### 2011-05-03 thesis.tex
Should have done this earlier...

\documentclass[10pt,a4paper]{report}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\author{James Pettit}
\begin{document}
\section{Introduction}
\section{The Game of Hex}
\section{State of the Art in Monte Carlo Tree Search}
UCT
AMAF
RAVE
Heuristics
Tree Bias
\section{Problem Statement}
\section{Solution}
\section{Conclusion}
\subsection{Conclusions}
Conclusions are not a rambling summary of the thesis: they are short, concise statements of the inferences that you have made because of your work. It helps to organize these as short numbered paragraphs, ordered from most to least important.
\subsection{Summary of Contributions}
The Summary of Contributions will be much sought and carefully read by the examiners. Here you list the contributions of new knowledge that your thesis makes. Of course, the thesis itself must substantiate any claims made here. There is often some overlap with the Conclusions, but that's okay. Concise numbered paragraphs are again best. Organize from most to least important.
\subsection{Future Research}
The Future Research subsection is included so that researchers picking up this work in future have the benefit of the ideas that you generated while you were working on the project. Again, concise numbered paragraphs are usually best.
\end{document}

### 2011-05-03 gui/main.py

@@ -1,0 +1,172 @@
+#!/usr/bin/env python
+import sys
+import pygtk  
+pygtk.require("2.0")  
+import gtk  
+import gtk.glade  
+import gobject
+import pygame
+import pygame.gfxdraw
+import os
+import sgflib
+import math
+
+class GUI(object):
+	def __init__(self):
+		gtk.settings_get_default().props.gtk_button_images = True
+		
+		self.builder = gtk.Builder()
+		self.builder.add_from_file(os.path.join(os.path.dirname(os.path.realpath(__file__)), "gui.glade"))
+		self.builder.connect_signals(self)
+		
+		self.builder.get_object("sgf_file_filter").add_pattern("*.sgf")
+		
+		self.builder.get_object("forward").set_sensitive(False)
+		self.builder.get_object("back").set_sensitive(False)
+		
+		canvas = self.builder.get_object("canvas")
+		os.putenv('SDL_WINDOWID', str(canvas.window.xid))
+		gtk.gdk.flush()
+		pygame.init()
+		pygame.display.set_mode((400, 400), 0, 0)
+		self.screen = pygame.display.get_surface()
+		self.size = None
+		self.last = None
+		gobject.idle_add(self.draw)
+		
+	def set_size(self, size):
+		self.size = size
+		self.board = [['empty' for i in range(self.size)] for j in range(self.size)]
+
+	def on_back_clicked(self, btn):
+		try:
+			self.cur.previous()
+			self.builder.get_object("forward").set_sensitive(True)
+			self.cur.next()
+			if 'B' in self.cur.node.data:
+				i = self.cur.node['B'][0][0]
+				j = self.cur.node['B'][0][1]
+			elif 'W' in self.cur.node.data:
+				i = self.cur.node['W'][0][0]
+				j = self.cur.node['W'][0][1]
+			else:
+				print 'error'
+			i, j = ord(i)-97, ord(j)-97
+			if i < self.size and j < self.size:
+				self.board[i][j] = 'empty'
+			self.cur.previous()
+			if 'B' in self.cur.node.data:
+				i = self.cur.node['B'][0][0]
+				j = self.cur.node['B'][0][1]
+			elif 'W' in self.cur.node.data:
+				i = self.cur.node['W'][0][0]
+				j = self.cur.node['W'][0][1]
[…]

### 2011-05-03 src/learning.go

@@ -10,6 +10,7 @@ import (
 	"fmt"
 	"strings"
 	"strconv"
+	"container/vector"
 )
 
 type Swarm struct {
@@ -161,6 +162,7 @@ func (s *Swarm) Step() {
 		children[i] = s.recombine(p)
 		s.mutate(children[i])
 	}
+	// propagate the 2 last best particles without change
 	children[0] = s.Particles[0].Copy()
 	children[1] = s.Particles[1].Copy()
 
@@ -174,11 +176,23 @@ func (s *Swarm) Step() {
 
 	// select mu parents from either children (,) or children + parents (+)
 	sort.Sort(children)
+	
+	f, err := os.Open(fmt.Sprintf("children-%d.gob", s.Generation), os.O_RDWR|os.O_TRUNC|os.O_CREAT, 0666)
+	if err != nil {
+		log.Println("failed to save swarm")
+		return
+	}
+	defer func() { f.Close() }()
+	e := gob.NewEncoder(f)
+	e.Encode(children)
+	
 	for i := uint(0); i < s.Mu; i++ {
 		s.Particles[i] = children[i]
 	}
 
 	s.Generation++
+	
+	s.SaveSwarm(fmt.Sprintf("swarm-%d.gob", s.Generation))
 }
 
 func randParticle(p Particles, e Particles) (r *Particle) {
@@ -302,7 +316,6 @@ func Train() {
 	for s.Generation < s.Generations {
 		start := time.Nanoseconds()
 		s.Step()
-		s.SaveSwarm(fmt.Sprintf("swarm-%d.gob", s.Generation))
 		log.Printf("generation %d/%d, best %.0f wins, took %d seconds",
 			s.Generation, s.Generations, s.Best().Fitness,
 			(time.Nanoseconds()-start)/1e9)
@@ -311,6 +324,49 @@ func Train() {
 
 func ShowSwarm(filename string) {
 	s := new(Swarm)
-	s.LoadSwarm(filename)
-	log.Println(s.Generation)
+	f, _ := os.Open("swarm-1.gob", os.O_RDONLY, 0)
+	defer func() { f.Close() }()
+	d := gob.NewDecoder(f)
+	d.Decode(s)
+	children := make(Particles, s.Lambda)
+	f, _ = os.Open("children-0.gob", os.O_RDONLY, 0)
[…]

### 2011-05-03 src/patterns.go

@@ -430,7 +430,7 @@ func drawHex(xoff, yoff, width float64, pos int, style1, style2 string, s *svg.S
 	}
 }
 
-func drawPat(bits string, xoff, yoff, width float64, pat []uint8, weights []float64, s *svg.SVG) {
+func drawPat(bits string, xoff, yoff, width float64, pat []uint8, weights []float64, index int, s *svg.SVG) {
 	sum := 0.0
 	for i := range pat {
 		sum += weights[i]
@@ -470,6 +470,7 @@ func drawPat(bits string, xoff, yoff, width float64, pat []uint8, weights []floa
 		}
 	}
 	s.Text(int(xoff+5*width+10), int(yoff+20), ws)
+	s.Text(int(xoff+5*width+10), int(yoff+40), fmt.Sprintf("%d", index))
 }
 
 func ShowPatterns() {
@@ -479,13 +480,15 @@ func ShowPatterns() {
 		defer func() { f.Close() }()
 		s := svg.New(f)
 		
-		max_pats := 1000
-		s.Start(1000, 40*6*max_pats+40)
+		pats := []int {
+16033,
+		}
+		s.Start(1000, 40*6*len(pats)+40)
 		p := matcher.(*Particle)
 		y := 0
-		for i := 0; i < max_pats; i++ {
+		for i := range pats {
 			pat := make([]uint8, 7)
-			bits := fmt.Sprintf("%.14b", i)
+			bits := fmt.Sprintf("%.14b", pats[i])
 			legal := true
 			off := 0
 			for j := 0; j < 7; j++ {
@@ -499,7 +502,7 @@ func ShowPatterns() {
 			}
 			if off % 2 != 0 { legal = false }
 			if legal {
-				drawPat(bits, 10, 40*6*float64(y)+40, 40, pat, p.Position[i*7:i*7+7], s)
+				drawPat(bits, 10, 40*6*float64(y)+40, 40, pat, p.Position[pats[i]*7:pats[i]*7+7], i, s)
 				y++
 			}
 		}

### 2011-05-03 gui/main.py
Working on background section.

@@ -16,7 +16,7 @@ class GUI(object):
 		gtk.settings_get_default().props.gtk_button_images = True
 		
 		self.builder = gtk.Builder()
-		self.builder.add_from_file(os.path.join(os.path.dirname(os.path.realpath(__file__)), "gui.glade"))
+		self.builder.add_from_file(get_resource_path("gui.glade"))
 		self.builder.connect_signals(self)
 		
 		self.builder.get_object("sgf_file_filter").add_pattern("*.sgf")
@@ -32,7 +32,10 @@ class GUI(object):
 		self.screen = pygame.display.get_surface()
 		self.size = None
 		self.last = None
+		self.font = pygame.font.Font(get_resource_path("DroidSansMono.ttf"), 16)
 		gobject.idle_add(self.draw)
+		if len(sys.argv) == 2:
+			self.set_file(sys.argv[1])
 		
 	def set_size(self, size):
 		self.size = size
@@ -90,13 +93,16 @@ class GUI(object):
 			self.builder.get_object("forward").set_sensitive(False)
 		
 	def on_file_set(self, chooser):
-		self.tree = sgflib.SGFParser(open(chooser.get_filename()).read()).parse()[0]
-		self.set_size(int(self.tree[0]['SZ'][0]))
-		self.cur = self.tree.cursor()
-		self.builder.get_object("forward").set_sensitive(True)
+		self.set_file(chooser.get_filename())
 		
 	def on_window_destroy(self, widget):
 		gtk.main_quit()
+		
+	def set_file(self, filename):
+		self.tree = sgflib.SGFParser(open(filename).read()).parse()[0]
+		self.set_size(int(self.tree[0]['SZ'][0]))
+		self.cur = self.tree.cursor()
+		self.builder.get_object("forward").set_sensitive(True)
 
 	def draw(self):
 		gobject.idle_add(self.draw)
@@ -109,6 +115,8 @@ class GUI(object):
 		black = (0, 0, 0)
 		white = (255, 255, 255)
 		gray = (200, 200, 200)
+		red = (255, 0, 0)
+		blue = (67, 83, 255)
 
 		C = 12
 		A = 0.5*C
@@ -139,18 +147,27 @@ class GUI(object):
 		ly.pop(len(ly)-1)
 		rx.pop(0)
 		ry.pop(0)
[…]

### 2011-05-06 training_progress.sh
Adding stuff to git.

@@ -1,0 +1,8 @@
+#!/bin/bash
+(
+while ((1)); do
+	percent=`grep "/" training.log | tail -n 1 | awk 'BEGIN { FS="/" } {print ($1/$2)*100}'`
+	echo $percent
+	sleep 1
+done
+) | dialog --shadow --guage "Training Progress" 7 100

### 2011-05-06 src/learning.go
Adding hooks to test swarm versus random, one function added to learning, one option added to main.

@@ -91,20 +91,13 @@ func (s Particles) Swap(i, j int) {
 	s[i], s[j] = s[j], s[i]
 }
 
-func (s *Swarm) evaluate(p *Particle) {
-	defer func() { recover() }()
-	var m PatternMatcher
-	if *tablepat {
-		m = p
-	} else {
-		m = &NeuralNet{s.Arch, p.Position}
-	}
+func playOneGame(black PatternMatcher, white PatternMatcher) Tracker {
 	t := NewTracker(*size)
 	passes := 0
 	var vertex int
 	for {
 		br := NewRoot(BLACK, t)
-		genmove(br, t, m)
+		genmove(br, t, black)
 		if br == nil || br.Best() == nil {
 			vertex = -1
 			passes++
@@ -117,7 +110,7 @@ func (s *Swarm) evaluate(p *Particle) {
 			break
 		}
 		wr := NewRoot(WHITE, t)
-		genmove(wr, t, m)
+		genmove(wr, t, white)
 		if wr == nil || wr.Best() == nil {
 			vertex = -1
 			passes++
@@ -130,6 +123,18 @@ func (s *Swarm) evaluate(p *Particle) {
 			break
 		}
 	}
+	return t
+}
+
+func (s *Swarm) evaluate(p *Particle) {
+	defer func() { recover() }()
+	var m PatternMatcher
+	if *tablepat {
+		m = p
+	} else {
+		m = &NeuralNet{s.Arch, p.Position}
+	}
+	t := playOneGame(m, m)
 	if t.Winner() == BLACK {
 		p.Fitness += 1
 	} else if t.Winner() == WHITE {
@@ -322,6 +327,7 @@ func Train() {
 	}
 }
 
+// converts gob file to arff for machine learning
 func ShowSwarm(filename string) {
 	s := new(Swarm)
 	f, _ := os.Open("swarm-1.gob", os.O_RDONLY, 0)
@@ -369,4 +375,31 @@ func ShowSwarm(filename string) {
 		fmt.Fprintln(f)
 		f.Sync()
 	}
+}
+
+// runs input swarm against random (generation 0)
+func TestSwarm(filename string) {
+	swarm := LoadTablePatternMatcher(filename)
+	rand := &RandomMatcher{}
+	swarm_black_wins, swarm_white_wins, rand_black_wins, rand_white_wins := 0.0, 0.0, 0.0, 0.0
+	rounds := 10
+	for i := 0; i < rounds; i++ {
[…]

### 2011-05-06 src/main.go

@@ -14,6 +14,7 @@ var server *string = flag.String("server", "", "Connect to redis server")
 var maxPlayouts *uint = flag.Uint("p", 10000, "Max number of playouts")
 var file *string = flag.String("file", "", "Load data from file")
 var showSwarm *bool = flag.Bool("showswarm", false, "show info on swarm")
+var testSwarm *bool = flag.Bool("testswarm", false, "run swarm tests")
 var showPatterns *bool = flag.Bool("showpat", false, "show patterns")
 var sgf *string = flag.String("sgf", "", "Load sgf file and generate move")
 var stats *bool = flag.Bool("stats", false, "Print out tree search statistics")
@@ -102,6 +103,8 @@ func main() {
 		TestPPS()
 	} else if *file != "" && *showSwarm {
 		ShowSwarm(*file)
+	}else if *file != "" && *testSwarm {
+		TestSwarm(*file)
 	} else if matcher != nil && *showPatterns {
 		ShowPatterns()
 	}

### 2011-08-17 gui/main.py
Mostly final version that is being used to generate results for thesis.

@@ -37,6 +37,10 @@ class GUI(object):
 		gobject.idle_add(self.draw)
 		if len(sys.argv) == 2:
 			self.set_file(sys.argv[1])
+		else:
+			self.set_size(9)
+			self.pb = "None"
+			self.pw = "None"
 		
 	def set_size(self, size):
 		self.size = size
@@ -49,11 +53,21 @@ class GUI(object):
 			if ord(key[0]) > ord('I'):
 				i -= 1
 			j = int(key[1]) - 1
+			black = float(value['black'])
+			white = float(value['white'])
 			self.board[i][j] = {
 				'occ': value['occ'],
-				'black': value['black'],
-				'white': value['white'],
+				'black': black,
+				'white': white,
 			}
+			if black > bmax:
+				bmax = black
+			if white > wmax:
+				wmax = white
+		for i in range(self.size):
+			for j in range(self.size):
+				self.board[i][j]['black'] /= bmax
+				self.board[i][j]['white'] /= wmax
 
 	def on_back_clicked(self, btn):
 		if type(self.cur) == int:
@@ -96,6 +110,7 @@ class GUI(object):
 				self.last = (i, j)
 			except sgflib.GameTreeEndError:
 				self.builder.get_object("back").set_sensitive(False)
+		gobject.idle_add(self.draw)
 
 	def on_forward_clicked(self, btn):
 		if type(self.cur) == int:
@@ -126,9 +141,11 @@ class GUI(object):
 						self.board[i][j] = color
 			except sgflib.GameTreeEndError:
 				self.builder.get_object("forward").set_sensitive(False)
+		gobject.idle_add(self.draw)
 		
 	def on_file_set(self, chooser):
 		self.set_file(chooser.get_filename())
+		gobject.idle_add(self.draw)
 		
 	def on_window_destroy(self, widget):
 		gtk.main_quit()
@@ -145,20 +162,28 @@ class GUI(object):
 			self.pb = "None"
 			self.pw = "None"
 			f = open(filename)
-			self.weights = map(json.loads, filter(lambda line: line.startswith("{"), f.readlines()))
+			lines = filter(lambda line: line.startswith("{"), f.readlines())
+			self.weights = []
+			for line in lines:
+				line = line.replace('+Inf', '"+Inf"')
+				try:
+					self.weights.append(json.loads(line))
[…]

### 2011-08-17 src/config.go

@@ -30,6 +30,7 @@ type Config struct {
 	// Different games
 	Go       bool
 	Hex      bool
+	HexFast  bool
 
 	// Game-specific variables
 	Size int
@@ -37,13 +38,13 @@ type Config struct {
 	Swapsafe bool
 
 	// Learning
-	Train       bool
-	Generations uint
-	Mu          uint
-	Parents     uint
-	Lambda      uint
-	Samples     uint
-	EvalErr     bool
+	Train        bool
+	Generations  uint
+	Mu           uint
+	Parents      uint
+	Lambda       uint
+	Samples      uint
+	Propagate    uint
 	
 	// Load/save different modules
 	Prefix string
@@ -53,6 +54,7 @@ type Config struct {
 	Sfile  string
 
 	// Tree exploration/expansion
+	TreeSearch  bool
 	Explore     float64
 	RAVE        float64
 	ExpandAfter float64
@@ -61,6 +63,10 @@ type Config struct {
 	Neighbors   bool
 	Ancestor    bool
 	Seed        bool
+	PlayoutProbs bool
+	PlayoutSuggest bool
+	PlayoutSuggestUniform bool
+	PlayoutSuggestUniformTenuki bool
 
 	// Logging
 	Verbose bool
@@ -87,7 +93,6 @@ type Config struct {
 	policy_weights   *Particle
 	
 	// log files
-	evalLog *os.File
 	probLog *os.File
 }
 
@@ -111,18 +116,19 @@ func NewConfig() *Config {
 
 	flag.BoolVar(&config.Go, "go", false, "Play Go")
 	flag.BoolVar(&config.Hex, "hex", false, "Play Hex")
+	flag.BoolVar(&config.HexFast, "hexfast", false, "Play Hex using fast tracker")
 
 	flag.IntVar(&config.Size, "size", 9, "Boardsize")
 	flag.Float64Var(&config.Komi, "komi", 6.5, "Komi")
 	flag.BoolVar(&config.Swapsafe, "swapsafe", false, "When playing hex, black will choose a swap-safe opening move")
 
-	flag.BoolVar(&config.Train, "train", false, "Do crazy unsupervised training stuff")
-	flag.UintVar(&config.Generations, "gens", 100, "Generations to train for")
+	flag.BoolVar(&config.Train, "train", false, "(Training) Do crazy unsupervised training stuff")
+	flag.UintVar(&config.Generations, "gens", 100, "(Training) Generations to train for")
[…]

### 2011-08-17 src/gotracker.go

@@ -419,8 +419,8 @@ func (t *GoTracker) updateWeights(vertex int) {
 			weight := t.get_weight(BLACK, vertex)
 			if math.IsNaN(weight) {
 				black_weight = 0
-			} else if black_weight + int(weight) > 0 {
-				black_weight += int(weight)
+			} else if black_weight + weight > 0 {
+				black_weight += weight
 			}
 		}
 		t.weights.Set(BLACK, vertex, black_weight)
@@ -429,8 +429,8 @@ func (t *GoTracker) updateWeights(vertex int) {
 			weight := t.get_weight(WHITE, vertex)
 			if math.IsNaN(weight) {
 				white_weight = 0
-			} else if white_weight + int(weight) > 0 {
-				white_weight += int(weight)
+			} else if white_weight + weight > 0 {
+				white_weight += weight
 			}
 		}
 		t.weights.Set(WHITE, vertex, white_weight)
@@ -692,10 +692,6 @@ func (t *GoTracker) String() (s string) {
 		}
 	}
 	return
-}
-
-func (t *GoTracker) WeightString() string {
-	return ""
 }
 
 func (t *GoTracker) dead() []int {

### 2011-08-17 src/hextracker.go

@@ -7,6 +7,7 @@ import (
 	"log"
 	"json"
 	"container/vector"
+	"rand"
 )
 
 type HexTracker struct {
@@ -98,6 +99,11 @@ func (t *HexTracker) Copy() Tracker {
 
 func (t *HexTracker) Play(color byte, vertex int) {
 	if vertex != -1 {
+		if t.board[vertex] != EMPTY {
+			log.Println(t.String())
+			log.Println(Ctoa(color), t.Vtoa(vertex))
+			panic("play on non-empty vertex")
+		}
 		root := find(vertex, t.parent)
 		for i := 0; i < 6; i++ {
 			adj := find(t.adj[vertex*6+i], t.parent)
@@ -123,7 +129,7 @@ func (t *HexTracker) Play(color byte, vertex int) {
 		// cannot play on occupied vertex
 		t.weights.Set(BLACK, vertex, 0)
 		t.weights.Set(WHITE, vertex, 0)
-		if t.config.policy_weights != nil {
+		if t.config.PlayoutProbs && t.config.policy_weights != nil {
 			t.updateNeighborWeights(vertex)
 		}
 		
@@ -135,36 +141,23 @@ func (t *HexTracker) Play(color byte, vertex int) {
 }
 
 func (t *HexTracker) updateNeighborWeights(vertex int) {
-	if t.board[vertex] == EMPTY {
-		return
-	}
 	for i := range t.neighbors[1][vertex] {
 		neighbor := t.neighbors[1][vertex][i]
 		if neighbor != -1 && t.board[neighbor] == EMPTY {
-			t.updateWeights(neighbor)
+			t.updateWeights(BLACK, neighbor, neighbor)
+			t.updateWeights(WHITE, neighbor, neighbor)
+			t.updateWeights(BLACK, vertex, neighbor)
+			t.updateWeights(WHITE, vertex, neighbor)
 		}
 	}
 }
 
-func (t *HexTracker) updateWeights(vertex int) {
-	black_weight := t.weights.Get(BLACK, vertex)
-	if black_weight != 0 {
-		weight := t.get_pattern_weight(BLACK, vertex)
-		black_weight = int(float64(black_weight) * weight)
-		if black_weight == 0 {
-			black_weight = 1
-		}
+func (t *HexTracker) updateWeights(color byte, v1, v2 int) {
+	weight := t.get_pattern_weight(color, v1) * t.weights.Get(color, v2)
[…]

### 2011-08-17 src/learning.go

@@ -11,9 +11,6 @@ import (
 	"log"
 	"container/vector"
 	"github.com/ajstarks/svgo"
-	"bufio"
-	"json"
-	"runtime"
 )
 
 type Swarm struct {
@@ -41,20 +38,7 @@ func NewSwarm(config *Config) *Swarm {
 	s.Generation = 0
 	s.Particles = make(Particles, s.Mu)
 	for i := uint(0); i < s.Mu; i++ {
-		s.Particles[i] = NewParticle(s, 0.01, 10.0)
-	}
-	if config.EvalErr {
-		s.evals = new(vector.Vector)
-		buf := bufio.NewReader(s.config.evalLog)
-		for line, _, err := buf.ReadLine(); err == nil; line, _, err = buf.ReadLine() {
-			eval := new(Eval)
-			err := json.Unmarshal(line, &eval)
-			if err != nil {
-				log.Println(err)
-			} else {
-				s.evals.Push(eval)
-			}
-		}
+		s.Particles[i] = NewParticle(s, 0, 100)
 	}
 	return s
 }
@@ -74,7 +58,7 @@ func NewParticle(swarm *Swarm, min, max float64) *Particle {
 	p.Position = make(map[uint32]float64)
 	p.Min = min
 	p.Max = max
-	p.Fitness = math.MaxFloat64
+	p.Fitness = 0
 	return p
 }
 
@@ -99,7 +83,7 @@ func (s Particles) Len() int {
 }
 
 func (s Particles) Less(i, j int) bool {
-	return s[i].Fitness < s[j].Fitness
+	return s[i].Fitness > s[j].Fitness
 }
 
 func (s Particles) Swap(i, j int) {
@@ -117,11 +101,11 @@ func (p *Particle) Init(i uint32) {
 	p.Position[i] = p.Min + (p.Max - p.Min) * rand.Float64()
 }
 
-func (s *Swarm) evalPlay(p *Particle) {
+func (s *Swarm) evalPlay(p1 *Particle, p2 *Particle) {
 	config := new(Config)
 	*config = *s.config
-	p.Fitness = 0
 	for sample := uint(0); sample < s.Samples; sample++ {
+		log.Printf("game %d / %d\n", sample, s.Samples)
 		t := NewTracker(config)
 		color := BLACK
 		target := BLACK
@@ -130,46 +114,31 @@ func (s *Swarm) evalPlay(p *Particle) {
 		}
 		for {
 			if color == target {
[…]

### 2011-08-17 src/main.go

@@ -50,25 +50,27 @@ func main() {
 		root := NewRoot(BLACK, t, config)
 		genmove(root, t)
 	} else if config.PlayGame {
-		t := NewTracker(config)
-		color := BLACK
-		move := 0
-		for {
-			var vertex int
-			if move == 0 && config.Swapsafe {
-				vertex = (3 * t.Boardsize()) + 2
-			} else {
-				root := NewRoot(color, t, config)
-				genmove(root, t)
-				vertex = root.Best().Vertex
+		for i := uint(0); i < config.Samples; i++ {
+			t := NewTracker(config)
+			color := BLACK
+			move := 0
+			for {
+				var vertex int
+				if move == 0 && config.Swapsafe {
+					vertex = (3 * t.Boardsize()) + 2
+				} else {
+					root := NewRoot(color, t, config)
+					genmove(root, t)
+					vertex = root.Best().Vertex
+				}
+				t.Play(color, vertex)
+				log.Println(t.String())
+				if t.Winner() != EMPTY {
+					break
+				}
+				move++
+				color = Reverse(color)
 			}
-			t.Play(color, vertex)
-			log.Println(t.String())
-			if t.Winner() != EMPTY {
-				break
-			}
-			move++
-			color = Reverse(color)
 		}
 	}
 	<-shutdown

### 2011-08-17 src/search.go

@@ -9,8 +9,6 @@ import (
 	"container/vector"
 	"os"
 	"gob"
-	"json"
-	"sort"
 )
 
 type Node struct {
@@ -67,7 +65,12 @@ func genmove(root *Node, t Tracker) {
 	root.copy_time = 0
 	root.next_count = 0
 	root.play_count = 0
-	playouts := float64(treeSearch(root, t))
+	var playouts float64
+	if root.config.TreeSearch {
+		playouts = float64(treeSearch(root, t))
+	} else {
+		playouts = float64(mcSearch(root, t))
+	}
 	elapsed := time.Nanoseconds() - start
 	elapsed_seconds := float64(elapsed) / 1e9
 	if root.config.Stats {
@@ -90,7 +93,6 @@ func genmove(root *Node, t Tracker) {
 			unaccounted := elapsed -
 				root.playout_time - root.update_time - root.win_calc_time - root.next_time - root.play_time - root.copy_time
 			log.Printf("%.2f seconds unaccounted\n", float64(unaccounted)/1e9)
-			root.LogEvals(t)
 		}
 		if root.config.Timelimit > 0 {
 			if elapsed_seconds > float64(root.config.Timelimit) {
@@ -124,24 +126,10 @@ func genmove(root *Node, t Tracker) {
 func treeSearch(root *Node, t Tracker) uint {
 	playouts := uint(0)
 	start := time.Nanoseconds()
-	s := time.Nanoseconds()
-	trackers := make([]Tracker, 10000)
-	for i := 0; i < len(trackers); i++ {
-		trackers[i] = t.Copy()
-	}
-	root.copy_time += (time.Nanoseconds() - s)
-	tracker := 0
 	for {
-		cp := trackers[tracker]
-		tracker++
-		if tracker >= len(trackers) {
-			s := time.Nanoseconds()
-			for i := 0; i < len(trackers); i++ {
-				trackers[i] = t.Copy()
-			}
-			root.copy_time += (time.Nanoseconds() - s)
-			tracker = 0
-		}
+		s := time.Nanoseconds()
+		cp := t.Copy()
+		root.copy_time += (time.Nanoseconds() - s)
 		root.step(cp)
 		playouts++
 		territory := cp.Territory(Reverse(root.Color))
@@ -173,6 +161,40 @@ func treeSearch(root *Node, t Tracker) uint {
 			}
 		} else if playouts >= root.config.MaxPlayouts {
 			break
+		}
+	}
+	return playouts
+}
+
[…]

### 2011-08-17 src/tracker.go

@@ -12,7 +12,7 @@ const (
 	UP_RIGHT   = 1
 	DOWN_LEFT  = 4
 	DOWN_RIGHT = 5
-	INIT_WEIGHT = 50
+	INIT_WEIGHT = 500
 )
 
 var SIDE_UP int
@@ -40,14 +40,17 @@ type Tracker interface {
 	String() string
 	Vtoa(vertex int) string
 	Atov(s string) int
-	WeightString() string
 }
 
 func NewTracker(config *Config) Tracker {
 	if config.Go {
 		return NewGoTracker(config)
 	} else if config.Hex {
-		return NewHexTracker(config)
+		if config.HexFast {
+			return NewFastHexTracker(config)
+		} else {
+			return NewHexTracker(config)
+		}
 	}
 	return nil
 }

### 2011-08-17 src/weight_tree.go

@@ -10,7 +10,8 @@ type WeightTree struct {
 	nodes int
 	interior int
 	leaves int
-	weights []int
+	black_weights []float64
+	white_weights []float64
 }
 
 func NewWeightTree(size int) *WeightTree {
@@ -23,45 +24,52 @@ func NewWeightTree(size int) *WeightTree {
 		t.nodes += 1 << i
 	}
 	t.interior = t.nodes - t.leaves
-	t.weights = make([]int, 2 * t.nodes)
+	t.black_weights = make([]float64, t.nodes)
+	t.white_weights = make([]float64, t.nodes)
 	return t
 }
 
-func (t *WeightTree) Set(color byte, vertex int, weight int) {
+func (t *WeightTree) Set(color byte, vertex int, weight float64) {
 	if color == BLACK {
-		t.update(t.interior + vertex, 0, weight, t.weights[t.interior + vertex])
+		t.black_weights[t.interior + vertex] = weight
+		for parent := t.parent(t.interior + vertex); parent >= 0; parent = t.parent(parent) {
+			t.black_weights[parent] = t.black_weights[2*parent+1] + t.black_weights[2*parent+2]
+		}
 	} else {
-		t.update(t.interior + vertex, t.nodes, weight, t.weights[t.nodes + t.interior + vertex])
+		t.white_weights[t.interior + vertex] = weight
+		for parent := t.parent(t.interior + vertex); parent >= 0; parent = t.parent(parent) {
+			t.white_weights[parent] = t.white_weights[2*parent+1] + t.white_weights[2*parent+2]
+		}
 	}
 }
 
-func (t *WeightTree) Get(color byte, vertex int) int {
+func (t *WeightTree) Get(color byte, vertex int) float64 {
 	if color == BLACK {
-		return t.weights[t.interior + vertex]
+		return t.black_weights[t.interior + vertex]
 	}
-	return t.weights[t.nodes + t.interior + vertex]
+	return t.white_weights[t.interior + vertex]
 }
 
 func (t *WeightTree) Prob(color byte, vertex int) float64 {
 	if color == BLACK {
-		return float64(t.Get(color, vertex)) / float64(t.weights[0])
+		return t.Get(color, vertex) / t.black_weights[0]
 	}
-	return float64(t.Get(color, vertex)) / float64(t.weights[t.nodes])
+	return t.Get(color, vertex) / t.white_weights[0]
 }
 
[…]

### 2011-08-17 thesis/thesis.tex
Adding thesis latex source.

\documentclass[10pt,a4paper]{report}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\newcommand{\black}{\texttt{Black}}
\newcommand{\white}{\texttt{White}}
\newcommand{\loc}[1]{\texttt{#1}}
\author{James Pettit}
\title{Simulation Learning in Computer Hex}
\begin{document}
\maketitle
\tableofcontents

\chapter{Introduction}
Abstract strategy games provide an easy testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally chosen for a variety of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundereds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. For at least Checkers, Chess, and similar games (Othello, Connect Four), brute force tree search has provided a silver bullet.

\section{Background: The Game of Hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by colors here \black{} and \white. The object of the game is to form an unbroken chain from one side of the board to the other. \black{} attempts to connect top to bottom, \white{} left to right. Figure \ref{fig:13x13inprogress} shows an example game in progress on a 9x9 board. Figure \ref{fig:13x13finished} shows a different game, where \black{} has just played the winning move, with an unbroken chain connecting top to bottom. Hex has properties that make it attractive for AI research. Once placed, a piece never leaves the board. The only criterion for a legal move is that the point be unoccupied (empty). Gale '79 proved the game cannot end in a draw, simplifying minimax methods. Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the strategy itself. Solutions for all opening positions for 7x7 and 8x8 boards have been found\cite{henderson2009solving}. A good portion, but not all, 9x9 openings have also been solved. Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.

Three Hex programs are currently in the running as viable world champions. Gabor Melis' \emph{Six} is the previous world champion (up to 2007). Six uses traditional alpha-beta search with and evaluation function based on an electical circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}. \emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves. \emph{MoHex}, also from Alberta, is based on the newer Monte-Carlo Tree Search technique, leveraging the open source \emph{Fuego} library. MoHex is the current world-champion, having finished ahead of Wolve in 2009 and 2010.

\begin{figure}
	\begin{center}
	\includegraphics[width=200px]{graphics/13x13-in-progress.png}
	\caption{13x13 Hex Game, \white{} to play.}
	\end{center}
	\label{fig:13x13inprogress}
\end{figure}
\begin{figure}
	\begin{center}
	\includegraphics[width=200px]{graphics/13x13-finished.png}
	\caption{Finished 13x13 Hex Game, \black{}'s play at \loc{b6} gives him an unbroken path from top to bottom.}
	\end{center}
	\label{fig:13x13finished}
\end{figure}

\section{Overview of the Thesis}\label{overview}
In Chapter \ref{mcts} we will start with the current state of the art in Monte Carlo Tree Search (MCTS). MCTS is a technique to stochastically sample paths through a game-tree. It has been found to work well in games where: the tree is too large to consider brute-force search, no good solution exists to prune ``bad'' nodes, and no good evaluation function exists. Section \ref{bias} covers the current methods of biasing the tree search, including some methods similar to the technique presented in this paper. Chapter \ref{problem} covers why finding a generic way to bias the tree search would be a step forward in MCTS, with a formal definition of biasing the search, allowing a comparison of different but similar techniques. Chapter \ref{solution} presents a way of biasing the tree that can be automatically evolved through self-play, and does not require outside ``expert'' knowledge. Finally Chapter \ref{results} summarizes the technique and its usefulness, and discusses what can be done in the future to improve and extend it.

\chapter{State of the Art in Monte Carlo Tree Search}\label{mcts}
As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS). The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a minimax search tree. The UCT algorithm was developed alongside MCTS to direct the growth of the search tree\cite{gelly2006exploration}. At any point, the search can be stopped and the current state of the tree used to select a move to play. Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.

All nodes in the tree corrospond to a fixed position, i.e. the occupancy state of the board and which player is next to move. Nodes also have statistics associated with each position - the number of winning simulations performed that used the node, and the total number of simulations that included the node. In general, at each iteration the algorithm will, starting at the root node, recursively traverse the tree until it reaches a leaf node whose estimated value is unknown. A simulation is run from the position of this leaf node and the result used to update all nodes along the path. The algorithm continues for some set number of iterations, each iteration corrosponding to one simulation.

\begin{verbatim}
function search(root):
  for i in range(simulations):
    step(root)

function step(node) returns outcome:
  if node.visits < threshold:
    outcome = simulate(node)
  else:
    next = select(node)
    outcome = step(next)
  if outcome is win for node's player:
    node.wins++
  node.visits++
  return outcome
\end{verbatim}

\subsection{Selection Policy}
The \texttt{select} function defines the selection policy. It must return the child of the given node the step into next. In general, selection must trade off between \emph{exploitation} and \emph{exploration}. It is advantageous to exploit child nodes with high win-rates, to discover possible weaknesses that would result in the opponent winning. Simultaneously, a node with a low win-rate might only have had a few unlucky simulations run through it, resulting in an incorrect value estimate. Nodes with a low number of visits must be explored to see if they result in a high-value position.

One function that balances this trade-off is Upper-Confidence for Trees (UCT). UCT incorporates the win-rate of a child with a term that increases in value as the ratio of the child's visits to its parents decreases. Eventually, no matter how bad the estimated value of a position is, it will be re-visited and the estimate updated. UCT has several variations, but the ``pure'' version has been proven to eventually yield the correct minimax value of a position, given an infinite number of simulations\cite{gelly2006exploration}.

\[
	\texttt{winrate} + C*\sqrt{\frac{\log{(\texttt{parent.visits})}}{\texttt{child.visits}}}
\]

$C$, the \emph{exploration coefficient}, can be varied to increase or decrease the level of exploration versus exploitation. Note that a very low value will yield a function that relies almost exclusively on the estimated win-rate of a position, while a high value will visit nodes more equally, placing less emphasis on the estimated value.

\subsection{Simulation Policy}
The \texttt{simulate} function plays out the rest of the game from the current position. The simplest policy plays a legal uniform random move until the game ends (in Hex, when one player has a winning path). The simulation policy has the largest effect on the strength of the entire search. Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move. The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stocastic choice between several candidates. Only when none of the rules match do they fall back to a uniform random move. It was quickly found that building a stronger simulation policy does not guarantee better performance when used in the full tree search. A test and check methodology is required, with changes to the simulation policy having unpredictable effects on overall performance.

Recent research has focused on learning a simulation policy...

\subsection{Update Policy}
AMAF/RAVE

\section{Biasing the Tree}\label{bias}
In general, there are three ways to bias the tree search. First, upon expansion of a node, an initial estimate of the node's value is used. Over time, the initial estimate is gradually replaced RAVE and progressive bias are examples

Mogo and others showed that using heuristics in both the expansion and playout policy can prove extremely beneficial. However, this requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games). It was also found that just improving the playout policy to play better was not enough. The playout policy can implicitly prune good moves, never allowing them to be played. If there is some small probability of playing a purely random move, the convergence properties of the MCT search will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence. Biasing tree expansion has similar problems. If the initial value estimate is incorrect, the selection policy will eventually expand the node enough times to find a winning sequence, but in practice the heuristics can hurt performance.

\chapter{Problem Statement}\label{problem}
Want to find a way to learn how to bias the search. Hand-crafted heuristics are tedious and don't extend to new games. Learning heuristics from expert games can be automatic, but requires a corpus, again, can't be used on new games.

\chapter{Solution}\label{solution}

\section{Evolution Strategies}
Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors. A general form of ES is given here, with some small modifications required for it to represent a biased simulation policy discussed in Section \ref{esmods}. More information on evolution strategies and evolutionary learning in general can be found in Eberhard and Shi's excellent textbook \emph{Computational Intelligence: Concepts to Implementations}\cite{eberhart2007computational}. The Evolution Strategies article on Scholarpedia.org also provides a good introduction to the topic\cite{Beyer:2007}.

\begin{enumerate}
\item A population (usually 20-200 individuals) is generated. In ES, each individual is represented by its vector, often called its ``position'' ($\bf{p}$), a single real-valued ``strategy'' ($\sigma$), and, once evaluation is complete, a single real-valued ``fitness'' attribute ($f$).
\item The algorithm loops, each loop forming one ``generation'':
	\begin{enumerate}
	\item From the $\mu$ parents, generate $\lambda$ children.
		\begin{enumerate}
		\item Each child is created by selecting $P$ parents.
		\item These parents (usually 2 or 3) are \emph{recombined} by averaging their positions component-wise to form the new child position. Similarly, the child's strategy is the average of its $P$ parents.
		\item	The child's strategy ($\sigma$) is \emph{mutated} by a Gaussian distributed random variable.
		\item The child's position is mutated component-wise by a vector of iid Gaussians, one per component of the dimension. These Gaussians are scaled by the strategy. Thus, one individual may mutate more (faster) than another, by using a larger strategy. These strategies mutate along with the positions, allowing the system to evolve the correct mutation weight along with the position.
		\end{enumerate}
	\item The best $b$ performing parents are propagated through without mutation, replacing $b$ children. This is done to preserve well-performing solutions.
	\item Evaluate the fitness of the $\lambda$ children (see \ref{fitness}).
	\item Sort the children by fitness.
	\item Cull the children by selecting only the top $\mu$. These now form the next generation of parents.
	\end{enumerate}
\end{enumerate}

Figure \ref{mutate} shows the differential equations for this evolutionary update. Note that $\sigma$ is multiplied at each time step, and can never reach zero. $\tau$, often called the ``learning rate'', can be scaled up or down to increase or decrease the variance of $\sigma$, a high values will give more mutation, low values less. The initial value of $\tau$ is $\frac{1}{\sqrt{2D}}$, $D$ the dimensionality of the search space. The rate decreases over time, as the number of generations approaches a set maximum, this forces the algorithm to converge, though not necessarily on a suitable solution. This choice of $\tau$ is driven by experimental literature suggesting it works in many cases\cite{Beyer:2007}.

\begin{figure}
\begin{flalign*}
	\tau_{t} = & \frac{1}{\sqrt{2D}} \left(1 - \frac{\textrm{generation}}{\textrm{max generations}}\right) \\
	\sigma_{t} = & \sigma_{t-1} e^{N(0, \tau^2)} \\
	p_{i_{t}} = & p_{i_{t-1}} N(0, \sigma^2)
\end{flalign*}
\caption{Differential equations for $\tau$, $\sigma$, and $\bf{p}$, indexed by time $t$. $D$ is the number of dimensions, i.e. $|\bf{p}|$}
\end{figure}\label{mutate}

\section{Using ES to Bias a Simulation}\label{esmods}

\section{Fitness Evaluation}\label{fitness}

\chapter{Results and Conclusions}\label{results}
\begin{itemize}
\item uniform local vs. control: 69.33\%
\item uniform local with tenuki vs. control: 61.33\%
\item learned local vs. control: 77.63\%
\item uniform local with tenuki vs. uniform local: 45.33\%
\item learned local vs. uniform local: 73.68\%
\item learned local vs. uniform local with tenuki: 75.00\%
\end{itemize}

Learned local play is the clear winner. Interestingly, the addition of a tenuki probability lowers performance in all cases. As expected, uniform local play significantly improves upon pure random playouts. The fact that evolutionary learning can find weights that boost performance versus uniform local in the same manner that uniform local improves upon pure random shows how much further there is to go, and shows the weaknesses of purely random simulations.

\bibliography{thesis}
\bibliographystyle{plain}
\end{document}

### 2011-08-19 thesis/thesis.tex
Adding to results, finalized title page conformant to grad division layout.

@@
 \newcommand{\loc}[1]{\texttt{#1}}
-\author{James Pettit}
-\title{Simulation Learning in Computer Hex}
 \begin{document}
-\maketitle
+
+\begin{titlepage}
+\begin{center}
+UNIVERSITY OF CALIFORNIA\\[0.5cm]
+SANTA CRUZ\\[0.75cm]
+\textbf{Simulation Learning in Computer Hex*}\\[0.5cm]
+\small{A report submitted in partial satisfaction}\\
+\small{of the requirements for the degree of}\\[0.5cm]
+MASTER OF SCIENCE\\[0.5cm]
+in\\[0.5cm]
+COMPUTER SCIENCE\\[0.5cm]
+by\\[0.5cm]
+\textbf{James Pettit}\\[0.5cm]
+\today
+\end{center}
+\vfill
+\begin{flushright}
+The Report of Simulation Learning in Computer Hex\\
+is approved:\\[1cm]
+\line(1,0){120}\\
+Professor David Helmbold\\[1cm]
+\line(1,0){120}\\
+Professor Robert Levinson\\[3cm]
+\end{flushright}
+\small{*Many thanks to David Helmbold, David Doshay, and Charlie McDowell for their advice, patience, and the computational resources required to complete this research.}
+\end{titlepage}
+
 \tableofcontents
@@
 
-Recent research has focused on learning a simulation policy...
+\subsection{Update Policy}
+Once a simulation is finished, the result is used to update the tree statistics.
+In the basic implementation, only the win/visit statistics along the traversed path are updated.
+It was quickly discovered that more sophisticated update strategies could be used to keep alternate statistics, and update them at nodes similar to those used in the playout that were not in the path.
+One simple example of this is the All-Moves-As-First (AMAF) heuristic.
+Briefly, it leverages the fact that moves played in the simulation could have been played in any order.
+Thus, a move deep in the simulation could be fictitiously referred to as the ``first'' move.
+In addition to updating the statistics of the nodes in the path, all siblings of the path nodes are also updated with a fictitious AMAF win/visit, if they participated in the simulation.
+Variations on AMAF have been tested, including Permutations AMAF, which updates node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
 
-\subsection{Update Policy}
-AMAF/RAVE
+Several methods exist to blend the AMAF results with the ``pure'' results.
+Formally, a given update policy must define the statistics kept for each node (e.g.
+\texttt{Pure\_Wins}, \texttt{Pure\_Visits}, \texttt{AMAF\_Wins}, \texttt{AMAF\_Visits}).
+These counts can be blended into the estimated win-rate used in the UCT selection policy.
+One simple policy is $\alpha$-AMAF, where a single value, $\alpha \in [0, 1]$, determines the relative weight given to each statistic, e.g.
 
-\section{Biasing the Tree}\label{bias}
-In general, there are three ways to bias the tree search.
-First, upon expansion of a node, an initial estimate of the node's value is used.
-Over time, the initial estimate is gradually replaced RAVE and progressive bias are examples
+\[
+\texttt{winrate} = \alpha\frac{\texttt{AMAF\_Wins}}{\texttt{AMAF\_Visits}} + (1 - \alpha)\frac{\texttt{Pure\_Wins}}{\texttt{Pure\_Visits}}
+\]
 
-Mogo and others showed that using heuristics in both the expansion and playout policy can prove extremely beneficial.
-However, this requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games).
-It was also found that just improving the playout policy to play better was not enough.
-The playout policy can implicitly prune good moves, never allowing them to be played.
-If there is some small probability of playing a purely random move, the convergence properties of the MCT search will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence.
-Biasing tree expansion has similar problems.
-If the initial value estimate is incorrect, the selection policy will eventually expand the node enough times to find a winning sequence, but in practice the heuristics can hurt performance.
+Other blending functions include Cutoff and Rapid Action Value Estimation (RAVE).
+Helmbold and Parker-Wood (2009) have a good summary of the various update policies and blending methods, along with experimental results showing marked improvement over the baseline of pure UCT\cite{helmbold2009all}.
+One noteworthy conjecture is that there is no ``silver bullet'' update policy.
+If the simulation policy is random and unbiased, the AMAF updates might be similarly unbiased.
+However, a simulation policy that preferentially plays certain moves might have an unpredictable effect on which nodes receive AMAF updates.
+This can then propagate to which nodes are selected during the UCT selection, and have chaotic effects on the search as a whole.
+
+\section{Biasing the Simulation}\label{bias}
+Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}.
+However, this frequently requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games).
+It was also found that improving the performance of playout policy did not guarantee an improvement in the tree search.
+For example, the playout policy can implicitly prune good moves, never allowing them to be played.
+In Go, the policy must correctly handle complicated situations like Nakade.
+If it fails to correctly play out a complex position, the tree might fail to see an opponent response.
+Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situation.
+If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence.
+In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in the selection policy\cite{gelly2007combining}.
+This puts most of the control of the search in the hands of the simulation policy.
 
 \chapter{Problem Statement}\label{problem}
-Want to find a way to learn how to bias the search.
-Hand-crafted heuristics are tedious and don't extend to new games.
-Learning heuristics from expert games can be automatic, but requires a corpus, again, can't be used on new games.
 
@@
 
-\section{Using ES to Bias a Simulation}\label{esmods}
+\section{Using ES to Bias the Simulation}\label{esmods}
 
@@
 \chapter{Results and Conclusions}\label{results}
+To test the effects of learning policy weights, the evolution process was ran for 41 generations.
+This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz.
+Although evolution strategies is easy to parallelize by running multiple evaluations simultaneously, the current implementation is single-threaded.
+After training was completed, the weights of top 4 individuals of the 41st generation were blended in proportion to their fitness to form a single individual.
+This final individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally.
+The control consisted of uniform random playouts.
+All players used a slightly modified UCT selection policy that included the variance term suggested by Gelly 2006\cite{gelly2006exploration}.
+A pure update policy was used, only the nodes directly in the path traversed by UCT were updated with the result of a simulation.
+
+200 games per pair were played, with time controls set at 5 seconds per move.
+All players have approximately the same performance with regard to simulations per second, so the 5 seconds per move resulted in around 10,000 simulated playouts.
+A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
+
 \begin{itemize}
-\item uniform local vs. control: 69.33\%
-\item uniform local with tenuki vs. control: 61.33\%
-\item learned local vs. control: 77.63\%
-\item uniform local with tenuki vs. uniform local: 45.33\%
-\item learned local vs. uniform local: 73.68\%
-\item learned local vs. uniform local with tenuki: 75.00\%
+\item uniform local vs. control: 65.00\%
+\item uniform local with tenuki vs. control: 58.50\%
+\item learned local vs. control: 81.50\%
+\item uniform local with tenuki vs. uniform local: 45.50\%
+\item learned local vs. uniform local: 76.50\%
+\item learned local vs. uniform local with tenuki: 79.00\%
 \end{itemize}
@@
 
+As an outside control, MoHex from the University of Alberta was compiled and ran with the default settings\cite{mohex}.
+Again, 200 games were played.
+As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move.
+Default time settings for MoHex are 10 seconds, but MoHex uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found.
+The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS.
+As a result, the time controls are not directly comparable.
+However, CPU time was tracked and can be used as another performance comparison.
+
+\begin{itemize}
+\item learned local vs. MoHex: 6.5\%, CPU=6.64x
+\item control vs. MoHex: 0\% CPU=5.55x
+\item learned local vs. control 92.5\% CPU=1.31x
+\end{itemize}
+
+Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played.
+Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength.
+
 \bibliography{thesis}

### 2011-08-19 thesis/thesis.tex
Starting on problem statement and solution.

@@
 \chapter{Solution}\label{solution}
+After each move, a random move is selected in the 6-connected local area.
+This move is used as the next move in the playout.
+All subsequent moves will also have their local area examined for a random response.
+If all 6 connected vertices are non-empty, a move is played using the uniform random policy (the \emph{default} policy).
+In the uniform local variation, a response is chosen uniformly from any empty moves in the local area.
+The uniform local variation with tenuki is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy.
+The learned local variation assigns a weight to each empty local vertex based on the local area.
+These weights are used to stochastically bias the local response.
+The weights are trained over time through self-play using Evolution Strategies.
 

### 2011-08-23 thesis/thesis.tex
A bit more in every section plus more references.

@@
 
-\chapter{Introduction}
+\renewcommand*\thesection{\arabic{section}}
+
+\section{Introduction}
 Abstract strategy games provide an easy testbed for artificial intelligence and machine learning applications.
@@
 
+Hex is very similar to the game of Havannah, another ``connection'' game.
+It also has similarities to the board game Go (also known as Igo, Weiqi, and Baduk).
+Go is a much older game than Hex or Havannah, and much research has been done on Go playing algorithms.
+Hex has simpler rules than Go, but has a similar branching factor and complexity.
+Much of this thesis will use Hex as a simpler analog for Go, with the techniques discussed being portable for use in any large branching game without easily crafted heuristics.
+
 \begin{figure}
@@
 
-\chapter{State of the Art in Monte Carlo Tree Search}\label{mcts}
+\section{State of the Art in Monte Carlo Tree Search}\label{mcts}
 As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS).
@@
 
-\section{Biasing the Simulation}\label{bias}
+\subsection{Biasing the Simulation}\label{bias}
 Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}.
@@
 
-\chapter{Problem Statement}\label{problem}
+\section{Problem Statement}\label{problem}
+MoGo, and Go programs in general, have the benefit of a rich history of play.
+Thousands of expert games are available for analysis.
+Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout\cite{chaslot2010adding}.
+More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move.
+Hex does have one simple pattern in the form of a bridge\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.
 
-\chapter{Solution}\label{solution}
+It would be ideal to automatically learn a better playout policy from self-play.
+The usual process of improvement requires hand-tuned weights that modify the playout policy.
+Any change must be tested to detect the change in strength it gives.
+Problematically, small changes in overall algorithm strength require thousands of games of testing to detect.
+Many times the improvement made by a small change in weight can be so small is will go unnoticed.
+Automatic weight tuning can improve this process, but still requires thousands of games to test.
+Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning\cite{silver2009monte}.
+This thesis presents a slightly different approach using evolutionary learning.
+It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
+
+\section{Solution}\label{solution}
 After each move, a random move is selected in the 6-connected local area.
@@
 
-\section{Evolution Strategies}
+\subsection{Evolution Strategies}
 Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors.
@@
 
-\section{Using ES to Bias the Simulation}\label{esmods}
+\subsection{Using ES to Bias the Simulation}\label{esmods}
 
-\section{Fitness Evaluation}\label{fitness}
+\subsection{Fitness Evaluation}\label{fitness}
+Every generation, each individual plays $n$ games against another random individual.
+Roughly half the games are played as black, half as white.
+The swap-safe rule was used to play the first move to further equalize the first-mover advantage.
+The winner of a game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.
 
-\chapter{Results and Conclusions}\label{results}
+\section{Results and Conclusions}\label{results}
 To test the effects of learning policy weights, the evolution process was ran for 41 generations.

### 2011-08-23 thesis/thesis.tex
Finished ES/bias section.

@@
 \newcommand{\white}{\texttt{White}}
+\newcommand{\blacks}{\texttt{Black's}}
+\newcommand{\whites}{\texttt{White's}}
 \newcommand{\loc}[1]{\texttt{#1}}
@@
 \subsection{Using ES to Bias the Simulation}\label{esmods}
+Evolution Strategies evolves a set of real numbered values that are used to bias the playouts.
+In most applications, the set of values is simply an array, and the array directly used in fitness evaluation.
+To apply ES to simulation biasing, an encoding is used that maps from a local pattern to a single 32-bit integer.
+The 32-bit integer is used to index a single real-valued weight for the pattern.
+To save space, a growable hashmap is used instead of an array with $2^{32}$ elements.
+In practice there are less than $2 * 4^6$ possible local patterns so a 16-bit integer could be used.
+However, this 32-bit hashmap implementation has small overhead and allows space to add properties into the encoding (for example, distance to one side) in the future.
+
+Figures \ref{fig:localpattern} and \ref{fig:encoding} show graphically how a local pattern is mapped to a 32-bit integer.
+The orange marker denotes the last move played.
+Around this move any empty vertices are examined for a local response.
+In the example shown in figure \ref{fig:localpattern}, the sum of the three weights is 11.
+Thus, vertex A will be selected with probability $P(A) = \frac{6}{11} \approx 0.54$.
+B will be selected with $P(B) \approx 0.27$ and C with $P(C) \approx 0.18$.
+Note that $P(A)+P(B)+P(C) = 1$ so one of A, B, or C will be selected.
+A pattern can have a weight of 0.
+If so, it will never be selected using the local response policy.
+If all patterns have weight 0, or if all adjacent vertices are non-empty, play falls back to the default policy (uniform random).
+
+Finally, note that in \ref{fig:encoding} the encoding includes a color bit in the 31st position.
+This denotes the color to play (either \black\ (1) or \white\ (0)).
+This bit is important because \black\ and \white\ have conflicting goals of moving, generally, on different diagonals.
+In fact, by using a reflection and flipping colors, one position can be translated into a similar position for the opposite color.
+A preliminary implementation was done that used this fact to compress the hashmap with the hope of accelerating learning.
+In practice this transform is tricky enough to perform that the simpler implementation using color bits was required to ensure correct results.
+
+Using a hashmap slightly complicates the recombination step of ES.
+The algorithm must now take the superset of patterns (i.e.
+hashmap keys) from all parents and use those to form the new individual.
+In practice the hashmap quickly grows to include all possible local patterns after just one generation.
+Mutation then operates on all values of the hashmap, instead of all values in an array.
+
+\begin{figure}
+	\begin{center}
+	\includegraphics[width=300px]{graphics/local-pattern.png}
+	\caption{A 5x5 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
+	\end{center}
+	\label{fig:localpattern}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+	\includegraphics[width=300px]{graphics/weight-pattern-map.png}
+	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
+	\end{center}
+	\label{fig:encoding}
+\end{figure}
 
 \subsection{Fitness Evaluation}\label{fitness}
+The ES population is initialized with $\lambda$ individuals, all with uniform random weights in the range $[0, 100]$.
+In practice, weights are not initialized until a pattern lookup is performed (a form of lazy evaluation).
+
 Every generation, each individual plays $n$ games against another random individual.

### 2011-08-24 thesis/thesis.tex
Needs editing and some comments on what to improve, I'm mostly out of ideas.

@@
 \subsection{Fitness Evaluation}\label{fitness}
-The ES population is initialized with $\lambda$ individuals, all with uniform random weights in the range $[0, 100]$.
-In practice, weights are not initialized until a pattern lookup is performed (a form of lazy evaluation).
-
-Every generation, each individual plays $n$ games against another random individual.
-Roughly half the games are played as black, half as white.
-The swap-safe rule was used to play the first move to further equalize the first-mover advantage.
+The ES population is initialized with $\lambda$ individuals, with uniform random weights in the range $[0, 100]$.
+In practice, weights are not initialized until a pattern lookup is performed (lazy lookup).
+Every generation, each individual plays $n$ complete games against another random individual.
+At the beginning of the game, one of the two individuals is randomly chosen to play \black, the other to play \white.
+The swap-safe rule was used to play the first move to further equalize first-mover advantage.
 The winner of a game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.
@@
 Although evolution strategies is easy to parallelize by running multiple evaluations simultaneously, the current implementation is single-threaded.
-After training was completed, the weights of top 4 individuals of the 41st generation were blended in proportion to their fitness to form a single individual.
+After training was completed, the 41st generation was used to form a single individual.
 This final individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally.
@@
 A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
+
+Instead of just using the best individual from the last generation, the weight vector used is a linear combination of the weights from all individuals in generation 41.
+The weights are blended based on relative fitness, so the final weight vector will most reflect the top performer of generation 41, but also includes the weights off all individuals.
+In some respect all individuals have ``survived'' 41 generations of recombination and mutation, and so were included in the final learned vector.
 

### 2011-08-24 thesis/thesis.tex
Spell checked, needs outside review.

@@
 This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov.
-These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundereds of moves.
+These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for 0 of moves.
 Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice.
@@
 Gabor Melis' \emph{Six} is the previous world champion (up to 2007).
-Six uses traditional alpha-beta search with and evaluation function based on an electical circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
+Six uses traditional alpha-beta search with and evaluation function based on an 0 circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
 \emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves.
@@
 
-All nodes in the tree corrospond to a fixed position, i.e.
+All nodes in the tree 0 to a fixed position, i.e.
 the occupancy state of the board and which player is next to move.
@@
 A simulation is run from the position of this leaf node and the result used to update all nodes along the path.
-The algorithm continues for some set number of iterations, each iteration corrosponding to one simulation.
+The algorithm continues for some set number of iterations, each iteration 0 to one simulation.
 
@@
 Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move.
-The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stocastic choice between several candidates.
+The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a 0 choice between several candidates.
 Only when none of the rules match do they fall back to a uniform random move.
@@
 For example, the playout policy can implicitly prune good moves, never allowing them to be played.
-In Go, the policy must correctly handle complicated situations like Nakade.
+In Go, the policy must correctly handle complicated situations like nakade.
 If it fails to correctly play out a complex position, the tree might fail to see an opponent response.
@@
 A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
+Experiments were ran using the Gomill software\cite{gomill}.
+Gomill required no modification to work with Hex playing programs.
 
@@
 
+\subsection{Conclusions}
+With infinite time, an unbiased playout policy can find the true minimax value of a position in the game of Hex.
+In practice, both resources and time are finite.
+Bias in the playout policy can be extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search.
+Evolutionary learning can provide a substantial performance boost over naive patterns.
+With self-play and evolution, weights can be learned automatically, without requiring expert knowledge or test-and-check tuning.
+The learning process can be done offline, and the final result can be used with little overhead.
+
+Source code for the Hivemind Computer Hex/Go player is available at \url{http://github.com/etherealmachine/hivemind}.
+The included README file has instructions on running the evolutionary learning process and using the results.
+
 \bibliography{thesis}

### 2011-08-24 thesis/thesis.tex
Larger font and double-spaced, more like the UC thesis requirements (ucthesis.cls).

@@
-\documentclass[10pt,a4paper]{report}
+\documentclass[11pt]{report}
 \usepackage[latin1]{inputenc}
@@
 \begin{center}
-UNIVERSITY OF CALIFORNIA\\[0.5cm]
+UNIVERSITY OF CALIFORNIA\\[0.4cm]
 SANTA CRUZ\\[0.75cm]
-\textbf{Simulation Learning in Computer Hex*}\\[0.5cm]
+\textbf{Simulation Learning in Computer Hex*}\\[0.4cm]
 \small{A report submitted in partial satisfaction}\\
-\small{of the requirements for the degree of}\\[0.5cm]
-MASTER OF SCIENCE\\[0.5cm]
+\small{of the requirements for the degree of}\\[0.4cm]
+MASTER OF SCIENCE\\[0.4cm]
 in\\[0.5cm]
-COMPUTER SCIENCE\\[0.5cm]
-by\\[0.5cm]
-\textbf{James Pettit}\\[0.5cm]
+COMPUTER SCIENCE\\[0.4cm]
+by\\[0.4cm]
+\textbf{James Pettit}\\[0.4cm]
 \today
@@
 \tableofcontents
+
+\newpage
+
+\def\dsp{\def\baselinestretch{1.5}\large\normalsize}
+\dsp
 

### 2011-08-25 thesis/thesis.tex
Fixed caption labels and 0's placed from spell checker.

@@
 This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov.
-These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for 0 of moves.
+These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves.
 Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice.
@@
 \black{} attempts to connect top to bottom, \white{} left to right.
-Figure \ref{fig:13x13inprogress} shows an example game in progress on a 9x9 board.
+Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board.
 Figure \ref{fig:13x13finished} shows a different game, where \black{} has just played the winning move, with an unbroken chain connecting top to bottom.
@@
 Gabor Melis' \emph{Six} is the previous world champion (up to 2007).
-Six uses traditional alpha-beta search with and evaluation function based on an 0 circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
+Six uses traditional alpha-beta search with and evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
 \emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves.
@@
 	\includegraphics[width=200px]{graphics/13x13-in-progress.png}
+	\end{center}
 	\caption{13x13 Hex Game, \white{} to play.}
-	\end{center}
 	\label{fig:13x13inprogress}
@@
 	\includegraphics[width=200px]{graphics/13x13-finished.png}
+	\end{center}
 	\caption{Finished 13x13 Hex Game, \black{}'s play at \loc{b6} gives him an unbroken path from top to bottom.}
-	\end{center}
 	\label{fig:13x13finished}
@@
 
-All nodes in the tree 0 to a fixed position, i.e.
+All nodes in the tree correspond to a fixed position, i.e.
 the occupancy state of the board and which player is next to move.
@@
 A simulation is run from the position of this leaf node and the result used to update all nodes along the path.
-The algorithm continues for some set number of iterations, each iteration 0 to one simulation.
+The algorithm continues for some set number of iterations, each iteration finishing with one simulation.
 
@@
 Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move.
-The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a 0 choice between several candidates.
+The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates.
 Only when none of the rules match do they fall back to a uniform random move.
@@
 \caption{Differential equations for $\tau$, $\sigma$, and $\bf{p}$, indexed by time $t$. $D$ is the number of dimensions, i.e. $|\bf{p}|$}
-\end{figure}\label{mutate}
+\label{mutate}
+\end{figure}
 
@@
 	\includegraphics[width=300px]{graphics/local-pattern.png}
+	\end{center}
 	\caption{A 5x5 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
-	\end{center}
 	\label{fig:localpattern}
@@
 	\includegraphics[width=300px]{graphics/weight-pattern-map.png}
+	\end{center}
 	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
-	\end{center}
 	\label{fig:encoding}

### 2011-08-25 src/gtp.go

@@ -102,6 +102,13 @@ func GTP(config *Config) {
 				fail = true
 			}
 			config.Size = boardsize
+			t = NewTracker(config)
+			color = WHITE
+			passcount = 0
+			movecount = 0
+			game_over = false
+			book = config.book
+			root = nil
 		case "clear_board":
 			t = NewTracker(config)
 			color = WHITE

### 2011-08-28 thesis/thesis.tex
A few grammatical errors and clean-ups. Added todo's for the few major additions required.

@@
 \section{Introduction}
-Abstract strategy games provide an easy testbed for artificial intelligence and machine learning applications.
+Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications.
 The game framework provides immediate feedback on algorithm performance.
@@
 
+This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
+% TODO: rephrase, introduce Hivemind
+
 \section{Background: The Game of Hex}
@@
 Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board.
-Figure \ref{fig:13x13finished} shows a different game, where \black{} has just played the winning move, with an unbroken chain connecting top to bottom.
+Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move, with an unbroken chain connecting top to bottom.
 Hex has properties that make it attractive for AI research.
 Once placed, a piece never leaves the board.
-The only criterion for a legal move is that the point be unoccupied (empty).
+The only criterion for a legal move is that the hexagon be unoccupied (empty).
 Gale '79 proved the game cannot end in a draw, simplifying minimax methods.
@@
 Gabor Melis' \emph{Six} is the previous world champion (up to 2007).
-Six uses traditional alpha-beta search with and evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
+Six uses traditional alpha-beta search with an evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)\cite{anshelevich2002hierarchical}.
 \emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves.
@@
 Hex has simpler rules than Go, but has a similar branching factor and complexity.
-Much of this thesis will use Hex as a simpler analog for Go, with the techniques discussed being portable for use in any large branching game without easily crafted heuristics.
+Much of this thesis will use Hex as a simpler analog for Go, with the techniques discussed being useful for any large branching game played on a regular board.
 
@@
 \section{Overview of the Thesis}\label{overview}
-In Chapter \ref{mcts} we will start with the current state of the art in Monte Carlo Tree Search (MCTS).
+In Chapter \ref{mcts} we will review the current state of the art in Monte Carlo Tree Search (MCTS).
 MCTS is a technique to stochastically sample paths through a game-tree.
@@
 Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.
+% TODO: enumerate and define 3 different policy types. Possibly with graphic.
 
@@
 
-function step(node) returns outcome:
+function play_from(node) returns outcome:
   if node.visits < threshold:
@@
   else:
-    next = select(node)
-    outcome = step(next)
+    child = select(node)
+    outcome = play_from(child)
   if outcome is win for node's player:
@@
 \end{verbatim}
+% TODO: figure with caption
 
@@
 The \texttt{select} function defines the selection policy.
-It must return the child of the given node the step into next.
+It must return the child of the given node to step into next.
 In general, selection must trade off between \emph{exploitation} and \emph{exploration}.
@@
 
-One function that balances this trade-off is Upper-Confidence for Trees (UCT).
+One function that balances this trade-off is the Upper-Confidence for Trees algorithm (UCT).
 UCT incorporates the win-rate of a child with a term that increases in value as the ratio of the child's visits to its parents decreases.
@@
 
+The UCT value of a child node is:
 \[
-	\texttt{winrate} + C*\sqrt{\frac{\log{(\texttt{parent.visits})}}{\texttt{child.visits}}}
+	\texttt{child.winrate} + C*\sqrt{\frac{\log{(\texttt{parent.visits})}}{\texttt{child.visits}}}.
 \]
 
-$C$, the \emph{exploration coefficient}, can be varied to increase or decrease the level of exploration versus exploitation.
+The \emph{exploration coefficient}, $C$, can be varied to increase or decrease the level of exploration versus exploitation.
 Note that a very low value will yield a function that relies almost exclusively on the estimated win-rate of a position, while a high value will visit nodes more equally, placing less emphasis on the estimated value.
@@
 The simulation policy has the largest effect on the strength of the entire search.
+%TODO: justify
 Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move.
@@
 Once a simulation is finished, the result is used to update the tree statistics.
-In the basic implementation, only the win/visit statistics along the traversed path are updated.
+In the basic implementation, only the win and visit statistics along the traversed path are updated.
 It was quickly discovered that more sophisticated update strategies could be used to keep alternate statistics, and update them at nodes similar to those used in the playout that were not in the path.
@@
 Thus, a move deep in the simulation could be fictitiously referred to as the ``first'' move.
-In addition to updating the statistics of the nodes in the path, all siblings of the path nodes are also updated with a fictitious AMAF win/visit, if they participated in the simulation.
+In addition to updating the statistics of the nodes in the path, all siblings corresponding to moves made in the simulation are also updated with a fictitious AMAF win and visit.
 Variations on AMAF have been tested, including Permutations AMAF, which updates node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
@@
 
-Other blending functions include Cutoff and Rapid Action Value Estimation (RAVE).
+Other blending functions include Cutoff and Rapid Action Value Estimation (RAVE)\cite{chaslot2008progressive}.
 Helmbold and Parker-Wood (2009) have a good summary of the various update policies and blending methods, along with experimental results showing marked improvement over the baseline of pure UCT\cite{helmbold2009all}.
@@
 Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}.
-However, this frequently requires expert knowledge (heuristics) that can be hard to obtain or not available for some games (particularly new games).
-It was also found that improving the performance of playout policy did not guarantee an improvement in the tree search.
+However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games).
+They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search.
 For example, the playout policy can implicitly prune good moves, never allowing them to be played.
-In Go, the policy must correctly handle complicated situations like nakade.
+In Go, the policy must correctly handle complicated situations like nakade (life and death).
 If it fails to correctly play out a complex position, the tree might fail to see an opponent response.
@@
 Problematically, small changes in overall algorithm strength require thousands of games of testing to detect.
-Many times the improvement made by a small change in weight can be so small is will go unnoticed.
+Many times the improvement made by a small change in weight can be so small it will go unnoticed.
 Automatic weight tuning can improve this process, but still requires thousands of games to test.
@@
 \section{Solution}\label{solution}
+% TODO: We consider simulation policies of the following forms: default, uniform local, uniform local (tenuki), learned local
 After each move, a random move is selected in the 6-connected local area.
@@
 In the uniform local variation, a response is chosen uniformly from any empty moves in the local area.
-The uniform local variation with tenuki is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy.
+The uniform local variation with tenuki (play away) is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy.
 The learned local variation assigns a weight to each empty local vertex based on the local area.
@@
 	\sigma_{t} = & \sigma_{t-1} e^{N(0, \tau^2)} \\
-	p_{i_{t}} = & p_{i_{t-1}} N(0, \sigma^2)
+	p_{i, t} = & p_{i, t-1} N(0, \sigma_{t}^2)
 \end{flalign*}
@@
 Evolution Strategies evolves a set of real numbered values that are used to bias the playouts.
-In most applications, the set of values is simply an array, and the array directly used in fitness evaluation.
+In most applications, the set of values is simply an array, and the array is used directly in fitness evaluation.
 To apply ES to simulation biasing, an encoding is used that maps from a local pattern to a single 32-bit integer.
@@
 In practice, weights are not initialized until a pattern lookup is performed (lazy lookup).
-Every generation, each individual plays $n$ complete games against another random individual.
-At the beginning of the game, one of the two individuals is randomly chosen to play \black, the other to play \white.
-The swap-safe rule was used to play the first move to further equalize first-mover advantage.
+Every generation, each individual plays $n$ complete games, each game played against a random individual.
+Competitors are drawn with replacement from the child population, so a given individual might play the same competitor more than once each generation.
+At the beginning of each game, one of the two individuals is randomly chosen to play \black, the other to play \white.
+A swap-safe move was used to play the first move to further equalize first-mover advantage.
 The winner of a game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.
+
+\subsection{The Hivemind System}
 
@@
 A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
-Experiments were ran using the Gomill software\cite{gomill}.
+Matches were run using the Gomill software\cite{gomill}.
 Gomill required no modification to work with Hex playing programs.
@@
 Instead of just using the best individual from the last generation, the weight vector used is a linear combination of the weights from all individuals in generation 41.
-The weights are blended based on relative fitness, so the final weight vector will most reflect the top performer of generation 41, but also includes the weights off all individuals.
+The weights are blended based on relative fitness, so the final weight vector will most reflect the top performer of generation 41, but also includes the weights of all individuals.
 In some respect all individuals have ``survived'' 41 generations of recombination and mutation, and so were included in the final learned vector.
+% TODO: test combination against single best individual
 
 \begin{itemize}
-\item uniform local vs. control: 65.00\%
-\item uniform local with tenuki vs. control: 58.50\%
-\item learned local vs. control: 81.50\%
+\item uniform local vs. pure random: 65.00\%
+\item uniform local with tenuki vs. pure random: 58.50\%
+\item learned local vs. pure random: 81.50\%
 \item uniform local with tenuki vs. uniform local: 45.50\%
@@
 \end{itemize}
+% TODO: tabular form
 
@@
 \begin{itemize}
-\item learned local vs. MoHex: 6.5\%, CPU=6.64x
-\item control vs. MoHex: 0\% CPU=5.55x
-\item learned local vs. control 92.5\% CPU=1.31x
+\item learned local vs. MoHex: 6.5\%, learned local CPU = 6.64 x MoHex
+\item pure random vs. MoHex: 0\% pure random CPU = 5.55 x MoHex
+\item learned local vs. pure random 92.5\% learned local CPU = 1.31 x pure random
 \end{itemize}

### 2011-08-29 src/config.go
Added option to generate best particle from combination of given generation.

@@ -45,6 +45,7 @@ type Config struct {
 	Lambda       uint
 	Samples      uint
 	Propagate    uint
+	Combine      bool
 	
 	// Load/save different modules
 	Prefix string
@@ -129,6 +130,7 @@ func NewConfig() *Config {
 	flag.UintVar(&config.Lambda, "lambda", 50, "(Training) Children")
 	flag.UintVar(&config.Samples, "samples", 7, "(Training) Evaluations per generation")
 	flag.UintVar(&config.Propagate, "prop", 2, "(Training) Propagate prop best from last generation")
+	flag.BoolVar(&config.Combine, "combine", false, "(Training) Use combination of all particles to form best")
 	
 	flag.StringVar(&config.Prefix, "prefix", "", "Prefix to use when saving file")
 	flag.StringVar(&config.Sfile, "sfile", "", "Load swarm from file")

### 2011-08-29 src/learning.go

@@ -247,28 +247,32 @@ func (s *Swarm) mutate(p *Particle) {
 	}
 }
 
-func (s *Swarm) Best() *Particle {
-	best := NewParticle(s, s.Particles[0].Min, s.Particles[0].Max)
-	best.Fitness = 0
-	superset := make(map[uint32]bool)
-	for i := range s.Particles {
-		for j := range s.Particles[i].Position {
-			superset[j] = true
-		}
-		best.Fitness += s.Particles[i].Fitness
+func (s *Swarm) Best() (best *Particle) {
+  if s.config.Combine {
+	  best = NewParticle(s, s.Particles[0].Min, s.Particles[0].Max)
+	  best.Fitness = 0
+	  superset := make(map[uint32]bool)
+	  for i := range s.Particles {
+		  for j := range s.Particles[i].Position {
+			  superset[j] = true
+		  }
+		  best.Fitness += s.Particles[i].Fitness
+	  }
+	  best.Fitness /= float64(len(s.Particles))
+	  for i := range superset {
+		  count := 0.0
+		  for j := range s.Particles {
+			  if _, exists := s.Particles[j].Position[i]; exists {
+				  best.Position[i] += s.Particles[j].Fitness * s.Particles[j].Position[i]
+				  count += s.Particles[j].Fitness
+			  }
+		  }
+		  best.Position[i] /= float64(count)
+	  }
+	} else {
+	  best = s.Particles[0]
 	}
-	best.Fitness /= float64(len(s.Particles))
-	for i := range superset {
-		count := 0.0
-		for j := range s.Particles {
-			if _, exists := s.Particles[j].Position[i]; exists {
-				best.Position[i] += s.Particles[j].Fitness * s.Particles[j].Position[i]
-				count += s.Particles[j].Fitness
-			}
-		}
-		best.Position[i] /= float64(count)
-	}
-	return best
+	return
 }
 
 func (s *Swarm) SaveSwarm() {

### 2011-09-22 thesis/thesis.tex
Prettied up some pseudocode with the listings package.

@@
 \usepackage{graphicx}
+\usepackage{listings}
 \usepackage{url}
@@
 
-\section{Overview of the Thesis}\label{overview}
+\section{Overview of the Thesis}
 In Chapter \ref{mcts} we will review the current state of the art in Monte Carlo Tree Search (MCTS).
@@
 
-\begin{verbatim}
-function search(root):
+\begin{lstlisting}[float,frame=single,language=Python,caption=MCTS Algorithm Pseudocode]
+def search(root):
   for i in range(simulations):
@@
 
-function play_from(node) returns outcome:
+def play_from(node) returns outcome:
   if node.visits < threshold:
@@
     outcome = play_from(child)
-  if outcome is win for node's player:
+  if outcome is win for node.player:
     node.wins++
@@
   return outcome
-\end{verbatim}
-% TODO: figure with caption
+\end{lstlisting}
 

### 2011-10-17 thesis/thesis.tex
Improved results tables.

@@
 \usepackage{url}
+\usepackage{havannah}
 \newcommand{\black}{\texttt{Black}}
@@
 
-This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
-% TODO: rephrase, introduce Hivemind
+The Hivemind system uses a slightly different approach.
+It directs a probabilistic (Monte Carlo) tree search with learned heuristics.
+The problem of requiring a strong outside player or expert knowledge is solved using evolutionary learning from games of self-play.
+Hivemind provides a framework for Monte Carlo Tree Search, with implementations that can play Hex and Go.
 
@@
 \section{State of the Art in Monte Carlo Tree Search}\label{mcts}
+
+\begin{figure}
+	\begin{center}
+	\begin{HexBoard}[board size=5]
+		\HGame{a3,c2}
+	\end{HexBoard}
+	\includegraphics[width=300px]{graphics/tree.png}
+	\end{center}
+	\caption{Partial MC Search Tree}
+	\label{fig:tree}
+\end{figure}
+
 As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS).
@@
 At any point, the search can be stopped and the current state of the tree used to select a move to play.
-Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.
-% TODO: enumerate and define 3 different policy types. Possibly with graphic.
+Performance of the search is primarily based on three algorithms.
+\begin{itemize}
+\item{The \emph{selection} policy recursively traverses the tree, selecting good-looking moves for exploration.}
+\item{The \emph{simulation} policy randomly or semi-randomly ``plays out'' the game from the current position until a final position.}
+\item{The \emph{update} policy modifies the win/loss statistics of nodes based on the simulation results.}
+\end{itemize}
+Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position.
+In the example, the selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}.
+The simulation policy ran, playing out the board with \texttt{E4} occupied by \black\ and \texttt{A1} occupied by \white.
+This resulted in a loss for \black.
+The update policy choose to update all nodes along the visited path, incrementing the wins for \white\ moves and incrementing the losses for \black\ moves.
 
@@
 The simplest policy plays a legal uniform random move until the game ends (in Hex, when one player has a winning path).
-The simulation policy has the largest effect on the strength of the entire search.
-%TODO: justify
-Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move.
+The simulation policy has a large effect on the strength of the entire search and, for Go, has a large body of work dedicated to improving it.
+Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move\cite{chaslot2010adding}.
 The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates.
 Only when none of the rules match do they fall back to a uniform random move.
-It was quickly found that building a stronger simulation policy does not guarantee better performance when used in the full tree search.
+It was quickly found that building a stronger simulation policy does not guarantee better performance when used in the full tree search\cite{gelly2006modification}.
 A test and check methodology is required, with changes to the simulation policy having unpredictable effects on overall performance.
@@
 \section{Problem Statement}\label{problem}
-MoGo, and Go programs in general, have the benefit of a rich history of play.
+MoGo specifically, and Go programs in general, have the benefit of a rich history of play.
 Thousands of expert games are available for analysis.
@@
 \section{Solution}\label{solution}
-% TODO: We consider simulation policies of the following forms: default, uniform local, uniform local (tenuki), learned local
-After each move, a random move is selected in the 6-connected local area.
-This move is used as the next move in the playout.
-All subsequent moves will also have their local area examined for a random response.
-If all 6 connected vertices are non-empty, a move is played using the uniform random policy (the \emph{default} policy).
-In the uniform local variation, a response is chosen uniformly from any empty moves in the local area.
-The uniform local variation with tenuki (play away) is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy.
-The learned local variation assigns a weight to each empty local vertex based on the local area.
-These weights are used to stochastically bias the local response.
-The weights are trained over time through self-play using Evolution Strategies.
+The simulation policy has the task of quickly selecting a legal move from an arbitrary position.
+The policy is applied iteratively until the game is finished.
+Ideally, several thousand games are simulated each second, so the simulation policy must be relatively simple and fast, compared to the full tree search.
+We consider simulation policies of the following forms:
+\begin{itemize}
+	\item{Default} - The next move is chosen uniformly from the set of legal moves.
+	\item{Uniform local} - The 6-connected (``local'') points surrounding the previous move are examined.
+The next move is chosen uniformly from any legal moves in this set.
+If no legal move exists on these 6 points, the default policy is used.
+	\item{Uniform local with ``tenuki''} - The tenuki (Japanese: ``play away'') variant includes a small probability ($p=\frac{1}{6}$) of using the default policy, even if a legal local move exists.
+	\item{Learned local} - Each local point is matched against a pattern database and assigned a weight.
+The weight is used as the probability of playing on the given point.
+Weights are trained over time through self-play using Evolution Strategies.
+\end{itemize}
+
+When used at part of a full Monte-Carlo Tree Search, the learned local policy consistently beats both the default policy and the non-learned, uniform local policies.
+It carries a small cost of computing the pattern hash and looking up weights in the pattern database, but most of the computation is done with offline learning.
+It is self-bootstrapping, using Evolution Strategies and self-play instead of a strong outside tutor or expert knowledge.
 
@@
 	\includegraphics[width=300px]{graphics/local-pattern.png}
-	\end{center}
 	\caption{A 5x5 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
 	\label{fig:localpattern}
+	\end{center}
 \end{figure}
@@
 	\includegraphics[width=300px]{graphics/weight-pattern-map.png}
-	\end{center}
 	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
 	\label{fig:encoding}
+	\end{center}
 \end{figure}
@@
 \subsection{The Hivemind System}
+% TODO: Hivemind system
 
@@
 
-\begin{itemize}
-\item uniform local vs. pure random: 65.00\%
-\item uniform local with tenuki vs. pure random: 58.50\%
-\item learned local vs. pure random: 81.50\%
-\item uniform local with tenuki vs. uniform local: 45.50\%
-\item learned local vs. uniform local: 76.50\%
-\item learned local vs. uniform local with tenuki: 79.00\%
-\end{itemize}
-% TODO: tabular form
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c | c c c}
+		 & default & uniform & uniform (tenuki) \\
+		\hline
+		uniform & 65\% & & \\
+		uniform (tenuki) & 58.5\% & 45.5\% & \\
+		learned & 81.5\% & 76.5\% & 79\% \\
+		\end{tabular}
+	\end{center}
+	\caption{All-play-all tournament of the 4 Hivemind variants.}
+	\label{fig:results}
+\end{figure}
 
@@
 As a result, the time controls are not directly comparable.
-However, CPU time was tracked and can be used as another performance comparison.
+However, CPU time was tracked and can be used for a cost comparison.
 
-\begin{itemize}
-\item learned local vs. MoHex: 6.5\%, learned local CPU = 6.64 x MoHex
-\item pure random vs. MoHex: 0\% pure random CPU = 5.55 x MoHex
-\item learned local vs. pure random 92.5\% learned local CPU = 1.31 x pure random
-\end{itemize}
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c c c c}
+		Player 1 & Player 2 & Win-rate & Relative CPU \\
+		\hline
+		Hivemind Default & MoHex & 6.5\% & 664\% \\
+		Hivemind Learned & MoHex & 0\% & 555\% \\
+		Hivemind Learned & Hivemind Default & 92.5\% & 131\% \\
+		\end{tabular}
+	\end{center}
+	\caption{Win-rate and relative CPU used of Hivemind versus MoHex. Relative CPU is the ratio of average CPU time per move for Player 1 vs. Player2.}
+	\label{fig:relativecpu}
+\end{figure}
 
@@
 Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength.
+In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results.
+This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 81.5\% with 10,000 moves.
+The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.
 

### 2011-10-17 thesis/thesis.tex
Changed from png graphics to havannah.sty board defs.

@@
 The object of the game is to form an unbroken chain from one side of the board to the other.
-\black{} attempts to connect top to bottom, \white{} left to right.
 Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board.
-Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move, with an unbroken chain connecting top to bottom.
+Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move, with an unbroken chain connecting both his sides.
 Hex has properties that make it attractive for AI research.
@@
 \begin{figure}
-	\begin{center}
-	\includegraphics[width=200px]{graphics/13x13-in-progress.png}
-	\end{center}
+	\begin{HexBoard}[board size=13]
+		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
+					 h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
+	\end{HexBoard}
 	\caption{13x13 Hex Game, \white{} to play.}
@@
 \end{figure}
+
 \begin{figure}
-	\begin{center}
-	\includegraphics[width=200px]{graphics/13x13-finished.png}
-	\end{center}
+	\begin{HexBoard}[board size=13]
+		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
+				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
+				   e8,e9,c10,k8,c9,g6,g7,c11,
+				   d10,h6,h7,i5,j5,j4,k4,k3,
+				   m2,l3,m3,l4,c6,e11,l5,m4,
+				   d11,b7,e6,d12,c12,c7,e7,f5,
+				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
+				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
+				   c13,c5,b6}
+	\end{HexBoard}
 	\caption{Finished 13x13 Hex Game, \black{}'s play at \loc{b6} gives him an unbroken path from top to bottom.}

### 2011-11-03 thesis/thesis.tex
Swapped out results to use the best individual instead of the combined.

@@
 
-\subsection{Evolution Strategies}
+\subsection{Evolution Strategies}\label{estrats}
 Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors.
@@
 \subsection{The Hivemind System}
-% TODO: Hivemind system
+Hivemind was designed for experiments in learning to play abstract strategy games on a regular board.
+The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface''.
+The tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe.
+It is designed to be possible to implement many different games in the tracker, abstracting away the learning and search process.
+Hivemind is open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
+
+\subsubsection{Learning}
+The learning module is responsible for driving the evolutionary search process.
+Currently, Hivemind uses Evolution Strategies (see Section \ref{estrats}), although previous versions experimented with Particle Swarm Optimization.
+As described earlier, the learning process consists of three steps, iterated once per generation: selection, mutation, and evaluation.
+In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that generates optimal results during the evaluation step.
+Evaluation is performed by selecting two individuals from the population and having them play a game against each other, each individual parameterized by their strategy.
+The individual that wins receives points.
+During the selection step, individuals are culled based on points received, and the process continues.
+To play a game, the learning module iteratively calls into the search module with the strategy, until the tracker module reports the game is over.
+The tracker module is also used to collect the game results.
+
+\subsubsection{Search}
+The search module is a simple generic implementation of Monte-Carlo Tree Search, using the UCT algorithm to traverse the tree.
+Section \ref{mcts} has a description of the MCTS process.
+The search module uses the Game Tracker to expand the tree and perform the random playouts.
+The entire source code for search is only about 600 lines of code.
+Using some extra information from the Game Tracker, search is configurable to use RAVE and AMAF heuristics so a single playout can provide information about the value of multiple nodes.
+
+\subsubsection{Game Tracker}
 
@@
 This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz.
-Although evolution strategies is easy to parallelize by running multiple evaluations simultaneously, the current implementation is single-threaded.
-After training was completed, the 41st generation was used to form a single individual.
-This final individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally.
+After training was completed, the best individual from the 41st generation was selected for comparison.
+This individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally.
 The control consisted of uniform random playouts.
@@
 
-Instead of just using the best individual from the last generation, the weight vector used is a linear combination of the weights from all individuals in generation 41.
-The weights are blended based on relative fitness, so the final weight vector will most reflect the top performer of generation 41, but also includes the weights of all individuals.
-In some respect all individuals have ``survived'' 41 generations of recombination and mutation, and so were included in the final learned vector.
-% TODO: test combination against single best individual
-
 \begin{figure}
@@
 		\hline
-		uniform & 65\% & & \\
-		uniform (tenuki) & 58.5\% & 45.5\% & \\
-		learned & 81.5\% & 76.5\% & 79\% \\
+		uniform & 63.50\% & & \\
+		uniform (tenuki) & 64.00\% & 50.00\% & \\
+		learned & 79.50\% & 81.50\% & 76.00\% \\
 		\end{tabular}
@@
 \end{figure}
+% TODO: run tests using best instead of combined
 

### 2011-11-04 thesis/thesis.tex
Added Fuego reference, and placeholder for content in Hivemind system detail section.
Working on the Hivemind description.

@@
 	\begin{HexBoard}[board size=13]
-		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
-					 h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
+		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
 	\end{HexBoard}
@@
 
-As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS).
+As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS)\cite{mohex}.
 The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}.
@@
 The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface''.
-The tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe.
-It is designed to be possible to implement many different games in the tracker, abstracting away the learning and search process.
-Hivemind is open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
+The design was influenced by Fuego, although Hivemind is written in a different language (Go).
+In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games\cite{Fuego}.
+Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe.
+The major contribution of Hivemind is the integration of evolutionary learning into the search process.
+Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
 
@@
 The entire source code for search is only about 600 lines of code.
-Using some extra information from the Game Tracker, search is configurable to use RAVE and AMAF heuristics so a single playout can provide information about the value of multiple nodes.
+Using some extra information from the Game Tracker, search is configurable to use RAVE and AMAF heuristics to update multiple nodes using the results of a single playout.
 
@@
 \end{figure}
-% TODO: run tests using best instead of combined
 

### 2011-11-04 thesis/thesis.tex
Fleshing out Hivemind design section.

@@
 In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that generates optimal results during the evaluation step.
-Evaluation is performed by selecting two individuals from the population and having them play a game against each other, each individual parameterized by their strategy.
+Evaluation is performed by selecting two individuals from the population and having them play a game against each other, each individual parameterized by their policy weights.
 The individual that wins receives points.
 During the selection step, individuals are culled based on points received, and the process continues.
-To play a game, the learning module iteratively calls into the search module with the strategy, until the tracker module reports the game is over.
-The tracker module is also used to collect the game results.
+To play a game, the learning module calls into the search module with the current tracker state.
+The search is parameterized by the policy weights.
+Once the tracker reports the game is over, the result (win/lose) is collected for use in the next selection step.
 
@@
 \subsubsection{Game Tracker}
+ToDo: Tracker both abstracts ``game'' notion to allow many different games with same learning and seach, and by defining what a ``game'' is, limits the scope of possible games.
+What games are possible with Hivemind (e.g.
+perfect info, 2-player, regular board), and what is off-limits/difficult (multiple players?
+stochastic info?).
+
+The game tracker has four primary responsibilities.
+\begin{enumerate}
+\item{Maintain required game state. A single function, \texttt{Play(color, vertex)}, is used by the search tree to apply a move for the given color.}
+\item{Report legal moves from a given game state.
+The \texttt{Legal(color, vertex)} function is used to query the legality of a given move.
+The search tree uses this to expand a parent node, adding all legal moves as children.}
+\item{Report on the status of the winner. For Hex, this is the special \texttt{Empty} player while the game is in progress, and either \black{} or \white{} once a connected path has been achieved.}
+\item{Playing out the game from the current game state.
+The tracker is responsible for the exact semantics of the policy weights, and for using them during the playout.
+The speed of the playout function is also critical, as more playouts-per-second generally result in better move selection.}
+\end{enumerate}
+The additional \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
+A few extra functions also provide for debugging by pretty-printing the tracker's internal state.
 
@@
 Learned local play is the clear winner.
-Interestingly, the addition of a tenuki probability lowers performance in all cases.
+Interestingly, the addition of a tenuki probability has no or little effect on performance in all cases.
 As expected, uniform local play significantly improves upon pure random playouts.

### 2011-11-28 thesis/thesis.tex
Cleaned up language and grammar.

--- 2011-11-04 thesis/thesis.tex
@@
 Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves.
-This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov.
+This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov.
 These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves.
@@
 Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons.
-The players are denoted by colors here \black{} and \white.
+The players are denoted by  a color, either \black{} and \white.
 The object of the game is to form an unbroken chain from one side of the board to the other.
@@
 In addition to updating the statistics of the nodes in the path, all siblings corresponding to moves made in the simulation are also updated with a fictitious AMAF win and visit.
-Variations on AMAF have been tested, including Permutations AMAF, which updates node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
+Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
 
@@
 If it fails to correctly play out a complex position, the tree might fail to see an opponent response.
-Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situation.
+Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situations.
 If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence.
@@
 Hivemind was designed for experiments in learning to play abstract strategy games on a regular board.
-The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface''.
-The design was influenced by Fuego, although Hivemind is written in a different language (Go).
+The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface'' used by the tree search to maintain game state information.
+The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go\cite{golang}).
 In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games\cite{Fuego}.
@@
 
-\subsubsection{Learning}
+\subsubsection{Learning}\label{learning}
 The learning module is responsible for driving the evolutionary search process.
-Currently, Hivemind uses Evolution Strategies (see Section \ref{estrats}), although previous versions experimented with Particle Swarm Optimization.
-As described earlier, the learning process consists of three steps, iterated once per generation: selection, mutation, and evaluation.
-In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that generates optimal results during the evaluation step.
-Evaluation is performed by selecting two individuals from the population and having them play a game against each other, each individual parameterized by their strategy.
-The individual that wins receives points.
-During the selection step, individuals are culled based on points received, and the process continues.
-To play a game, the learning module iteratively calls into the search module with the strategy, until the tracker module reports the game is over.
-The tracker module is also used to collect the game results.
+In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that performs well during the evaluation step.
+Hivemind currently uses Evolution Strategies, although previous versions experimented with Particle Swarm Optimization.
+Selection and mutation are implemented as described in section \ref{estrats}.
+Evaluation is described in section \ref{fitness}.
 
@@
 \subsubsection{Game Tracker}
+The game tracker allows complete games to be played vi Monte-Carlo Tree search. The implementation of a game has four responsibilities:
+\begin{enumerate}
+\item{Maintain all required game state. A single function, \texttt{Play(color, vertex)}, is used by the search tree to apply a move for the given color.}
+\item{Report legal moves from a given game state.
+The \texttt{Legal(color, vertex)} function is used to query the legality of a given move.
+The search tree uses this to expand a parent node, adding all legal moves as children.}
+\item{Report on the status of the winner.
+The \texttt{Winner()} function is used to determine the current winner of the game.
+The special \texttt{Empty} player is returned while the game is in progress, and either \black{} or \white{} once a winner has been determined.}
+\item{Playing out the game from the current game state.
+The tracker is parameterized with the policy weights, and responsible for using them during the playout.
+The speed of the playout function is also critical, as more playouts-per-second generally result in better move selection.
+This emphasis on speed is why the tracker is required to implement a monolithic \texttt{Playout} function.
+Many games have optimizations that can be performed in the playout that the search tree is unaware of, for example, the Hex tracker keeps a simple list of randomized empty positions, so random move selection is a O(1) operation.
+This is because in Hex, any empty position is legal.
+This is not the case for Go, so the Go engine uses other optimizations for fast random move selection.}
+\end{enumerate}
+The additional \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
+A few extra functions also provide for debugging by pretty-printing the tracker's internal state.
+
+The game tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework.
+It also constrains the space of games that can be easily implemented in Hivemind.
+In particular, Hivemind allows for 2 player games with perfect information.
+The game space is not required to be a regular board.
+Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children.
+Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.
 
 \section{Results and Conclusions}\label{results}
-To test the effects of learning policy weights, the evolution process was ran for 41 generations.
-This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz.
-After training was completed, the best individual from the 41st generation was selected for comparison.
+To test the effects of learning policy weights, the evolution process was run for 100 generations.
+The Hivemind learning parameters used were: \texttt{generations=100, lambda=40, mu=30, parents=2, samples=5}.
+See sections \ref{estrats} and \ref{learning} for a description of the learning parameters.
+Each game was played with boardsize of 7x7 and 1000 playouts per move.
+After 100 generations were complete, the best individual from the 100th generation was selected for comparison.
 This individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally.
@@
 All players have approximately the same performance with regard to simulations per second, so the 5 seconds per move resulted in around 10,000 simulated playouts.
-A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
+A boardsize of 7x7 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
 Matches were run using the Gomill software\cite{gomill}.
 Gomill required no modification to work with Hex playing programs.
+Results are shown in figure \ref{fig:results}.
 
@@
 		\hline
-		uniform & 63.50\% & & \\
-		uniform (tenuki) & 64.00\% & 50.00\% & \\
-		learned & 79.50\% & 81.50\% & 76.00\% \\
+		uniform & 70.50\% & & \\
+		uniform (tenuki) & 61.00\% & 50.00\% & \\
+		learned & 90.00\% & 84.00\% & 86.00\% \\
 		\end{tabular}
+	\caption{All-play-all tournament of the 4 Hivemind variants. Each element is the percent win-rate of the row variant versus the column variant.}
+	\label{fig:results}
 	\end{center}
-	\caption{All-play-all tournament of the 4 Hivemind variants.}
-	\label{fig:results}
 \end{figure}
@@
 
-As an outside control, MoHex from the University of Alberta was compiled and ran with the default settings\cite{mohex}.
+As an outside control, MoHex from the University of Alberta was compiled and run with the default settings\cite{mohex}.
 Again, 200 games were played.
 As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move.
-Default time settings for MoHex are 10 seconds, but MoHex uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found.
+MoHex played with the default time settings of 10 seconds per move, but it uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found.
 The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS.
@@
 However, CPU time was tracked and can be used for a cost comparison.
+Figure \ref{fig:relativecpu} shows the results of this series.
 
@@
 		\hline
-		Hivemind Default & MoHex & 6.5\% & 664\% \\
-		Hivemind Learned & MoHex & 0\% & 555\% \\
+		Hivemind Default & MoHex & 0\% & 555\% \\
+		Hivemind Learned & MoHex & 6.5\% & 664\% \\
 		Hivemind Learned & Hivemind Default & 92.5\% & 131\% \\
 		\end{tabular}
-	\end{center}
 		\caption{Win-rate and relative CPU used of Hivemind versus MoHex. Relative CPU is the ratio of average CPU time per move for Player 1 vs. Player2.}
 		\label{fig:relativecpu}
+	\end{center}
 \end{figure}
@@
 Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played.
-Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength.
+MoHex strongly outperformed Hivemind, but used only a fraction (about 1/6th) of the CPU time, showing a large distance to cover to get Hivemind up to competition strength.
 In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results.
-This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 81.5\% with 10,000 moves.
+This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 90\% with 10,000 moves.
 The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.

### 2011-12-06 thesis/thesis.tex
Added appendix A1

@@
 
-200 games per pair were played, with time controls set at 5 seconds per move.
-All players have approximately the same performance with regard to simulations per second, so the 5 seconds per move resulted in around 10,000 simulated playouts.
+200 games per pair were played, with time controls set at 10,000 playouts per move.
 A boardsize of 7x7 was used, and to control for the swap rule a swap-safe move was used as the first move for black.
@@
 
+\section{Appendix}\label{appendix}
+\subsection{A1 - Stability of Evolution}
+Stochastic processes like evolutionary search can have large variations in their end results.
+To test the variation of Hivemind, 10 independent individuals were produced.
+Each individual was the best individual from the 100th generation of an independent run of the learning process.
+The parameters used for each run were that same as those used for the final results of self-play,  \texttt{generations=100, lambda=40, mu=30, parents=2, samples=5}, with 1000 playouts per move on a 7x7 board.
+To test, each independent individual (Learned 1 through 10) played 200 games on a 7x7 board, against a player using the default policy (Default), all players simulating 10,000 playouts per move.
+
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c c c c}
+		& Default \\
+		\hline
+		Learned 1 & 91.0\% \\
+		Learned 2 & 90.5\% \\
+		Learned 3 & 90.5\% \\
+		Learned 4 & 87.5\% \\
+		Learned 5 & 90.5\% \\
+		Learned 6 & 93.5\% \\
+		Learned 7 & 93.0\% \\
+		Learned 8 & 91.5\% \\
+		Learned 9 & 87.0\% \\
+		Learned 10 & 92.5\% \\
+		Min & 87.0\% \\
+		Max & 93.5\% \\
+		Mean & 90.75\% \\
+		Std. Dev. & ~2.14\% \\
+		\end{tabular}
+		\caption{Results of 10 independent evolutionary processes versus the default policy.}
+		\label{fig:a1results}
+	\end{center}
+\end{figure}
+
+Figure \ref{fig:a1results} shows the results of these games.
+Overall, the 10 individuals had a standard deviation of about 2.14\%, with a mean of 90.75\%.
+This indicated that the evolutionary processes is relatively stable.
+With only a few independent runs of the evolutionary processes, an implementer can reasonably assume that the best individual would perform as well as after several hundred runs.
+
+\subsection{A2 - Effects of different board sizes}
+The computational time required of learning process greatly depends on the boardsize used in learning.
+For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage.
+
 \bibliography{thesis}

### 2011-12-27 thesis/thesis.tex
Better swap-safe moves, only for 7x7 and 11x11 though. Also gofmt'd.

@@
 \subsection{A1 - Stability of Evolution}
-Stochastic processes like evolutionary search can have large variations in their end results.
-To test the variation of Hivemind, 10 independent individuals were produced.
-Each individual was the best individual from the 100th generation of an independent run of the learning process.
-The parameters used for each run were that same as those used for the final results of self-play,  \texttt{generations=100, lambda=40, mu=30, parents=2, samples=5}, with 1000 playouts per move on a 7x7 board.
-To test, each independent individual (Learned 1 through 10) played 200 games on a 7x7 board, against a player using the default policy (Default), all players simulating 10,000 playouts per move.
-
-\begin{figure}
+\begin{figure}[t]
 	\begin{center}
-		\begin{tabular}{c c c c}
+		\begin{tabular}{c c}
 		& Default \\
@@
 		\end{tabular}
-		\caption{Results of 10 independent evolutionary processes versus the default policy.}
+		\caption{Results of 10 independent evolutionary processes versus the default policy. Percentage shown is average win-rate of row player versus column player.}
 		\label{fig:a1results}
@@
 
+Stochastic processes like evolutionary search can have large variations in their end results.
+To test the variation of Hivemind, 10 independent individuals were produced.
+Each individual was the best individual from the 100th generation of an independent run of the learning process.
+The parameters used for each run were that same as those used for the final results of self-play,  \texttt{generations=100, lambda=40, mu=30, parents=2, samples=5}, with 1000 playouts per move on a 7x7 board.
+To test, each independent individual (Learned 1 through 10) played 200 games on a 7x7 board, against a player using the default policy (Default), all players simulating 10,000 playouts per move.
+
 Figure \ref{fig:a1results} shows the results of these games.
@@
 This indicated that the evolutionary processes is relatively stable.
-With only a few independent runs of the evolutionary processes, an implementer can reasonably assume that the best individual would perform as well as after several hundred runs.
+With only a few independent runs of the evolutionary processes, an implementer can reasonably assume that the best individual would perform well.
 
 \subsection{A2 - Effects of different board sizes}
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{c c c c}
+		& Learned \\
+		Default 11x11 & 96.8\% \\
+		Default 13x13 & 100.0\% \\
+		Uniform 11x11 & 82.76\% \\
+		Uniform 13x13 & 92.86\% \\
+		\hline
+		\end{tabular}
+		\caption{Results of evolutionary process learned on a 7x7 board, playing on an 11x11 and 13x13 board. Percentage shown is average win-rate of column player versus row player.}
+		\label{fig:a2results}
+	\end{center}
+\end{figure}
+
 The computational time required of learning process greatly depends on the boardsize used in learning.
 For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage.
+The best individual from the 100th generation of this training was used in matches against the default and uniform policy, but a boardsize of 11x11 and 13x13 was used for the matches.
+Because of the increased boardsize, players used 100,000 simulated playouts per move.
+The computational time required for these games is significantly higher than that required for the 7x7 games, so the results shown in Figure \ref{fig:a2results} are only after 30 games per match.
+
+Even with the low number of games, the results shown in Figure \ref{fig:a2results} are positive.
+Learning on a 7x7 board provides results that can generalize to larger boardsizes.
+This is especially useful because of the exponential increase in computational time required to play games for evolutionary learning when learning on larger board sizes.
 

### 2011-12-27 gui/main.py

@@ -38,7 +38,7 @@ class GUI(object):
 		if len(sys.argv) == 2:
 			self.set_file(sys.argv[1])
 		else:
-			self.set_size(9)
+			self.set_size(7)
 			self.pb = "None"
 			self.pw = "None"
 		

### 2011-12-27 src/asgtypes.go

@@ -1,9 +1,9 @@
 package main
 
 import (
-	"strings"
 	"fmt"
 	"os"
+	"strings"
 	"time"
 )
 
@@ -130,17 +130,17 @@ func StatsBoard(root *Node, t Tracker) (s string) {
 	}
 	for row := 0; row < t.Boardsize(); row++ {
 		for col := 0; col < t.Boardsize(); col++ {
-			v := row * t.Boardsize() + col
+			v := row*t.Boardsize() + col
 			if board[v] == "" {
 				s += "\"\""
 			} else {
 				s += board[v]
 			}
-			if col != t.Boardsize() - 1 {
+			if col != t.Boardsize()-1 {
 				s += " "
 			}
 		}
-		if row != t.Boardsize() - 1 {
+		if row != t.Boardsize()-1 {
 			s += "\n"
 		}
 	}

### 2011-12-27 src/cluster.go

@@ -2,8 +2,8 @@ package main
 
 import (
 	"github.com/etherealmachine/redis.go"
-	"log"
 	"json"
+	"log"
 )
 
 func InitCluster(shutdown chan bool) {
@@ -49,38 +49,38 @@ func process(configs chan Config) {
 	sem := make(chan int, 4)
 	for config := range configs {
 		switch config.MsgType {
-			case "eval":
-				go eval(config, sem)
-			case "play":
-				go play(config, sem)
+		case "eval":
+			go eval(config, sem)
+		case "play":
+			go play(config, sem)
 		}
 	}
 }
 
 func eval(config Config, sem chan int) {
 	/*
-	sem <- 1
-	t := NewTracker(&config)
-	color := BLACK
-	for i := range config.Moves {
-		t.Play(color, config.Moves[i])
-		color = Reverse(color)
-	}
-	root := NewRoot(color, t, &config)
-	genmove(root, t)
-	eval := &Eval{config.Moves, root.Best().Vertex, root.Best().Wins / root.Best().Visits}
-	bytes, err := json.Marshal(eval)
-	if err != nil {
-		log.Println(err)
-	} else {
-		var client redis.Client
-		client.Addr = "127.0.0.1:6379"
-		err := client.Publish("evals", bytes)
+		sem <- 1
+		t := NewTracker(&config)
+		color := BLACK
+		for i := range config.Moves {
+			t.Play(color, config.Moves[i])
+			color = Reverse(color)
+		}
+		root := NewRoot(color, t, &config)
+		genmove(root, t)
+		eval := &Eval{config.Moves, root.Best().Vertex, root.Best().Wins / root.Best().Visits}
+		bytes, err := json.Marshal(eval)
 		if err != nil {
 			log.Println(err)
+		} else {
+			var client redis.Client
+			client.Addr = "127.0.0.1:6379"
+			err := client.Publish("evals", bytes)
+			if err != nil {
+				log.Println(err)
+			}
 		}
-	}
-	<-sem
+		<-sem
 	*/
 }
 
@@ -88,14 +88,14 @@ func play(config Config, sem chan int) byte {
 	sem <- 1
 	t := NewTracker(&config)
 	var vertex int
-	for move := 0;; {
+	for move := 0; ; {
 		config.policy_weights = config.Black_policy_weights
 		br := NewRoot(BLACK, t, &config)
[…]

### 2011-12-27 src/config.go

@@ -2,20 +2,20 @@ package main
 
 import (
 	"flag"
-	"os"
 	"json"
 	"log"
+	"os"
 )
 
 type Config struct {
 	// Modes
-	Help  bool
-	Gtp   bool
-	Book  bool
-	Genmove bool
+	Help     bool
+	Gtp      bool
+	Book     bool
+	Genmove  bool
 	PlayGame bool
-	SGF string
-	Cluster bool
+	SGF      string
+	Cluster  bool
 
 	// Time limits
 	MaxPlayouts uint
@@ -28,25 +28,25 @@ type Config struct {
 	Gfx bool
 
 	// Different games
-	Go       bool
-	Hex      bool
-	HexFast  bool
+	Go      bool
+	Hex     bool
+	HexFast bool
 
 	// Game-specific variables
-	Size int
-	Komi float64
+	Size     int
+	Komi     float64
 	Swapsafe bool
 
 	// Learning
-	Train        bool
-	Generations  uint
-	Mu           uint
-	Parents      uint
-	Lambda       uint
-	Samples      uint
-	Propagate    uint
-	Combine      bool
-	
+	Train       bool
+	Generations uint
+	Mu          uint
+	Parents     uint
+	Lambda      uint
+	Samples     uint
+	Propagate   uint
+	Combine     bool
+
 	// Load/save different modules
 	Prefix string
 	Bfile  string
@@ -55,44 +55,44 @@ type Config struct {
 	Sfile  string
 
 	// Tree exploration/expansion
-	TreeSearch  bool
-	Explore     float64
-	RAVE        float64
-	ExpandAfter float64
-	Var         bool
-	AMAF        bool
-	Neighbors   bool
-	Ancestor    bool
-	Seed        bool
-	PlayoutProbs bool
-	PlayoutSuggest bool
-	PlayoutSuggestUniform bool
+	TreeSearch                  bool
+	Explore                     float64
+	RAVE                        float64
+	ExpandAfter                 float64
+	Var                         bool
+	AMAF                        bool
+	Neighbors                   bool
+	Ancestor                    bool
+	Seed                        bool
+	PlayoutProbs                bool
+	PlayoutSuggest              bool
+	PlayoutSuggestUniform       bool
 	PlayoutSuggestUniformTenuki bool
 
 	// Logging
-	Verbose bool
-	VeryVerbose bool
-	Verify  bool
-	PrintWeights  bool
-	Lfile   string
-	
+	Verbose      bool
[…]

### 2011-12-27 src/gotracker.go

@@ -1,13 +1,13 @@
 package main
 
 import (
-	"fmt"
 	"container/vector"
-	"rand"
-	"strings"
-	"strconv"
+	"fmt"
 	"log"
 	"math"
+	"rand"
+	"strconv"
+	"strings"
 )
 
 // Tracks a game of Go
@@ -16,27 +16,27 @@ import (
 // liberties returns the number of liberties for the chain
 // it is only correct for the root of the set
 type GoTracker struct {
-	boardsize      int
-	sqsize         int
-	parent         []int
-	rank           []int
-	liberties      [][2]uint64
-	board          []byte
-	weights        *WeightTree
-	atari          []map[int]int
-	komi           float64
-	koVertex       int
-	koColor        byte
-	played         []byte
-	adj            [][]int
-	mask           [][4]uint64
-	neighbors      [][][]int
-	passes         int
-	winner         byte
-	superko        bool
-	moves          *vector.IntVector
-	history        *vector.Vector
-	config         *Config
+	boardsize int
+	sqsize    int
+	parent    []int
+	rank      []int
+	liberties [][2]uint64
+	board     []byte
+	weights   *WeightTree
+	atari     []map[int]int
+	komi      float64
+	koVertex  int
+	koColor   byte
+	played    []byte
+	adj       [][]int
+	mask      [][4]uint64
+	neighbors [][][]int
+	passes    int
+	winner    byte
+	superko   bool
+	moves     *vector.IntVector
+	history   *vector.Vector
+	config    *Config
 }
 
 // parent must be initialized so each element is a pointer to itself
@@ -133,7 +133,7 @@ func (t *GoTracker) Copy() Tracker {
 func (t *GoTracker) Play(color byte, vertex int) {
 	if vertex != -1 {
 		t.passes = 0
-		
+
 		if t.koVertex != -1 {
 			t.weights.Set(t.koColor, t.koVertex, INIT_WEIGHT)
 			t.koVertex = -1
@@ -149,12 +149,12 @@ func (t *GoTracker) Play(color byte, vertex int) {
 			cp.Play(color, vertex)
 			t.history.Push(*MakeHash(cp))
 		}
-		
+
 		// modify the board
[…]

### 2011-12-27 src/gtp.go

@@ -1,12 +1,12 @@
 package main
 
 import (
-	"os"
 	"bufio"
 	"fmt"
-	"strings"
-	"strconv"
 	"log"
+	"os"
+	"strconv"
+	"strings"
 )
 
 var supported_commands = `name
@@ -102,13 +102,6 @@ func GTP(config *Config) {
 				fail = true
 			}
 			config.Size = boardsize
-			t = NewTracker(config)
-			color = WHITE
-			passcount = 0
-			movecount = 0
-			game_over = false
-			book = config.book
-			root = nil
 		case "clear_board":
 			t = NewTracker(config)
 			color = WHITE
@@ -164,9 +157,10 @@ func GTP(config *Config) {
 				vertex := -1
 				// HEX, swap-safe: if black and first move of game, play a move that should be safe from swapping
 				if config.Hex && color == BLACK && config.Swapsafe && movecount == 0 {
-					vertex = (3 * t.Boardsize()) + 2
-					// Pass if: no time left, game definitely won
-				} else if config.Timelimit != 0 && t.Winner() == EMPTY && !game_over {
+					vertex = swap_safe(boardsize)
+				}
+				// Pass if: no time left, game definitely won
+				if vertex == -1 && config.Timelimit != 0 && t.Winner() == EMPTY && !game_over {
 					if book != nil {
 						best := book.Best()
 						if best.Visits > 100 {
@@ -201,7 +195,7 @@ func GTP(config *Config) {
 				if book != nil {
 					book = book.Play(color, vertex, t)
 				}
-				if game_over && t.Winner() == Reverse(color) {
+				if (game_over && t.Winner() == Reverse(color)) || (config.Hex && vertex == -1) {
 					res = "resign"
 				} else {
 					res = t.Vtoa(vertex)
@@ -256,7 +250,7 @@ func GTP(config *Config) {
 			}
 			res = TerritoryBoard(value, 1, t)
 		case "legal":
-			res = LegalBoard(t, map[byte]string{BOTH:"green",BLACK:"black",WHITE:"white",EMPTY:"none"})
+			res = LegalBoard(t, map[byte]string{BOTH: "green", BLACK: "black", WHITE: "white", EMPTY: "none"})
 		case "time_settings":
 			main_time, _ = strconv.Atoi(args[1])
 			byo_yomi_time, _ := strconv.Atoi(args[2])

### 2011-12-27 src/hextracker.go

@@ -1,13 +1,13 @@
 package main
 
 import (
-	"strings"
-	"strconv"
-	"fmt"
-	"log"
-	"json"
 	"container/vector"
+	"fmt"
+	"json"
+	"log"
 	"rand"
+	"strconv"
+	"strings"
 )
 
 type HexTracker struct {
@@ -42,7 +42,7 @@ func NewHexTracker(config *Config) *HexTracker {
 	t.rank = make([]int, t.sqsize+4)
 	t.weights = NewWeightTree(t.sqsize)
 	// initialize union-find data structure
-	for i := 0; i < t.sqsize + 4; i++ {
+	for i := 0; i < t.sqsize+4; i++ {
 		t.parent[i] = i
 		if i < t.sqsize {
 			t.rank[i] = 1
@@ -58,9 +58,9 @@ func NewHexTracker(config *Config) *HexTracker {
 	t.winner = EMPTY
 
 	t.played = make([]byte, t.sqsize)
-	
+
 	t.moves = new(vector.IntVector)
-	
+
 	t.config = config
 
 	return t
@@ -88,10 +88,10 @@ func (t *HexTracker) Copy() Tracker {
 	cp.winner = t.winner
 
 	cp.played = make([]byte, cp.sqsize)
-	
+
 	cp.moves = new(vector.IntVector)
 	*cp.moves = t.moves.Copy()
-	
+
 	cp.config = t.config
 
 	return cp
@@ -132,7 +132,7 @@ func (t *HexTracker) Play(color byte, vertex int) {
 		if t.config.PlayoutProbs && t.config.policy_weights != nil {
 			t.updateNeighborWeights(vertex)
 		}
-		
+
 		if t.played[vertex] == EMPTY {
 			t.played[vertex] = color
 		}
@@ -178,7 +178,7 @@ func (t *HexTracker) suggestion(color byte, last int) int {
 				weights[i] = 1
 			} else {
 				hash := hex_min_hash[hex_hash(color, t.board, t.neighbors[1][n])]
-				hash |= 1<<30
+				hash |= 1 << 30
 				weights[i] = t.config.policy_weights.Get(hash)
 			}
 			weightSum += weights[i]
@@ -209,7 +209,7 @@ func (t *HexTracker) Playout(color byte) {
 			log.Println(Ctoa(color) + t.Vtoa(vertex))
 		}
 		t.Play(color, vertex)
-		
+
 		if t.config.VeryVerbose {
[…]

### 2011-12-27 src/hive_test.go

@@ -1,10 +1,10 @@
 package main
 
 import (
-	"testing"
 	"fmt"
 	"log"
 	"os"
+	"testing"
 )
 
 var config *Config
@@ -52,7 +52,7 @@ func TestGoSwarm(t *testing.T) {
 	config.Go = true
 	config.Hex = false
 	config.MaxPlayouts = 10000
-        config.policy_weights = LoadBest("./swarm.1.gob", config)
+	config.policy_weights = LoadBest("./swarm.1.gob", config)
 	tracker := NewTracker(config)
 	root := NewRoot(BLACK, tracker, config)
 	genmove(root, tracker)
@@ -63,7 +63,7 @@ func TestHexSwarm(t *testing.T) {
 	config.Go = false
 	config.Hex = true
 	config.MaxPlayouts = 10000
-        config.policy_weights = LoadBest("./swarm.1.gob", config)
+	config.policy_weights = LoadBest("./swarm.1.gob", config)
 	tracker := NewTracker(config)
 	fmt.Println(tracker)
 	root := NewRoot(BLACK, tracker, config)
@@ -84,5 +84,5 @@ func TestWeightTree(t *testing.T) {
 			count++
 		}
 	}
-	fmt.Println(tree.Prob(BLACK, target), float64(count) / float64(samples))
+	fmt.Println(tree.Prob(BLACK, target), float64(count)/float64(samples))
 }

### 2011-12-27 src/learning.go

@@ -1,16 +1,16 @@
 package main
 
 import (
-	"math"
+	"container/vector"
+	"fmt"
+	"github.com/ajstarks/svgo"
 	"gob"
+	"log"
+	"math"
 	"os"
 	"rand"
 	"sort"
 	"time"
-	"fmt"
-	"log"
-	"container/vector"
-	"github.com/ajstarks/svgo"
 )
 
 type Swarm struct {
@@ -44,11 +44,11 @@ func NewSwarm(config *Config) *Swarm {
 }
 
 type Particle struct {
-	Strategy       float64
-	Position       map[uint32]float64
-	Min, Max       float64
-	Fitness        float64
-	swarm          *Swarm
+	Strategy float64
+	Position map[uint32]float64
+	Min, Max float64
+	Fitness  float64
+	swarm    *Swarm
 }
 
 func NewParticle(swarm *Swarm, min, max float64) *Particle {
@@ -98,7 +98,7 @@ func (p *Particle) Get(i uint32) float64 {
 }
 
 func (p *Particle) Init(i uint32) {
-	p.Position[i] = p.Min + (p.Max - p.Min) * rand.Float64()
+	p.Position[i] = p.Min + (p.Max-p.Min)*rand.Float64()
 }
 
 func (s *Swarm) evalPlay(p1 *Particle, p2 *Particle) {
@@ -157,7 +157,7 @@ B_o offspring, lambda = |B_o|
 func (s *Swarm) step() {
 
 	parents := s.Particles
-	
+
 	// generate lambda (lambda >= mu for comma) children
 	s.Particles = make(Particles, s.Lambda)
 	for i := range s.Particles {
@@ -170,7 +170,7 @@ func (s *Swarm) step() {
 		s.mutate(s.Particles[i])
 		s.Particles[i].Fitness = 0
 	}
-	
+
 	// propagate the last best particles without change
 	for i := uint(0); i < s.config.Propagate; i++ {
 		s.Particles[i] = parents[i].Copy()
@@ -248,29 +248,29 @@ func (s *Swarm) mutate(p *Particle) {
 }
 
 func (s *Swarm) Best() (best *Particle) {
-  if s.config.Combine {
-	  best = NewParticle(s, s.Particles[0].Min, s.Particles[0].Max)
-	  best.Fitness = 0
-	  superset := make(map[uint32]bool)
-	  for i := range s.Particles {
-		  for j := range s.Particles[i].Position {
[…]

### 2011-12-27 src/main.go

@@ -1,10 +1,10 @@
 package main
 
 import (
-	"os"
 	"flag"
 	"fmt"
 	"log"
+	"os"
 	"rand"
 	"time"
 )
@@ -12,14 +12,14 @@ import (
 func main() {
 	rand.Seed(time.Nanoseconds())
 	config := NewConfig()
-	
+
 	shutdown := make(chan bool, 1)
 	if config.Cluster {
 		InitCluster(shutdown)
 	} else {
 		shutdown <- true
 	}
-	
+
 	if config.Help {
 		flag.Usage()
 		os.Exit(0)

### 2011-12-27 src/search.go

@@ -1,14 +1,14 @@
 package main
 
 import (
-	"fmt"
-	"math"
-	"time"
-	"log"
-	"rand"
 	"container/vector"
-	"os"
+	"fmt"
 	"gob"
+	"log"
+	"math"
+	"os"
+	"rand"
+	"time"
 )
 
 type Node struct {
@@ -349,7 +349,7 @@ func (node *Node) update(t Tracker) {
 
 func (node *Node) recalc() {
 	if node.Visits == 10 {
-		node.value = 1 + 0.1 * rand.Float64()
+		node.value = 1 + 0.1*rand.Float64()
 		return
 	}
 	node.Mean = node.Wins / node.Visits
@@ -400,9 +400,9 @@ func (node *Node) recalc() {
 	r := math.Log1p(node.parent.Visits) / node.Visits
 	v := 1.0
 	if node.config.Var {
-		v = math.Fmin(0.25, node.blendedMean - (node.blendedMean * node.blendedMean) + math.Sqrt(2*r))
+		v = math.Fmin(0.25, node.blendedMean-(node.blendedMean*node.blendedMean)+math.Sqrt(2*r))
 	}
-	node.value = node.blendedMean + node.config.Explore * math.Sqrt(r*v)
+	node.value = node.blendedMean + node.config.Explore*math.Sqrt(r*v)
 }
 
 // return node's grandparent's sibling corresponding to node's move
@@ -626,7 +626,7 @@ func (node *Node) String(depth, maxdepth int, t Tracker) (s string) {
 	if depth < maxdepth && node.Child != nil {
 		for child := node.Child; child != nil; child = child.Sibling {
 			if child.Visits > 0 {
-				s += child.String(depth + 1, maxdepth, t)
+				s += child.String(depth+1, maxdepth, t)
 			}
 		}
 	}
@@ -634,6 +634,6 @@ func (node *Node) String(depth, maxdepth int, t Tracker) (s string) {
 }
 
 type Eval struct {
-	Moves *vector.IntVector
+	Moves        *vector.IntVector
 	Wins, Visits float64
 }

### 2011-12-27 src/tracker.go

@@ -4,14 +4,14 @@ import "container/vector"
 import "rand"
 
 const (
-	UP         = 0
-	DOWN       = 1
-	LEFT       = 2
-	RIGHT      = 3
-	UP_LEFT    = 0
-	UP_RIGHT   = 1
-	DOWN_LEFT  = 4
-	DOWN_RIGHT = 5
+	UP          = 0
+	DOWN        = 1
+	LEFT        = 2
+	RIGHT       = 3
+	UP_LEFT     = 0
+	UP_RIGHT    = 1
+	DOWN_LEFT   = 4
+	DOWN_RIGHT  = 5
 	INIT_WEIGHT = 500
 )
 

### 2011-12-27 src/weight_tree.go

@@ -6,10 +6,10 @@ import (
 )
 
 type WeightTree struct {
-	size int
-	nodes int
-	interior int
-	leaves int
+	size          int
+	nodes         int
+	interior      int
+	leaves        int
 	black_weights []float64
 	white_weights []float64
 }
@@ -31,12 +31,12 @@ func NewWeightTree(size int) *WeightTree {
 
 func (t *WeightTree) Set(color byte, vertex int, weight float64) {
 	if color == BLACK {
-		t.black_weights[t.interior + vertex] = weight
+		t.black_weights[t.interior+vertex] = weight
 		for parent := t.parent(t.interior + vertex); parent >= 0; parent = t.parent(parent) {
 			t.black_weights[parent] = t.black_weights[2*parent+1] + t.black_weights[2*parent+2]
 		}
 	} else {
-		t.white_weights[t.interior + vertex] = weight
+		t.white_weights[t.interior+vertex] = weight
 		for parent := t.parent(t.interior + vertex); parent >= 0; parent = t.parent(parent) {
 			t.white_weights[parent] = t.white_weights[2*parent+1] + t.white_weights[2*parent+2]
 		}
@@ -45,9 +45,9 @@ func (t *WeightTree) Set(color byte, vertex int, weight float64) {
 
 func (t *WeightTree) Get(color byte, vertex int) float64 {
 	if color == BLACK {
-		return t.black_weights[t.interior + vertex]
+		return t.black_weights[t.interior+vertex]
 	}
-	return t.white_weights[t.interior + vertex]
+	return t.white_weights[t.interior+vertex]
 }
 
 func (t *WeightTree) Prob(color byte, vertex int) float64 {
@@ -95,15 +95,15 @@ func (t *WeightTree) rand(color byte, node int, root int) int {
 	if node >= t.interior {
 		return node - t.interior
 	}
-	left := 2 * node + 1
-	right := 2 * node + 2
+	left := 2*node + 1
+	right := 2*node + 2
 	var weights []float64
 	if color == BLACK {
 		weights = t.black_weights
 	} else {
 		weights = t.white_weights
 	}
-	if rand.Float64() * weights[root + node] < weights[root + left] {
+	if rand.Float64()*weights[root+node] < weights[root+left] {
 		return t.rand(color, left, root)
 	}
 	return t.rand(color, right, root)

### 2012-03-18 thesis/paper/paper.tex
Paper submitted for publication to Foundations of Digital Games 2012.

\documentclass{acm_proc_article-sp}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\newcommand{\black}{\texttt{Black}}
\newcommand{\white}{\texttt{White}}
\newcommand{\blacks}{\texttt{Black's}}
\newcommand{\whites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}

\title{Learning to Bias MCTS Playouts}
\numberofauthors{2}
\author{ 
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
James Pettit \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{ james.l.pettit@gmail.com }
% 2nd. author
\alignauthor
David Helmbold \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{dph@soe.ucsc.edu}
}

\begin{document}

\maketitle

\begin{abstract}
Monte-Carlo Tree Search (MCTS) 
grows a partial game tree and uses a large number of random simulations to
approximate the values of the nodes.   
It has proven effective in games with such as Go and Hex where the large search space and
difficulty of evaluating positions cause difficulties for standard methods.
The best MCTS players use carefully hand-crafted rules to bias the random simulations.
Obtaining good  hand-crafting rules is a very difficult process, as even rules promoting better
simulation play can result in a weaker MCTS system~\cite{gelly2006modification}.
Our Hivemind system uses evolutionary strategies to automatically learn effective rules for biasing the 
random simulations.
We have built a MCTS player using Hivemind for the game Hex. 
The Hivemind learned rules result in a 90\% win rate against a baseline MCTS system, and 
significant improvement against the computer Hex world champion, MoHex.
\end{abstract}

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. 
Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can be effective.

In contrast, cutting and connecting games like Go, Hex, and Havannah, have much larger search trees and long-range interactions make it difficult to create accurate evaluation functions.
In these domains, Monte-Carlo Tree Search (MCTS) has proven effective~\cite{}, 
with the world champion Go and Hex players both using MCTS~\cite{}.
MCTS picks its next move by growing a partial game tree while using
a large number of randomized simulations to evaluate the possible moves.
Each simulation first uses a child-selection policy to navigate through the partial game tree.
After reaching a leaf of the partial game tree, a randomized continuation policy is used to complete the game 
and determine a winner.  
Finally, an update policy is used to update the partial game tree with the result of the simulation.
Although all three policies have an effect on the algorithm's playing strength, here we focus on
continuation policies for the game Hex.
Section~\ref{mcts} contains a fuller description of MCTS.

MoGo specifically, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better continuation policies.

The Hivemind approach automatically learns better continuation policies from self-play. 
The usual process of improvement requires hand-tuned weights that bias the continuation policy. 
Any change must be tested to detect the change in strength it gives,
and different weight parameters can have synergistic effects.
Small changes in overall algorithm strength require thousands of games of testing to detect. 
Often the improvement made by a small change in weight can be so small it will go unnoticed. 
Automatic weight tuning can improve this process, but still requires thousands of games to test. 
Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. 
Hivemind is an alternative approach using evolutionary learning. 
Because it relies on self-play, Hivemind avoids the need for extensive expert knowledge or a strong outside player.
Although our experiments have focused on Hex, 
the Hivemind framework is suitable for generic Monte-Carlo Tree Search, and has been implemented for
both Go and Hex.

The continuation biases learned by Hivemind dramatically improve the playing strength of simple MCTS players.
Our experiments show that the Hivemind biases have a 90\% win rate over an unbiased continuation policy,
and about 85\% win rates over two other continuation policies that favor local moves.  
Hivemind also wins 6.5\% of its games against the world champion MoHex.
Although 6.5\% indicates that there is still a lot of room for improvement, 
we note that the other simple policies were unable to win \emph{any} games against MoHex. 
Furthermore, Hivemind is improving only the continuation policy, 
improvements in the child-selection and tree updating policies could have greatly synergistic benefits.

The evolutionary strategies approach used by Hivemind is surprisingly stable.
In 10 different evolutionary runs, the win rates over the unbiased continuation policy remained
between 87\% and 93.5\%.  
Furthermore, the learned strategies generalize well to different board sizes.
Despite using $7 \times 7$ boards for faster training, 
the learned biases performed as well (and usually better!) against the simple policies on $11 \times 11$ and $13 \times 13$ boards.

The remainder of the paper begins with some background on the game of Hex in Section~\ref{s:hex} and
an overview of MCTS in Section~\ref{mcts}.
The Hivemind system is presented in Section~\ref{hivemind}
and Section~\ref{results} describes our experimental setup and results.

\section{Background: The Game of Hex}
\label{s:hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by  a color, either \black{} and \white. 
The players alternate placing markers of their color on empty hexagons. 
The object of the game is to form an unbroken chain from one side of the board to the other. 
Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board, note that the numbers indicate
the order in which the markers were played and have no effect on the game.
Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move (71), completing an unbroken chain connecting both his sides.

Hex is a cutting and connecting game that
has properties that make it attractive for AI research. 
Once placed, a piece never leaves the board. 
The only criterion for a legal move is that the hexagon be unoccupied (empty). 
Gale '79 proved the game cannot end in a draw, simplifying minimax methods. 
Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the winning strategy.
Solutions for all opening positions for 7x7 and 8x8 boards have been found~\cite{henderson2009solving}. 
A good portion, but not all of the 9x9 openings have also been solved. 
Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.

There are three exceptionally strong Hex programs.  Gabor Melis' \emph{Six} is the previous world champion (up to 2007). 
Six uses traditional alpha-beta search with an evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)~\cite{anshelevich2002hierarchical}. 
\emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves. 
\emph{MoHex}, also from Alberta, is based on the newer Monte-Carlo Tree Search technique, leveraging the open source \emph{Fuego} library. 
MoHex is the current world-champion, having finished ahead of Wolve in 2008, 2009 and 2010.

Go is a much older game than Hex or Havannah, and much research has been done on Go playing algorithms. 
Despite its simpler rules, Hex has a similar branching factor and position evaluation complexity as Go. 
Although we emphasize Hex here, the techniques discussed generalize to any large branching game played on a regular board
(such as Go or Havannah).

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
	\end{HexBoard}
	}
	\caption{13x13 Hex Game, \white{} to play.}
	\label{fig:13x13inprogress}
\end{figure}

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
				   e8,e9,c10,k8,c9,g6,g7,c11,
				   d10,h6,h7,i5,j5,j4,k4,k3,
				   m2,l3,m3,l4,c6,e11,l5,m4,
				   d11,b7,e6,d12,c12,c7,e7,f5,
				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
				   c13,c5,b6}
	\end{HexBoard}
	}
	\caption{Finished 13x13 Hex Game, \black{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
	\label{fig:13x13finished}
\end{figure}

\section{Monte Carlo Tree Search} \label{mcts}

\begin{figure}
	\begin{center}
%	\centerline{\begin{HexBoard}[board size=5]
%		\HGame{a3,c2}
%	\end{HexBoard}
%	}
	\includegraphics[width=3.0in]{graphics/tree.png}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

%As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS)~\cite{mohex}.

The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann~\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a partial  search tree. 
The UCT algorithm was developed alongside MCTS to direct the growth of the search tree~\cite{gelly2006exploration}. 
At any point, the search can be stopped and the current state of the tree used to select a move to play.

All nodes in the tree correspond to a fixed board position, i.e. the occupancy state of the board and which player is next to move. 
Nodes also have statistics associated with their position:  the number of winning simulations performed that used the node, and the total number of simulations that included the node. 
For each simulation the algorithm will start at the root node, recursively traverse the (partial) tree until it reaches a leaf node.
The simulation continues from the position represented by this leaf node until a winner is determined.
The result of the simulation used to update the statistics at nodes in the tree.
If the leaf node has been visited enough times (50 in our case), then it is expanded and nodes for all possible moves from 
that position are added to the search tree.

The effectiveness of MCTS depends on three policies:
\begin{itemize}
\item{The \emph{child-selection} policy recursively traverses the tree, selecting promising moves for further exploration.}
\item{The \emph{continuation} policy randomly or semi-randomly ``plays out'' the game from a leaf position in the tree until a final position.}
\item{The \emph{update} policy modifies the win/loss statistics of nodes based on the simulation results, and determines when to grow the search tree.}
\end{itemize}
Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position. In the example, the child-selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}. 
The continuation policy ran, completing the game from the positiion with \texttt{E4} occupied by \black\ and \texttt{A1} occupied by \white.
This resulted in a loss for \black. 
The update policy choose to update all nodes along the visited path, incrementing the wins for \white\ moves and incrementing the losses for \black\ moves.

\begin{lstlisting}[float,frame=single,language=Python,caption=MCTS Algorithm Pseudocode]
def search(root):
  for i in range(simulations):
    step(root)

def play_from(node) returns outcome:
  if node.visits < threshold:
    outcome = simulate(node)
  else:
    child = select(node)
    outcome = play_from(child)
  if outcome is win for node.player:
    node.wins++
  node.visits++
  return outcome
\end{lstlisting}

\subsection{Child Selection Policy}
The \texttt{select} function defines the child selection policy. It must return the child of the given node to step into next. 
In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
It is important to ``exploit" those child nodes with high win-rates, both
to discover possible weaknesses that would result in the opponent winning and to obtain accurate win-rate estimates. 
Simultaneously, a child with a low number of visits must also be explored since a low win-rate
could be due to a few unlucky simulations.
This exploration/exploitation tradeoff is related to a variance reduction problem.  
We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move
could actually be very good.

One way to balance this exploration/exploitation trade-off is the Upper-Confidence for Trees algorithm (UCT). 
UCT incorporates the win-rate of a child with a term that increases in value as the ratio of the child's visits to its parents decreases. 
Eventually, no matter how bad the estimated value of a position is, it will be re-visited and the estimate updated. 
UCT has several variations, but the ``pure'' version has been proven to eventually yield the correct minimax value of a position, given an infinite number of simulations~\cite{gelly2006exploration}.

The UCT priority of a child node is:
\[
	\texttt{child.winrate} + C*\sqrt{\frac{\log{(\texttt{parent.visits})}}{\texttt{child.visits}}}.
\]

The \emph{exploration coefficient}, $C$, can be varied to increase or decrease the level of exploration versus exploitation. Note that a very low value will yield a function that relies almost exclusively on the estimated win-rate of a position, while a high value will visit nodes more equally, placing less emphasis on the estimated value.

\subsection{Continuation Policy}

The \texttt{continuation} function plays out the rest of the game from the current position. 
The simplest policy, which we call the \emph{default policy},
 plays uniformly at random from the legal moves until the game ends (in Hex, when one player has a winning path). 
The continuation policy has a large effect on the strength of the MCTS player.
% and the Go literature has a large body of work dedicated to improving it.
Initial research in MCTS Go players focused on using a strong continuation policy, with expert-derived heuristics to select the next move~\cite{chaslot2010adding}. 
The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates. 
Only when none of the rules match do they fall back to a uniform random move. 
It was quickly found that building a stronger continuation policy does not guarantee better performance when used in the full tree search~\cite{gelly2006modification}. 
In other words, when the continuation policies are used as players, the relative strengths of the continuation policies are poor predictors
of how well the MCTS algorithms relying on these continuation policies will perform.
This surprising result may be due to stronger continuation policies under-exploring the search space.
In any case, this means that hand crafting an effective continuation policy requires benchmarking and frequent tests of the entire
MCTS system as small changes to the simulation policy can have unpredictable effects on overall performance.
This illustrates the need for automated systems for tuning the continuation policy.

Another constraint on the continuation policy is its speed.  
MCTS algorithms require large numbers of random simulations to compute accurate statistics and select good moves, therefore the continuation policy must be not be overly complex and implemented efficiently.

\subsection{Update Policy}
Once a simulation is finished, the result is used to update the search tree.
Basic MCTS systems updated the win and visit statistics only along the traversed path.
It was quickly discovered that more aggressive update strategies are often profitable.
One simple example of this is the All-Moves-As-First (AMAF) heuristic~\cite{}
that exploits the fact that moves played in a simulation could have been played in any order. 
Thus, a move deep in the simulation could be fictitiously treated as the ``first'' move from an intermediate position. 
In addition to updating the statistics of the nodes in the path, all children of nodes on the path corresponding to moves made later
in the simulation have their AMAF win and visit counts updated. 
Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.

Several methods can be used to blend the AMAF results with the ``pure'' results, such as a weighted average.
Other blending functions include Cutoff and Rapid Action Value Estimation (RAVE)~\cite{chaslot2008progressive}. 
Helmbold and Parker-Wood (2009) have a summary of the various update policies and blending methods, along with experimental results showing marked improvement over the baseline of pure UCT~\cite{helmbold2009all}. 
One noteworthy conjecture is that there is no ``silver bullet'' update policy. If the simulation policy is random and unbiased, the AMAF updates might be similarly unbiased. However, a simulation policy that preferentially plays certain moves might have an unpredictable effect on which nodes receive AMAF updates. This can then propagate to which nodes are selected during the UCT selection, and have chaotic effects on the search as a whole.

Our Hivemind experiments use only the basic update policy, updating win and visit counts for only those nodes on the traversed path.

% TODO
\textbf{does MoHex do something fancier?}
% MoHex uses one local pattern - bridge preservation, and AMAF updates with RAVE (http://webdocs.cs.ualberta.ca/~hayward/hex/#mohex)

The tree update policy also determines how the partial search tree is grown.
Generally, systems will add the possible children of a leaf to the tree after the leaf has been visited a sufficient number of times.
This number is a parameter in Hivemind, and our experiments have set it to the (untuned) value of 50.

\subsection{Biasing the Continuation -- Incorporate into earlier subsections}\label{bias}

Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial~\cite{gelly2006modification,gelly2008achieving}. However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games). They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like nakade (life and death). If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situations. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive number of simulations.
In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in its child selection policy~\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.
%
%\section{Problem Statement  -- delete? Incorporate points into MCTS section?}\label{problem}
%MoGo specifically, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.
%
%It would be ideal to automatically learn a better playout policy from self-play. The usual process of improvement requires hand-tuned weights that modify the playout policy. Any change must be tested to detect the change in strength it gives. Problematically, small changes in overall algorithm strength require thousands of games of testing to detect. Many times the improvement made by a small change in weight can be so small it will go unnoticed. Automatic weight tuning can improve this process, but still requires thousands of games to test. Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
%

\section{The Hivemind System}\label{hivemind}

Hivemind is a system that learns to play abstract strategy games on a regular board. 
The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``Game Tracker interface'' used by the tree search to maintain game state information. 
The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go~\cite{golang}). 
In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games~\cite{Fuego}. 
Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. 
The major contribution of Hivemind is the integration of evolutionary learning to automatically learn appropriate continuation policies.
Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
Preliminary work tried learning with particle swarm optimization, but we found that evolutionary strategies gave
superior results.

Using evolutionary strategies to learn continuation policies requires that the continuation policies are encoded as vectors.
We first describe this encoding.  Section~\ref{s:tracker} describes the Game Tracker module, and Section~\ref{s:es} describes the
use of Evolutionary Strategies.

\subsection{Continuation Policies and Local Patterns}
\label{s:policies}

Hivemind uses local patterns around the previous move to bias the random continuations.
Responding to an opponent's threat is often necessary in cutting and connecting games
and local patterns around the last move can be efficiently evaluated, allowing for efficient simulation.
Strong MCTS Go programs like MoGo~\cite{gelly2006modification} use hand-crafted local patterns to direct their  continuation policies.

Each empty location around this last move is a candidate local response, and a pattern database gives each of these candidates a
weight indicating its relative importance.
During a continuation play-out, a candidate local response is selected with probability proportional to its weight.
Figures \ref{fig:localpattern} and \ref{fig:encoding} show graphically how a local patterns can be mapped to an integer. 
The orange marker denotes the last move played. 
In Hex there are less than $2 * 4^6=2^{13}$ possible local patterns, and 
assigning a weight to each of these patterns creates a continuation strategy.
Thus the continuation strategies we consider correspond to an array of $2^{13}$ weights, one for each local pattern.
% TODO
\textbf{How about tenuki weights? does hivemind consider them?}
% Early experiments with tenuki weights were less successful. The default policy is only used if no legal moves exist in the local area.

In the example shown in figure \ref{fig:localpattern}, the sum of the three weights is 11. Thus, vertex A will be selected with probability $P(A) = \frac{6}{11} \approx 0.54$. 
B will be selected with $P(B) \approx 0.27$ and C with $P(C) \approx 0.18$. 
Note that $P(A)+P(B)+P(C) = 1$ so one of A, B, or C will be selected.

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/local-pattern.png}
	\caption{A 4x4 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/weight-pattern-map.png}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
	\end{center}
\end{figure}

A local pattern can assign a candidate move a weight of 0. If so, it will never be selected using the local response policy. 
If all patterns adjacent to the last move have weight 0, or if all adjacent vertices are already occupied, 
the continuation policy makes a move at random from all legal possibilities.

Although conceptually an array with $2^{13}$ entries, Hivemind implements continuation strategies as
32-bit hashmaps.  
The additional overhead is modest and the larger hashmaps  have space for additional properties (such as distance to a side).
Note that in Figure~\ref{fig:encoding} the encoding includes a color bit in the 31st position. This denotes the color to play (either \black\ (1) or \white\ (0)). 
This bit is important because \black\ and \white\ have different goals -- they are trying to connect along different diagonals. 
By using a reflection and flipping colors, one position can be translated into a symmetric  position for the opposite color. 
We experimented with this approach, but it proved more troublesome than beneficial.

There are three simple continuation policies that we use as baseline comparators for our learned policies.

\begin{description}
	\item[Default] - The next move is chosen uniformly at random from the set of all legal moves.
	This policy corresponds to setting all of the weights to 0 in our encoding .
	\item[Uniform local] - The next move is chosen uniformly at random from the empty locations adjacent to the last move.
	If no legal move exists on these 6 points, then the move is chosen uniformly from the set of all legal moves 
	(i.e. using the default policy).  This policy corresponds to setting all the weights to 1.
	\item[Uniform local with ``tenuki''] - The tenuki (Japanese for  ``play away'') variant is a mixture of the Default and Uniform local policies.  
	With probability 5/6 the move is chosen with the Uniform Local policy and with probability 1/6 it is chosen with the Default policy. 
\end{description}

\subsection{Hivemind Game Tracker module}
\label{s:tracker}

The Game Tracker module allows real games to be played via Monte-Carlo Tree Search.
Its design was influenced by Fuego~\cite{Fuego},  and the Game Tracker interface is very similar to Fuego's subclassing system for implementing multiple games.
Note that playing a single \emph{real game} involves playing millions of  \emph{simulated} games when
moves in the real game are chosen using Monte-Carlo Tree Search.
Part of the Game Tracker's responsibility is to maintain the 
current simulation.
The Game Tracker also stores the continuation policies of the two players as well as both
the real-game board state and the board state in the current simulation.
Note that the player making the move in the ``real game" uses their continuation policy for both players in the simulated games.

The MCTS  module is a simple 
generic implementation of Monte-Carlo Tree Search, using the UCT algorithm for child selection
(see Section \ref{mcts}).
The MCTS module uses the Game Tracker to keep track of the simulated game state 
as it runs simulated games and expands the search tree.
The MCTS module is only about 600 lines of source code. 
Using some extra information from the Game Tracker,  
the MCTS module can be configurable to use RAVE and AMAF heuristics to 
update multiple nodes using the results of a single simulation.

The Game Tracker uses the policy weights for the players to implement the appropriate continuation policy and complete simulated games. 
It also implements the three baseline continuation policies described in Section~\ref{s:policies}.
Completing simulated games quickly is critical, as more simulations-per-second generally results in better move selection. 
This emphasis on speed is why the tracker uses a monolithic \texttt{Playout} function
to implement continuation policies.
Many games have optimizations that can be performed in this \texttt{Playout} function that the search tree is unaware of. 
For example, the Hex tracker initializes a randomized list of empty positions so default policy move selection is a O(1) operation. 
This exploits the fact that in Hex, any empty location is a legal play. 
This is not the case for Go, so the Go engine uses other optimizations for fast random move selection.

The Game Tracker module uses an interface that is easily extendible to other games. 
In addition to the \texttt{Playout} function, the Game Tracker provides the following functionality.
\begin{enumerate}
\item A single function, \texttt{Play(color, vertex)}, is used by the search tree to make a move in the simulated game.
\item The \texttt{Legal(color, vertex)} function is used to query the legality of a given move in the simulated game. 
The MCTS module uses this when the tree is gown as each time a leaf node is expanded,
one child is added for each legal move from the leaf node's state.
\item The \texttt{Winner()} function is used to determine the current winner of the game. 
	A special \texttt{Empty} player is returned if the game has not finished.
\item The \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
\end{enumerate}
A few extra functions also provide for debugging by pretty-printing the tracker's internal state.

The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. It also constrains the space of games that can be easily implemented in Hivemind. 
In particular, Hivemind allows for 2 player games with perfect information. 
The game space is not required to be a regular board. 
Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children. Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.

\subsection{Evolution Strategies}\label{s:es}

Hivemind implements evolutionary strategies to learn good continuation policies.
Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors. 
More information on evolution strategies and evolutionary learning in general can be found in Eberhard and Shi's excellent textbook \emph{Computational Intelligence: Concepts to Implementations}~\cite{eberhart2007computational}. 
The Evolution Strategies article on Scholarpedia.org also provides a good introduction to the topic~\cite{Beyer:2007}.

Our implementation of EW works as follows.
\begin{enumerate}
\item An initial population (we use 30 individuals) is randomly generated. 
Each individual is represented by its vector of $2^{13}$ local pattern weights 
(each pattern weight is randomly chosen from $[0 .. 100]$)
and a strategy parameter $\sigma$ that controls how fast it evolves.
During assessment, the individual will also be assigned a real-valued ``fitness'' attribute.
\item The algorithm iterates the following with each iteration forming one ``generation'':
	\begin{enumerate}
	\item From the 30 parents, generate 35 children.
		\begin{enumerate}
		\item Each child is created by randomly selecting 2 parents and then 
		averaging the parents' local patten weight vectors and strategy parameters.
		\item	The child's strategy parameter $\sigma$ is \emph{mutated} to $\sigma e^{ {\cal N}(0,\tau^2)}$
			(see below).
		\item Each local pattern weight of the child is mutated by adding a random variable drawn from the Normal Distribution
		${\cal N}(0, \sigma^2)$.
		\end{enumerate}
	\item The best 5 performing parents are propagated through without mutation.  This is done to preserve well-performing solutions.
	\item Evaluate the fitness of the new pool of 40 individuals, and keep only the top 30 as parents for the next generation.
	\end{enumerate}
\end{enumerate}

The numbers of:  individuals starting generations, children generated, and good parents kept without mutation
were picked arbitrarily, and do not have deep significance.

\subsubsection*{Mutation of $\sigma$}
Note that $\sigma$ is multiplied by a positive factor at each time step, and can never reach zero. 
The ``learning rate" $\tau$  can be scaled up or down to increase or decrease the variance of $\sigma$, a high values will give more mutation, low values less. 
We use an initial $\tau = \frac{1}{\sqrt{2^{13}}}$ (recall that there are $2^{13}$  pattern weights) and decrease $\tau$ to zero 
as the number of generations approaches a set maximum (100 in our case). 
This forces the algorithm to converge, though not necessarily to an optimal solution. 
This choice of $\tau$ is driven by experimental literature suggesting it works in many cases~\cite{Beyer:2007}.

%\begin{figure}
%\begin{flalign*}
%	\tau_{t} = & \frac{1}{\sqrt{2D}} \left(1 - \frac{\textrm{generation}}{\textrm{max generations}}\right) \\
%	\sigma_{t} = & \sigma_{t-1} e^{N(0, \tau^2)} \\
%	p_{i, t} = & p_{i, t-1} N(0, \sigma_{t}^2)
%\end{flalign*}
%\caption{Differential equations for $\tau$, $\sigma$, and $\bf{p}$, indexed by time $t$. $D$ is the number of dimensions, i.e. $|\bf{p}|$}
%\label{mutate}
%\end{figure}
%

\subsubsection*{Fitness Evaluation}
%\label{fitness}

Every generation, each individual plays $n$ complete games, each game played against an opponent chosen  randomly from the pool.
Opponents are drawn with replacement, so a given individual could play the same opponent more than once each generation. 
At the beginning of each game, one of the two players is randomly chosen to play \black, the other to play \white. 
A swap-safe move was used to play the first move to greatly reduce the first-mover advantage. 
The winner of each game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.
% TODO
\textbf{This means that individuals expect to play $2n$ games total, but the exact number will vary?}
% Correct

%\subsection{The Hivemind System}
%Hivemind was designed for experiments in learning to play abstract strategy games on a regular board. The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``Game Tracker interface'' used by the tree search to maintain game state information. The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go~\cite{golang}). In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games~\cite{Fuego}. Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. The major contribution of Hivemind is the integration of evolutionary learning into the search process. Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
%
%\subsubsection{Learning}\label{learning}
%The learning module is responsible for driving the evolutionary search process. In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that performs well during the evaluation step. Hivemind currently uses Evolution Strategies, although previous versions experimented with Particle Swarm Optimization. Selection and mutation are implemented as described in section \ref{estrats}. Evaluation is described in section~\ref{fitness}. 
%
%\subsubsection{Search}
%The search module is a simple generic implementation of Monte-Carlo Tree Search, using the UCT algorithm to traverse the tree. Section \ref{mcts} has a description of the MCTS process. The search module uses the Game Tracker to expand the tree and perform the random playouts. The entire source code for search is only about 600 lines of code. Using some extra information from the Game Tracker, search is configurable to use RAVE and AMAF heuristics to update multiple nodes using the results of a single playout.
%
%\subsubsection{Game Tracker}
%The Game Tracker allows complete games to be played vi Monte-Carlo Tree Search. The implementation of a game has four responsibilities:
%\begin{enumerate}
%\item{Maintain all required game state. A single function, \texttt{Play(color, vertex)}, is used by the search tree to apply a move for the given color.}
%\item{Report legal moves from a given game state. The \texttt{Legal(color, vertex)} function is used to query the legality of a given move. The search tree uses this to expand a parent node, adding all legal moves as children.}
%\item{Report on the status of the winner. The \texttt{Winner()} function is used to determine the current winner of the game. The special \texttt{Empty} player is returned while the game is in progress, and either \black{} or \white{} once a winner has been determined.}
%\item{Playing out the game from the current game state. The tracker is parameterized with the policy weights, and responsible for using them during the playout. The speed of the playout function is also critical, as more playouts-per-second generally result in better move selection. This emphasis on speed is why the tracker is required to implement a monolithic \texttt{Playout} function. Many games have optimizations that can be performed in the playout that the search tree is unaware of, for example, the Hex tracker keeps a simple list of randomized empty positions, so random move selection is a O(1) operation. This is because in Hex, any empty position is legal. This is not the case for Go, so the Go engine uses other optimizations for fast random move selection.}
%\end{enumerate}
%The additional \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic. A few extra functions also provide for debugging by pretty-printing the tracker's internal state.
%
%The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. It also constrains the space of games that can be easily implemented in Hivemind. In particular, Hivemind allows for 2 player games with perfect information. The game space is not required to be a regular board. Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children. Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.

\section{Results and Conclusions}\label{results}

We experimented with Hivemind and evolution strategies to answer the following four questions:
\begin{enumerate}
\item Do learned continuation policies have an advantage over the three baseline MCTS polices?
\item Are the results of evolution strategies learning stable and predictable?
\item How effective are the learned policies against world-class competitors?
\item Do the learned continuation policies generalize well to other board sizes?
\end{enumerate}

To train the learned policy we used the evolution process as described in Section~\ref{s:es} was run for 100 generations.  
Each evaluation game was played on a 7x7 board and used 1000 simulations per move in the fitness evaluation games. 
The individual from the 100th generation having the highest fitness was selected as the \emph{learned} continuation policy.
The learned continuation policy, as well as the three basic continuation policies described in Section~\ref{s:policies}, were coupled with
a slightly modified UCT child selection policy that included the variance term suggested by Gelly 2006~\cite{gelly2006exploration} and a
pure update policy, where only those nodes directly in the path traversed by UCT are updated with the result of a simulation.
Matches were run using the Gomill software~\cite{gomill}. 
Gomill required no modification to work with Hex playing programs.

\subsubsection*{Benefits of learned policy}
To determine the benefits of the learned continuation policy we played a round-robin set of matches between the learned, 
default, uniform local, and uniform with tenuki policies.
Each match consisted of 200 games on a 7x7 board where the first player alternated was forced to start with a
swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.}
move.
Each player was allowed 10,000 simulations per move and the results are shown in Figure~\ref{fig:results}.

\begin{figure}
	\begin{center}
		\begin{tabular}{c | c c c}
		& \multispan{3}{\hfil opponent \hfil} \\
		 player & default & uniform & uniform (tenuki) \\
		\hline
		uniform & 70.50\% & & \\
		uniform (tenuki) & 61.00\% & 50.00\% & \\
		learned & 90.00\% & 84.00\% & 86.00\% \\
		\end{tabular}
	\caption{All-play-all tournament of the 4 Hivemind variants. Each element is the percent win-rate of the row variant versus the column variant.}
	\label{fig:results}
	\end{center}
\end{figure}

The learned policy is the decisive winner, winning almost 90\% of its games.
This shows the clear benefits of automatically learned weights over the default continuation policies.
We originally thought that the addition of a tenuki probability would add beneficial diversity to the continuation searches, but
it has little or no improvement over the uniform local policy.
As expected, uniform local play significantly improves upon the default (random selection over all legal moves)  continuation policy.

%
%The fact that evolutionary learning can find weights that boost performance versus uniform local in the same manner that uniform 
% local improves upon pure random shows how much further there is to go, and shows the weaknesses of purely random simulations.

\subsubsection*{Stability of Evolution}

%\begin{figure}[t]
%	\begin{center}
%		\begin{tabular}{c c}
%		& Default \\
%		\hline
%		Learned 1 & 91.0\% \\
%		Learned 2 & 90.5\% \\
%		Learned 3 & 90.5\% \\
%		Learned 4 & 87.5\% \\
%		Learned 5 & 90.5\% \\
%		Learned 6 & 93.5\% \\
%		Learned 7 & 93.0\% \\
%		Learned 8 & 91.5\% \\
%		Learned 9 & 87.0\% \\
%		Learned 10 & 92.5\% \\
%		Min & 87.0\% \\
%		Max & 93.5\% \\
%		Mean & 90.75\% \\
%		Std. Dev. & ~2.14\% \\
%		\end{tabular}
%		\caption{Results of 10 independent evolutionary processes versus the default policy. Percentage shown is average win-rate of row player versus column player.}
%		\label{fig:a1results}
%	\end{center}
%\end{figure}

A natural concern is that the evolutionary strategies could have large variations in their end results and our first learned policy was
just a ``lucky'' happenstance. 
To protect against this, we re-ran the learning process described above another 9 times, creating a total of 10 different learned continuation policies.
We then played 200 game matches between each of the resulting policies and the default policy using the above protocol.
The following table gives the resulting win-rates for the learned policies.

\medskip

\centerline{
\begin{tabular}{c c c c  c c c c  c c c}
\multispan{10}{\hfil Winning \% for Learned Policy vs. Default \hfil} \\
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\
91 & 90.5  & 90.5  & 87.5  & 90.5  & 93.5  & 93 & 91.5  & 87 & 92.5 
\end{tabular}
}

The 10 learned policies were tightly clustered with a mean of 90.75\% and a standard deviation of only 2.14\%, with a mean of 90.75\%.
This surprising result indicates that the evolutionary processes is far more stable than we expected.

\subsubsection*{Comparison against MoHex}

A potential problem with self-learning is that one might only be learning how to exploit the weaknesses of their siblings, and
not making a general improvement in playing strength.  
To examine this we used an outside control:  the MoHex program~\cite{mohex} from the University of Alberta, compiled and run with its default settings~\cite{mohex}. 
The optimization and domain expertise in MoHex make it a very strong  competitor --  MoHex has won the computer hex world championships in 2009 and 2010.

In the 7x7 matches against MoHex, 
MoHex played with the default time settings of 10 seconds per move while the Hivemind programs again used
10,000 simulations per move.
\textbf{MoHex uses some of its time to perform game theoretic search and pruning before the Monte-Carlo phase}. 
The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. 
\textbf{Although the time controls are not directly comparable, we did track CPU time and use that for a comparison of 
the computational resources used by the various systems.  \emph{for 7x7 or larger size only?}}
Since we were unable to force MoHex to play a particular first move, all games had the Hivemind policy moving first with a swap-safe move.  The Hivemind win percentages (400 game matches) against MoHex are given below.
% TODO
% Swap-safe is C4 on 7x7, C8 on 11x11

\begin{centering}
\begin{tabular}{ccc}
Learned policy & Uniform local & Default policy \\
42\% & 26\% & 11.75\%
\end{tabular}
\end{centering}

As expected, the learned policy does significantly better against MoHex than the other policies,
with the Learned policy winning above 40\% of its games (the results for the other 9 learned policies are similar,
with win rates from 39\% to 47\%).   
The good performance against MoHex is a little surprising, and could be due to the relatively small board size and/or 
the swap-safe move still favors the first player (after any first move, one of the players will have a forced win).
Despite this, the superiority of the learned policy over the baseline ones is dramatically evident.

\subsubsection*{Generalization across board sizes}

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{c c c c}
		& Learned \\
		Default 11x11 & 96.8\% \\
		Default 13x13 & 100.0\% \\
		Uniform 11x11 & 82.76\% \\
		Uniform 13x13 & 92.86\% \\
		\hline
		\end{tabular}
		\caption{Results of evolutionary process learned on a 7x7 board, playing on an 11x11 and 13x13 board. Percentage shown is average win-rate of column player versus row player.}
		\label{fig:a2results}
	\end{center}
\end{figure}

The computational time required of learning process greatly depends on the boardsize used in learning. For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage. The best individual from the 100th generation of this training was used in matches against the default and uniform policy, but a boardsize of 11x11 and 13x13 was used for the matches. Because of the increased boardsize, players used 100,000 simulated playouts per move. The computational time required for these games is significantly higher than that required for the 7x7 games, so the results shown in Figure \ref{fig:a2results} are only after 30 games per match.

Even with the low number of games, the results shown in Figure \ref{fig:a2results} are positive. Learning on a 7x7 board provides results that can generalize to larger boardsizes. This is especially useful because of the exponential increase in computational time required to play games for evolutionary learning when learning on larger board sizes.

Figure \ref{fig:relativecpu} shows the results of these matches.

\begin{figure}
	\begin{center}
		\begin{tabular}{c  | c c c}
		& &	Hivemind  & Hivemind \\
		Hivemind & Opponent & Win-rate & Relative CPU \\
		\hline
		Default & MoHex & 0\% & 555\% \\
		Learned & MoHex & 6.5\% & 664\% \\
		Learned & Hivemind Default & 92.5\% & 131\% \\
		\end{tabular}
		\caption{Win-rate and relative CPU used of Hivemind versus MoHex on an 11x11 board. Relative CPU is the ratio of average CPU time per move for Hivemind vs. opponent.
		% TODO
		% These were on an 11x11 board, but with alternating colors for the opening position. I.e. Hivemind and Default used the swap-safe move when they played black, while MoHex was allowed any move when it was playing black. If only the games where Hivemind or Default played the swap-safe move are allowed, there are 100 samples, Hivemind won 11 games, and Default still won 0.
		}
		\label{fig:relativecpu}
	\end{center}
\end{figure}

Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 games played. 
MoHex strongly outperformed Hivemind, but used only a fraction (about 1/6th) of the CPU time, showing a large distance to cover to get Hivemind up to competition strength. 
In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results. This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 90\% with 10,000 moves. The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.

\subsubsection* {Conclusions}
With infinite time, an unbiased playout policy can find the true minimax value of a position in the game of Hex. In practice, both resources and time are finite. Bias in the playout policy can be extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search. Evolutionary learning can provide a substantial performance boost over naive patterns. With self-play and evolution, weights can be learned automatically, without requiring expert knowledge or test-and-check tuning. The learning process can be done offline, and the final result can be used with little overhead.

It is worth emphasizing two points.
First the learned continuation policies are coupled with relatively naive update and child-selection
polices.
Second, that the learned policies are produced automatically, without domain expertise beyond the basic rules.
Yet even so, they perform well against strong programs (at least on small board sizes).

Source code for the Hivemind Computer Hex/Go player is available at \url{http://github.com/etherealmachine/hivemind}. The included README file has instructions on running the evolutionary learning process and using the results.

\bibliography{paper}
\bibliographystyle{plain}
\end{document}

### 2012-03-18 thesis/paper/paper.tex
Working through revisions suggested by reviewers.

@@
 \usepackage{havannah}
-\newcommand{\black}{\texttt{Black}}
-\newcommand{\white}{\texttt{White}}
-\newcommand{\blacks}{\texttt{Black's}}
-\newcommand{\whites}{\texttt{White's}}
+\newcommand{\hblack}{\texttt{Black}}
+\newcommand{\hwhite}{\texttt{White}}
+\newcommand{\hblacks}{\texttt{Black's}}
+\newcommand{\hwhites}{\texttt{White's}}
 \newcommand{\loc}[1]{\texttt{#1}}
 
-\title{Learning to Bias MCTS Playouts}
+\title{Evolutionary Learning of Policies for MCTS Simulations}
 \numberofauthors{2}
@@
        \affaddr{Santa Cruz, California}\\
-       \email{ james.l.pettit@gmail.com }
+       \email{ jpettit@soe.ucsc.edu}
 % 2nd. author
@@
 simulation play can result in a weaker MCTS system~\cite{gelly2006modification}.
-Our Hivemind system uses evolutionary strategies to automatically learn effective rules for biasing the 
+Our Hivemind system uses evolution strategies to automatically learn effective rules for biasing the 
 random simulations.
@@
 
+\keywords{Monte Carlo Tree Search, Computer Hex, Evolution Strategies, Evolutionary Learning, Adversarial Search, Monte Carlo}
+
 \section{Introduction}
-Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications.
-The game framework provides immediate feedback on algorithm performance.
-The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels.
+Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
+%The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels. 
 Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. 
@@
 Chinook went so far as to solve checkers, proving that perfect play will end in a draw.
-Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can be effective.
+Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can beat the best human players.
 
-In contrast, cutting and connecting games like Go, Hex, and Havannah, have much larger search trees and long-range interactions make it difficult to create accurate evaluation functions.
-In these domains, Monte-Carlo Tree Search (MCTS) has proven effective~\cite{}, 
-with the world champion Go and Hex players both using MCTS~\cite{}.
+In contrast, cutting and connecting games (like Go, Hex, and Havannah)
+have much larger search trees and long-range interactions make it difficult to create accurate evaluation functions.
+In these domains, Monte-Carlo Tree Search (MCTS) has proven effective, 
+with the computer world champion Go and Hex players both using MCTS.
 MCTS picks its next move by growing a partial game tree while using
@@
 Each simulation first uses a child-selection policy to navigate through the partial game tree.
-After reaching a leaf of the partial game tree, a randomized continuation policy is used to complete the game 
-and determine a winner.  
+After reaching a leaf of the partial game tree, a randomized continuation policy is used to play out the rest of the game 
+so a winner can be determined.
 Finally, an update policy is used to update the partial game tree with the result of the simulation.
@@
 
-MoGo specifically, and Go programs in general, have the benefit of a rich history of play.
-Thousands of expert games are available for analysis.
+Go programs have the benefit of a rich history of play.
+Thousands of Go expert games are available for analysis.
 Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}.
@@
 and different weight parameters can have synergistic effects.
-Small changes in overall algorithm strength require thousands of games of testing to detect. 
-Often the improvement made by a small change in weight can be so small it will go unnoticed. 
-Automatic weight tuning can improve this process, but still requires thousands of games to test. 
+Detecting small changes in overall algorithm strength require thousands of testing games on a full-sized board, and
+the improvement made by a small change in weight can go unnoticed.
+Automatic weight tuning can improve this process, but still requires thousands of learning games. 
 Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. 
@@
 Because it relies on self-play, Hivemind avoids the need for extensive expert knowledge or a strong outside player.
+By using an evolutionary approach, Hivemind can handle the subtle interactions between weights that can have a strong effect on overall playing strength.
 Although our experiments have focused on Hex, 
@@
 both Go and Hex.
+
+Our experiments answer several questions, including: \\
+$\bullet$  How beneficial are the learned policies ?  \\
+$\bullet$   How stable is the evolutionary learning process? \\
+$\bullet$   Do the learned policies work on different board sizes?
 
@@
 
-The evolutionary strategies approach used by Hivemind is surprisingly stable.
+The evolution strategies approach used by Hivemind is surprisingly stable.
 In 10 different evolutionary runs, the win rates over the unbiased continuation policy remained
@@
 \label{s:hex}
-Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by  a color, either \black{} and \white. 
-The players alternate placing markers of their color on empty hexagons. 
+Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons.
+The players are \hblack{} and \hwhite{} and alternate placing markers of their color on empty hexagons.
 The object of the game is to form an unbroken chain from one side of the board to the other. 
@@
 the order in which the markers were played and have no effect on the game.
-Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move (71), completing an unbroken chain connecting both his sides.
+Figure \ref{fig:13x13finished} shows the final position, \hblack{} having just played the winning move (71), completing an unbroken chain connecting both his sides.
 
-Hex is a cutting and connecting game that
-has properties that make it attractive for AI research. 
+Hex is a cutting and connecting game whose high computational complexity and simple scoring mechanism makes it attractive for AI research. 
 Once placed, a piece never leaves the board. 
@@
 Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the winning strategy.
-Solutions for all opening positions for 7x7 and 8x8 boards have been found~\cite{henderson2009solving}. 
-A good portion, but not all of the 9x9 openings have also been solved. 
-Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.
+Solutions for all opening moves on a 7x7 and 8x8 board have been found~\cite{henderson2009solving}. 
+Of the 81 possible 9x9 openings, only 13 remain unsolved (taking symmetry into account). 13x13 is the default board size in computer Hex competitions.
 
-There are three exceptionally strong Hex programs.  Gabor Melis' \emph{Six} is the previous world champion (up to 2007). 
+There are three Hex programs that have historically competed for world champion status in the Computer Olympiad. Gabor Melis' \emph{Six} is the previous world champion (up to 2007). 
 Six uses traditional alpha-beta search with an evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)~\cite{anshelevich2002hierarchical}. 
@@
 
-Go is a much older game than Hex or Havannah, and much research has been done on Go playing algorithms. 
-Despite its simpler rules, Hex has a similar branching factor and position evaluation complexity as Go. 
-Although we emphasize Hex here, the techniques discussed generalize to any large branching game played on a regular board
+%Go is a much older game than Hex, and much research has been done on Go playing algorithms. 
+Despite its simpler rules, Hex has a similar branching factor as Go and long-range interactions make it difficult to 
+evaluate Hex positions.
+Although we emphasize Hex here, the techniques we propose generalize to any cutting and connecting game
 (such as Go or Havannah).
@@
 	}
-	\caption{13x13 Hex Game, \white{} to play.}
+	\caption{13x13 Hex Game, \hwhite{} to play.}
 	\label{fig:13x13inprogress}
@@
 	}
-	\caption{Finished 13x13 Hex Game, \black{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
+	\caption{Finished 13x13 Hex Game, \hblack{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
 	\label{fig:13x13finished}
@@
 %	}
-	\includegraphics[width=3.0in]{graphics/tree.png}
+	\includegraphics[width=3.0in]{graphics/tree.pdf}
 	\end{center}
@@
 
-The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann~\cite{brugmann1993monte}.
+The basic Monte-Carlo technique for Computer Go was proposed by Br\"{u}gmann~\cite{brugmann1993monte}.
 His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value.
-To contrast, MCTS uses Monte-Carlo simulations to grow a partial  search tree.
+Later MCTS techniques use Monte-Carlo simulations to grow a partial  search tree. 
 The UCT algorithm was developed alongside MCTS to direct the growth of the search tree~\cite{gelly2006exploration}. 
@@
 
-All nodes in the tree correspond to a fixed board position, i.e. the occupancy state of the board and which player is next to move. 
+All nodes in the tree correspond to a fixed board position, i.e.~the occupancy state of the board and which player is next to move. 
 Nodes also have statistics associated with their position:  the number of winning simulations performed that used the node, and the total number of simulations that included the node. 
@@
 The simulation continues from the position represented by this leaf node until a winner is determined.
-The result of the simulation used to update the statistics at nodes in the tree.
+The result of the simulation is used to update the statistics at nodes in the tree
+(see Listing 1).
 If the leaf node has been visited enough times (50 in our case), then it is expanded and nodes for all possible moves from 
@@
 \end{itemize}
+
+Gelly and others have shown that using heuristics in both the child-selection and continuation policies 
+can prove extremely beneficial~\cite{gelly2006modification,gelly2008achieving} for Go.
+However, this requires expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games).
+
 Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position. In the example, the child-selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}. 
-The continuation policy ran, completing the game from the positiion with \texttt{E4} occupied by \black\ and \texttt{A1} occupied by \white.
-This resulted in a loss for \black. 
-The update policy choose to update all nodes along the visited path, incrementing the wins for \white\ moves and incrementing the losses for \black\ moves.
+The continuation policy ran, completing the game from the positiion with \texttt{E4} occupied by \hblack\ and \texttt{A1} occupied by \hwhite.
+This resulted in a loss for \hblack. 
+The update policy choose to update all nodes along the visited path, incrementing the wins for \hwhite\ moves and incrementing the losses for \hblack\ moves.
 
-\begin{lstlisting}[float,frame=single,language=Python,caption=MCTS Algorithm Pseudocode]
-def search(root):
+\begin{lstlisting}[float,frame=single,language=Pascal,caption=MCTS Algorithm Pseudocode]
+function search(root):
   for i in range(simulations):
-    step(root)
+    simulate_from (root)
 
-def play_from(node) returns outcome:
-  if node.visits < threshold:
-    outcome = simulate(node)
+function simulate_from(node) returns outcome
+  if node is leaf
+    outcome =  continue_from (node)
   else:
-    child = select(node)
-    outcome = play_from(child)
-  if outcome is win for node.player:
+    child = select_child (node)
+    outcome = simulate_from (child)
+  node.visits++
+  if outcome is  node.player win
     node.wins++
-  node.visits++
+  if (node.visits ==  50) expand (node)
   return outcome
@@
 \subsection{Child Selection Policy}
-The \texttt{select} function defines the child selection policy. It must return the child of the given node to step into next. 
+The \verb+select_child+ function defines the child selection policy. It must return the child of the given node to step into next. 
 In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
 It is important to ``exploit" those child nodes with high win-rates, both
-to discover possible weaknesses that would result in the opponent winning and to obtain accurate win-rate estimates. 
+to discover possible weaknesses and to obtain highly accurate win-rate estimates. 
 Simultaneously, a child with a low number of visits must also be explored since a low win-rate
@@
 
-The \texttt{continuation} function plays out the rest of the game from the current position. 
+The \verb+continue_from+ function plays out the rest of the game from the current position. 
 The simplest policy, which we call the \emph{default policy},
@@
 The continuation policy has a large effect on the strength of the MCTS player.
+In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in its child selection policy,
+putting most of the control of the search in the hands of the continuation policy~\cite{gelly2007combining}.
 % and the Go literature has a large body of work dedicated to improving it.
@@
 It was quickly discovered that more aggressive update strategies are often profitable.
-One simple example of this is the All-Moves-As-First (AMAF) heuristic~\cite{}
+One simple example is the All-Moves-As-First (AMAF) heuristic~\cite{brugmann1993monte}
 that exploits the fact that moves played in a simulation could have been played in any order. 
@@
 In addition to updating the statistics of the nodes in the path, all children of nodes on the path corresponding to moves made later
-in the simulation have their AMAF win and visit counts updated. 
-Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
+in the simulation also have their AMAF win and visit counts updated. 
+%Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
 
@@
 % TODO
-\textbf{does MoHex do something fancier?}
+%\textbf{does MoHex do something fancier?}
 % MoHex uses one local pattern - bridge preservation, and AMAF updates with RAVE (http://webdocs.cs.ualberta.ca/~hayward/hex/#mohex)
@@
 This number is a parameter in Hivemind, and our experiments have set it to the (untuned) value of 50.
-
-\subsection{Biasing the Continuation -- Incorporate into earlier subsections}\label{bias}
-
-Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial~\cite{gelly2006modification,gelly2008achieving}.
-However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games).
-They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search.
-For example, the playout policy can implicitly prune good moves, never allowing them to be played.
-In Go, the policy must correctly handle complicated situations like nakade (life and death).
-If it fails to correctly play out a complex position, the tree might fail to see an opponent response.
-Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situations.
-If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive number of simulations.
-In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in its child selection policy~\cite{gelly2007combining}.
-This puts most of the control of the search in the hands of the simulation policy.
-%
-%\section{Problem Statement  -- delete? Incorporate points into MCTS section?}\label{problem}
-%MoGo specifically, and Go programs in general, have the benefit of a rich history of play.
-Thousands of expert games are available for analysis.
-Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}.
-More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move.
-Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.
-%
-%It would be ideal to automatically learn a better playout policy from self-play.
-The usual process of improvement requires hand-tuned weights that modify the playout policy.
-Any change must be tested to detect the change in strength it gives.
-Problematically, small changes in overall algorithm strength require thousands of games of testing to detect.
-Many times the improvement made by a small change in weight can be so small it will go unnoticed.
-Automatic weight tuning can improve this process, but still requires thousands of games to test.
-Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}.
-This thesis presents a slightly different approach using evolutionary learning.
-It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
-%
 
@@
 The major contribution of Hivemind is the integration of evolutionary learning to automatically learn appropriate continuation policies.
-Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
-Preliminary work tried learning with particle swarm optimization, but we found that evolutionary strategies gave
+% Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
+Preliminary work tried learning with particle swarm optimization, but we found that evolution strategies gave
 superior results.
 
-Using evolutionary strategies to learn continuation policies requires that the continuation policies are encoded as vectors.
+Using evolution strategies to learn continuation policies requires that the continuation policies are encoded as vectors.
 We first describe this encoding.  Section~\ref{s:tracker} describes the Game Tracker module, and Section~\ref{s:es} describes the
@@
 % TODO
-\textbf{How about tenuki weights? does hivemind consider them?}
+%\textbf{How about tenuki weights? does hivemind consider them?}
 % Early experiments with tenuki weights were less successful. The default policy is only used if no legal moves exist in the local area.
@@
 	\begin{center}
-	\includegraphics[width=3.0in]{graphics/local-pattern.png}
-	\caption{A 4x4 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
+	\includegraphics[width=3.0in]{graphics/local-pattern.pdf}
+	\caption{A 4x4 board with the 3 patterns shown around \hblacks\ last move. Example weights are shown in the center vertex.}
 	\label{fig:localpattern}
@@
 	\begin{center}
-	\includegraphics[width=3.0in]{graphics/weight-pattern-map.png}
+	\includegraphics[width=3.0in]{graphics/weight-pattern-map.pdf}
 	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
@@
 The additional overhead is modest and the larger hashmaps  have space for additional properties (such as distance to a side).
-Note that in Figure~\ref{fig:encoding} the encoding includes a color bit in the 31st position. This denotes the color to play (either \black\ (1) or \white\ (0)). 
-This bit is important because \black\ and \white\ have different goals -- they are trying to connect along different diagonals. 
+Note that in Figure~\ref{fig:encoding} the encoding includes a color bit in the 31st position. This denotes the color to play (either \hblack\ (1) or \hwhite\ (0)). 
+This bit is important because \hblack\ and \hwhite\ have different goals -- they are trying to connect along different diagonals. 
 By using a reflection and flipping colors, one position can be translated into a symmetric  position for the opposite color. 
@@
 
-The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. It also constrains the space of games that can be easily implemented in Hivemind. 
-In particular, Hivemind allows for 2 player games with perfect information. 
-The game space is not required to be a regular board. 
-Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children.
-Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.
+The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. 
+%It also constrains the space of games that can be easily implemented in Hivemind. 
+In particular, Hivemind is well suited for 2 player placement games with perfect information. 
+Since node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children,
+Hivemind is not constrained to games with regular boards.
 
-\subsection{Evolution Strategies}\label{s:es}
+\subsection{Evolution Strategies}
+\label{s:es}
 
-Hivemind implements evolutionary strategies to learn good continuation policies.
+Hivemind implements evolution strategies to learn good continuation policies.
 Evolution strategies (ES) is a collection of algorithms that operate on real-valued vectors. 
@@
 
-Every generation, each individual plays $n$ complete games, each game played against an opponent chosen  randomly from the pool.
+Every generation, each individual plays $n=5$ complete games against opponents chosen  randomly from the pool.
 Opponents are drawn with replacement, so a given individual could play the same opponent more than once each generation. 
-At the beginning of each game, one of the two players is randomly chosen to play \black, the other to play \white. 
+At the beginning of each game, one of the two players is randomly chosen to play \hblack, the other to play \hwhite. 
 A swap-safe move was used to play the first move to greatly reduce the first-mover advantage. 
 The winner of each game receives 1 point added to their fitness, the loser has 1 point deducted from their fitness.
-% TODO
-\textbf{This means that individuals expect to play $2n$ games total, but the exact number will vary?}
-% Correct
+This means that individuals will be evaluated on about $10$ games, but the exact number depends on how often they are
+chosen as opponents.
 
@@
 The \texttt{Winner()} function is used to determine the current winner of the game.
-The special \texttt{Empty} player is returned while the game is in progress, and either \black{} or \white{} once a winner has been determined.}
+The special \texttt{Empty} player is returned while the game is in progress, and either \hblack{} or \hwhite{} once a winner has been determined.}
 %\item{Playing out the game from the current game state.
@@
 Each evaluation game was played on a 7x7 board and used 1000 simulations per move in the fitness evaluation games. 
-The individual from the 100th generation having the highest fitness was selected as the \emph{learned} continuation policy.
-The learned continuation policy, as well as the three basic continuation policies described in Section~\ref{s:policies}, were coupled with
+The individual from the 100th generation having the highest fitness was selected as the \emph{learned continuation policy}.
+This learned continuation policy, as well as the three basic continuation policies described in Section~\ref{s:policies}, were coupled with
 a slightly modified UCT child selection policy that included the variance term suggested by Gelly 2006~\cite{gelly2006exploration} and a
@@
 Each match consisted of 200 games on a 7x7 board where the first player alternated was forced to start with a
-swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.}
+swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.
+We used C4 in our 7x7 games and C8 in the 11x11 games.}
 move.
@@
 
-A natural concern is that the evolutionary strategies could have large variations in their end results and our first learned policy was
+A natural concern is that the evolution strategies could have large variations in their end results and our first learned policy was
 just a ``lucky'' happenstance. 
@@
 
-The 10 learned policies were tightly clustered with a mean of 90.75\% and a standard deviation of only 2.14\%, with a mean of 90.75\%.
-This surprising result indicates that the evolutionary processes is far more stable than we expected.
+The 10 learned policies were tightly clustered with a mean of 90.75\% and a standard deviation of only 2.14\%.
+This surprising result indicates that the evolutionary process is far more stable than we expected.
 
@@
 
-A potential problem with self-learning is that one might only be learning how to exploit the weaknesses of their siblings, and
+A potential problem with self-learning is that the system might only be learning how to exploit the weaknesses of its competitors, and
 not making a general improvement in playing strength.  
-To examine this we used an outside control:  the MoHex program~\cite{mohex} from the University of Alberta, compiled and run with its default settings~\cite{mohex}. 
-The optimization and domain expertise in MoHex make it a very strong  competitor --  MoHex has won the computer hex world championships in 2009 and 2010.
+To examine this we used an outside control:  the MoHex program from the University of Alberta, compiled and run with its default settings~\cite{mohex}. 
+The optimization and domain expertise in MoHex make it a very strong  competitor --  MoHex has won the computer hex world championships in 2008, 2009 and 2010.
 
 In the 7x7 matches against MoHex, 
-MoHex played with the default time settings of 10 seconds per move while the Hivemind programs again used
+MoHex played with the default time limit of 10 seconds per move while the Hivemind programs again used
 10,000 simulations per move. 
-\textbf{MoHex uses some of its time to perform game theoretic search and pruning before the Monte-Carlo phase}. 
+MoHex uses some of its time to perform game theoretic search and pruning before the Monte-Carlo phase. 
 The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. 
-\textbf{Although the time controls are not directly comparable, we did track CPU time and use that for a comparison of 
-the computational resources used by the various systems.  \emph{for 7x7 or larger size only?}}
-Since we were unable to force MoHex to play a particular first move, all games had the Hivemind policy moving first with a swap-safe move.
-The Hivemind win percentages (400 game matches) against MoHex are given below.
-% TODO
-% Swap-safe is C4 on 7x7, C8 on 11x11
+Although the time controls are not directly comparable, we did track CPU seconds per game.
+MoHex used about 35 CPU seconds per game while the Hivemind programs were a bit faster.
+Since we were unable to force MoHex to play a particular first move, all games had the 
+Hivemind variant moving first with the swap-safe move.  
+The Hivemind win percentages (400 game matches) and average CPU seconds used against MoHex are given below.
 
-\begin{centering}
-\begin{tabular}{ccc}
-Learned policy & Uniform local & Default policy \\
-42\% & 26\% & 11.75\%
+\medskip
+\centerline{
+\begin{tabular}{lccc}
+& Learned policy & Uniform local & Default policy \\
+win \% & 42\% & 26\% & 11.75\% \\
+cpu secs & 20.34 & 15.79 & 13.3 
 \end{tabular}
-\end{centering}
+}
 
@@
 The good performance against MoHex is a little surprising, and could be due to the relatively small board size and/or 
-the swap-safe move still favors the first player (after any first move, one of the players will have a forced win).
+a swap-safe move that favors the first player (after any first move, one of the players will have a forced win).
 Despite this, the superiority of the learned policy over the baseline ones is dramatically evident.
@@
 
-\begin{figure}[t]
-	\begin{center}
-		\begin{tabular}{c c c c}
-		& Learned \\
-		Default 11x11 & 96.8\% \\
-		Default 13x13 & 100.0\% \\
-		Uniform 11x11 & 82.76\% \\
-		Uniform 13x13 & 92.86\% \\
-		\hline
+%\begin{figure}[t]
+%	\begin{center}
+%		\begin{tabular}{c c c c}
+%		& Learned \\
+%		Default 11x11 & 96.8\% \\
+%		Default 13x13 & 100.0\% \\
+%		Uniform 11x11 & 82.76\% \\
+%		Uniform 13x13 & 92.86\% \\
+%		\hline
+%		\end{tabular}
+%		\caption{Results of evolutionary process learned on a 7x7 board, playing on an 11x11 and 13x13 board. Percentage shown is average win-rate of column player versus row player.}
+%		\label{fig:a2results}
+%	\end{center}
+%\end{figure}
+
+The time required to learn greatly depends on the boardsize used. For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage. 
+We now examine how well the policy learned on 7x7 boards generalizes to larger boards. 
+We played 200 game matches on 11x11 and 13x13 boards between the learned policy and the 
+default and uniform local policies.  
+Because of the increased boardsize,  each player used 100,000 simulated playouts per move.
+
+\centerline{
+\begin{tabular}{c | cc cc}
+ & \multispan{2}{\hfil default \hfil} & \multispan{2}{\hfil uniform local\hfil} \\
+ & 11x11 & 13x13 & 11x11 & 13x13 \\ \hline
+ learned win \%	& 92.5 \% & 94 \%  & 88.5\% & 85\% 
  \end{tabular}
-		\caption{Results of evolutionary process learned on a 7x7 board, playing on an 11x11 and 13x13 board. Percentage shown is average win-rate of column player versus row player.}
-		\label{fig:a2results}
-	\end{center}
-\end{figure}
+}
 
-The computational time required of learning process greatly depends on the boardsize used in learning.
-For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage.
-The best individual from the 100th generation of this training was used in matches against the default and uniform policy, but a boardsize of 11x11 and 13x13 was used for the matches.
-Because of the increased boardsize, players used 100,000 simulated playouts per move.
-The computational time required for these games is significantly higher than that required for the 7x7 games, so the results shown in Figure \ref{fig:a2results} are only after 30 games per match.
+The policy learned on a 7x7 board provides results that can generalize to larger boardsizes. 
+In fact, the policy's win rate increases on the larger boards (but not significantly).
+This generalization is especially useful because of the exponential increase in computational time required to play games for evolutionary learning when learning on larger board sizes.
 
-Even with the low number of games, the results shown in Figure \ref{fig:a2results} are positive.
-Learning on a 7x7 board provides results that can generalize to larger boardsizes.
-This is especially useful because of the exponential increase in computational time required to play games for evolutionary learning when learning on larger board sizes.
+On larger boards the learned policy's superiority over the baselines is also evident against MoHex.
+The following table gives the results of 200 game matches on 11x11 boards between Hivemind policies (moving first in a swap-safe location, 100,000 simulations per move) and MoHex.
 
-Figure \ref{fig:relativecpu} shows the results of these matches.
-
-\begin{figure}
-	\begin{center}
+\centerline{
 		\begin{tabular}{c  | c c c}
@@
 		\hline
-		Default & MoHex & 0\% & 555\% \\
-		Learned & MoHex & 6.5\% & 664\% \\
-		Learned & Hivemind Default & 92.5\% & 131\% \\
+		Default & MoHex & 0.5\% & 510\% \\
+		Local & MoHex & 0.0\% & 545\% \\
+		Learned & MoHex & 11\% & 620\% \\
 		\end{tabular}
-		\caption{Win-rate and relative CPU used of Hivemind versus MoHex on an 11x11 board. Relative CPU is the ratio of average CPU time per move for Hivemind vs. opponent.
-		% TODO
-		% These were on an 11x11 board, but with alternating colors for the opening position.
-I.e.
-Hivemind and Default used the swap-safe move when they played black, while MoHex was allowed any move when it was playing black.
-If only the games where Hivemind or Default played the swap-safe move are allowed, there are 100 samples, Hivemind won 11 games, and Default still won 0.
 }
-		\label{fig:relativecpu}
-	\end{center}
-\end{figure}
 
-Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 games played. 
-MoHex strongly outperformed Hivemind, but used only a fraction (about 1/6th) of the CPU time, showing a large distance to cover to get Hivemind up to competition strength. 
-In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results.
-This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 90\% with 10,000 moves.
-The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.
+The learned policy provided 22 of the 23 wins against MoHex.
+Whereas the Hivemind programs were faster on the 7x7 boards, MoHex's optimizations allow it to use only 1/5 to 1/6 the CPU time
+on the larger board.
+Note that the pattern lookup required by the learned weights slightly increases the time per simulation.
 
@@
 In practice, both resources and time are finite. 
-Bias in the playout policy can be extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search.
-Evolutionary learning can provide a substantial performance boost over naive patterns.
-With self-play and evolution, weights can be learned automatically, without requiring expert knowledge or test-and-check tuning.
-The learning process can be done offline, and the final result can be used with little overhead.
+Bias in the playout policy is extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search
+and evolutionary learning can provide a substantial performance boost over naive patterns. 
+We showed how self-play and evolution can automatically learn good weights, without requiring expert knowledge or manual tuning. 
+The learning process can be done offline, and the final result requires little overhead.
 
-It is worth emphasizing two points.
-First the learned continuation policies are coupled with relatively naive update and child-selection
+It is worth emphasizing 
+that  the learned continuation policies are coupled with relatively naive update and child-selection
 polices.  
-Second, that the learned policies are produced automatically, without domain expertise beyond the basic rules.
-Yet even so, they perform well against strong programs (at least on small board sizes).
+There is still a large gap between the automatically learned Hivemind policy and current champion programs,
+but it remains to be seen how much of this gap can be closed with improved child-selection and update policies.
 
@@
 
+%\newpage
+
 \bibliography{paper}

### 2012-03-18 thesis/paper/paper.tex
Removed old commented sections from thesis conversion.
6 TODO's left out of 13 total comments... about halfway there.

@@
 \author{ 
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
-%
 % 1st. author
@@
 Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
-%The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels. 
 Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. 
@@
 
-%Go is a much older game than Hex, and much research has been done on Go playing algorithms. 
 Despite its simpler rules, Hex has a similar branching factor as Go and long-range interactions make it difficult to 
@@
 	\begin{center}
-%	\centerline{\begin{HexBoard}[board size=5]
-%		\HGame{a3,c2}
-%	\end{HexBoard}
-%	}
 	\includegraphics[width=3.0in]{graphics/tree.pdf}
@@
 \end{figure}
-
-%As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS)~\cite{mohex}.
 
@@
 that position are added to the search tree.
+% TODO 7
+% comment: how did you come with the constant of 50?
+% answer: balancing exploration versus computational requirements and preliminary results
+% also, this was suggested as a good starting point in the computer Go mailing list (5, 10, and 50 were some common values)
 
@@
 This resulted in a loss for \hblack. 
-The update policy choose to update all nodes along the visited path, incrementing the wins for \hwhite\ moves and incrementing the losses for \hblack\ moves.
+The update policy updates all nodes along the visited path, incrementing the wins for \hwhite\ moves and incrementing the losses for \hblack\ moves.
 
@@
 putting most of the control of the search in the hands of the continuation policy~\cite{gelly2007combining}.
-% and the Go literature has a large body of work dedicated to improving it.
 Initial research in MCTS Go players focused on using a strong continuation policy, with expert-derived heuristics to select the next move~\cite{chaslot2010adding}. 
@@
 in the simulation also have their AMAF win and visit counts updated.
-%Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.
 
@@
 Our Hivemind experiments use only the basic update policy, updating win and visit counts for only those nodes on the traversed path.
-
-% TODO
-%\textbf{does MoHex do something fancier?}
-% MoHex uses one local pattern - bridge preservation, and AMAF updates with RAVE (http://webdocs.cs.ualberta.ca/~hayward/hex/#mohex)
 
@@
 The major contribution of Hivemind is the integration of evolutionary learning to automatically learn appropriate continuation policies.
-% Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
 Preliminary work tried learning with particle swarm optimization, but we found that evolution strategies gave
@@
 Thus the continuation strategies we consider correspond to an array of $2^{13}$ weights, one for each local pattern.
-% TODO
-%\textbf{How about tenuki weights? does hivemind consider them?}
-% Early experiments with tenuki weights were less successful. The default policy is only used if no legal moves exist in the local area.
 
@@
 \end{description}
+
+% TODO 9
+% comment: would it be possible to include patterns/weights from competition-level programs such as MoGo (in Go)?
+% answer: yes
 
@@
 The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. 
-%It also constrains the space of games that can be easily implemented in Hivemind. 
 In particular, Hivemind is well suited for 2 player placement games with perfect information. 
@@
 
-Our implementation of EW works as follows.
+Our implementation of ES works as follows.
 \begin{enumerate}
@@
 	\item The best 5 performing parents are propagated through without mutation.  This is done to preserve well-performing solutions.
+	% TODO 10
+	% comment: this appears to be against the spirit of natural evolution where mutation is always (presumably) present
+	% answer: 5/35 children will be unmutated, so mutation is definitely still present in the system.
+	% This is also standard in a lot the genetic algorithm literature.
 	\item Evaluate the fitness of the new pool of 40 individuals, and keep only the top 30 as parents for the next generation.
@@
 
-%\begin{figure}
-%\begin{flalign*}
-%	\tau_{t} = & \frac{1}{\sqrt{2D}} \left(1 - \frac{\textrm{generation}}{\textrm{max generations}}\right) \\
-%	\sigma_{t} = & \sigma_{t-1} e^{N(0, \tau^2)} \\
-%	p_{i, t} = & p_{i, t-1} N(0, \sigma_{t}^2)
-%\end{flalign*}
-%\caption{Differential equations for $\tau$, $\sigma$, and $\bf{p}$, indexed by time $t$. $D$ is the number of dimensions, i.e. $|\bf{p}|$}
-%\label{mutate}
-%\end{figure}
-%
-
 \subsubsection*{Fitness Evaluation}
-%\label{fitness}
 
@@
 chosen as opponents.
-
-%\subsection{The Hivemind System}
-%Hivemind was designed for experiments in learning to play abstract strategy games on a regular board.
-The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``Game Tracker interface'' used by the tree search to maintain game state information.
-The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go~\cite{golang}).
-In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games~\cite{Fuego}.
-Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe.
-The major contribution of Hivemind is the integration of evolutionary learning into the search process.
-Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
-%
-%\subsubsection{Learning}\label{learning}
-%The learning module is responsible for driving the evolutionary search process.
-In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that performs well during the evaluation step.
-Hivemind currently uses Evolution Strategies, although previous versions experimented with Particle Swarm Optimization.
-Selection and mutation are implemented as described in section \ref{estrats}.
-Evaluation is described in section~\ref{fitness}.
-%
-%\subsubsection{Search}
-%The search module is a simple generic implementation of Monte-Carlo Tree Search, using the UCT algorithm to traverse the tree.
-Section \ref{mcts} has a description of the MCTS process.
-The search module uses the Game Tracker to expand the tree and perform the random playouts.
-The entire source code for search is only about 600 lines of code.
-Using some extra information from the Game Tracker, search is configurable to use RAVE and AMAF heuristics to update multiple nodes using the results of a single playout.
-%
-%\subsubsection{Game Tracker}
-%The Game Tracker allows complete games to be played vi Monte-Carlo Tree Search. The implementation of a game has four responsibilities:
-%\begin{enumerate}
-%\item{Maintain all required game state. A single function, \texttt{Play(color, vertex)}, is used by the search tree to apply a move for the given color.}
-%\item{Report legal moves from a given game state.
-The \texttt{Legal(color, vertex)} function is used to query the legality of a given move.
-The search tree uses this to expand a parent node, adding all legal moves as children.}
-%\item{Report on the status of the winner.
-The \texttt{Winner()} function is used to determine the current winner of the game.
-The special \texttt{Empty} player is returned while the game is in progress, and either \hblack{} or \hwhite{} once a winner has been determined.}
-%\item{Playing out the game from the current game state.
-The tracker is parameterized with the policy weights, and responsible for using them during the playout.
-The speed of the playout function is also critical, as more playouts-per-second generally result in better move selection.
-This emphasis on speed is why the tracker is required to implement a monolithic \texttt{Playout} function.
-Many games have optimizations that can be performed in the playout that the search tree is unaware of, for example, the Hex tracker keeps a simple list of randomized empty positions, so random move selection is a O(1) operation.
-This is because in Hex, any empty position is legal.
-This is not the case for Go, so the Go engine uses other optimizations for fast random move selection.}
-%\end{enumerate}
-%The additional \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic.
-A few extra functions also provide for debugging by pretty-printing the tracker's internal state.
-%
-%The Game Tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework.
-It also constrains the space of games that can be easily implemented in Hivemind.
-In particular, Hivemind allows for 2 player games with perfect information.
-The game space is not required to be a regular board.
-Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children.
-Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.
+% TODO 12
+% comment: so the pool consisted of three basic policies? Why wasn't MoHex a part of the pool? In fact, the authors touch on this later in the paper:
+% "A potential problem with self-learning is that the system might only be learning how to exploit the weaknesses of its competitors, and not making
+% a general improvement in playing strength." This sentence is also problematic.
+Namely, it is unclear what a "general improvement in playing strength" is.
+A strong player against player A may be weak against player B.
+It is unclear how to define an opponent-independent measure of playing
+% strength. However, by not including MoHex in the pool during the evolution the authors did not specifically evolve their continuation policies
+% against MoHex. I wonder why.
+% answer: To preserve MoHex as an outside comparator. Learning from self-play is also the research question.
 
@@
 
-%
-%The fact that evolutionary learning can find weights that boost performance versus uniform local in the same manner that uniform 
-% local improves upon pure random shows how much further there is to go, and shows the weaknesses of purely random simulations.
-
 \subsubsection*{Stability of Evolution}
-
-%\begin{figure}[t]
-%	\begin{center}
-%		\begin{tabular}{c c}
-%		& Default \\
-%		\hline
-%		Learned 1 & 91.0\% \\
-%		Learned 2 & 90.5\% \\
-%		Learned 3 & 90.5\% \\
-%		Learned 4 & 87.5\% \\
-%		Learned 5 & 90.5\% \\
-%		Learned 6 & 93.5\% \\
-%		Learned 7 & 93.0\% \\
-%		Learned 8 & 91.5\% \\
-%		Learned 9 & 87.0\% \\
-%		Learned 10 & 92.5\% \\
-%		Min & 87.0\% \\
-%		Max & 93.5\% \\
-%		Mean & 90.75\% \\
-%		Std. Dev. & ~2.14\% \\
-%		\end{tabular}
-%		\caption{Results of 10 independent evolutionary processes versus the default policy. Percentage shown is average win-rate of row player versus column player.}
-%		\label{fig:a1results}
-%	\end{center}
-%\end{figure}
 
@@
 
-%\begin{figure}[t]
-%	\begin{center}
-%		\begin{tabular}{c c c c}
-%		& Learned \\
-%		Default 11x11 & 96.8\% \\
-%		Default 13x13 & 100.0\% \\
-%		Uniform 11x11 & 82.76\% \\
-%		Uniform 13x13 & 92.86\% \\
-%		\hline
-%		\end{tabular}
-%		\caption{Results of evolutionary process learned on a 7x7 board, playing on an 11x11 and 13x13 board. Percentage shown is average win-rate of column player versus row player.}
-%		\label{fig:a2results}
-%	\end{center}
-%\end{figure}
-
 The time required to learn greatly depends on the boardsize used. For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage.
+% TODO 13
+% comment: explain
+% answer: 
 We now examine how well the policy learned on 7x7 boards generalizes to larger boards. 
@@
 but it remains to be seen how much of this gap can be closed with improved child-selection and update policies.
+% TODO 13
+% comment: yes, I think the results of this paper would be even more interesting if either (i) only continuation policies were compared
+% e.g. by disabling all other techniques in MoHex) or (ii) other advanced techniques (e.g., child-selection policy, the update policy, etc.)
+% were added to Hivemind as well.
+% answer: Comparing just continuation policies is a tricky issue. Hivemind specifically scores candidates on winning a full game using MCTS, it does
+% not attempt to build a strong continuation policy. It is true that a strong continuation policy can lead to stronger overall play, but especially
+% the simulation balancing paper shows that strong playouts are not always the best approach to a stronger overall player.
+% As to (ii), improving Hivemind's other areas is definitely a good idea
 
@@
 
-%\newpage
-
 \bibliography{paper}

### 2012-03-23 thesis/paper/paper.tex
More cleanup addressing reviewers comments.

@@
 
-The basic Monte-Carlo technique for Computer Go was proposed by Br\"{u}gmann~\cite{brugmann1993monte}.
+The basic Monte-Carlo technique for computer Go was proposed by Br\"{u}gmann~\cite{brugmann1993monte}.
 His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value.
@@
 (see Listing 1).
-If the leaf node has been visited enough times (50 in our case), then it is expanded and nodes for all possible moves from 
+If the leaf node has been visited enough times (50 in our case\footnote{Lower numbers promote accurate exploration at the cost of larger trees and increased computation.
+Our choice of 50 is somewhat arbitrary, values from 1 to 100 are used in UCT programs on the CGOS Computer Go Server\cite{cgosbots}.}), then it is expanded and nodes for all possible moves from
 that position are added to the search tree.
-% TODO 7
-% comment: how did you come with the constant of 50?
-% answer: balancing exploration versus computational requirements and preliminary results
-% also, this was suggested as a good starting point in the computer Go mailing list (5, 10, and 50 were some common values)
 
@@
 Initial research in MCTS Go players focused on using a strong continuation policy, with expert-derived heuristics to select the next move~\cite{chaslot2010adding}. 
-The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates.
+The top computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates.
 Only when none of the rules match do they fall back to a uniform random move. 
@@
 
-There are three simple continuation policies that we use as baseline comparators for our learned policies.
+%There are three simple continuation policies that we use as baseline comparators for our learned policies.
+In principle any player could be used to evaluate the effectiveness of learning.
+We used three simple continuation policies that were easily implementable in Hivemind as baseline comparators for our learned policies.
+% TODO 9
+% comment: would it be possible to include patterns/weights from competition-level programs such as MoGo (in Go)?
+% answer: yes
+% DPH: the more I think about it the more I think I don't understand the comment.  Is it about representing MoGo patterns with the 
+% pattern encodings, or about using mogo patterns as a comparator?
 
@@
 \end{description}
-
-% TODO 9
-% comment: would it be possible to include patterns/weights from competition-level programs such as MoGo (in Go)?
-% answer: yes
 
@@
 		\end{enumerate}
-	\item The best 5 performing parents are propagated through without mutation.  This is done to preserve well-performing solutions.
-	% TODO 10
-	% comment: this appears to be against the spirit of natural evolution where mutation is always (presumably) present
-	% answer: 5/35 children will be unmutated, so mutation is definitely still present in the system.
-	% This is also standard in a lot the genetic algorithm literature.
+	\item The best 5 performing parents are propagated through without mutation.  This is commonly done to preserve well-performing solutions. 
 	\item Evaluate the fitness of the new pool of 40 individuals, and keep only the top 30 as parents for the next generation.
@@
 
-Every generation, each individual plays $n=5$ complete games against opponents chosen  randomly from the pool.
+Every generation, each individual plays $n=5$ complete games against opponents chosen  randomly from the current population.
 Opponents are drawn with replacement, so a given individual could play the same opponent more than once each generation. 
@@
 chosen as opponents.
-% TODO 12
-% comment: so the pool consisted of three basic policies? Why wasn't MoHex a part of the pool? In fact, the authors touch on this later in the paper:
-% "A potential problem with self-learning is that the system might only be learning how to exploit the weaknesses of its competitors, and not making
-% a general improvement in playing strength." This sentence is also problematic.
-Namely, it is unclear what a "general improvement in playing strength" is.
-A strong player against player A may be weak against player B.
-It is unclear how to define an opponent-independent measure of playing
-% strength. However, by not including MoHex in the pool during the evolution the authors did not specifically evolve their continuation policies
-% against MoHex. I wonder why.
-% answer: To preserve MoHex as an outside comparator. Learning from self-play is also the research question.
 
@@
 swap-safe\footnote{A swap-safe first move is a neutral or mediocre move by the first player in an attempt to eliminate the first-player advantage.
+This is similar to the \emph{Komi} rule in Go.
 We used C4 in our 7x7 games and C8 in the 11x11 games.}
@@
 
-A potential problem with self-learning is that the system might only be learning how to exploit the weaknesses of its competitors, and
-not making a general improvement in playing strength.  
-To examine this we used an outside control:  the MoHex program from the University of Alberta, compiled and run with its default settings~\cite{mohex}. 
-The optimization and domain expertise in MoHex make it a very strong  competitor --  MoHex has won the computer hex world championships in 2008, 2009 and 2010.
+A potential problem with self-learning is that the system might only be learning how to exploit the weaknesses of its competitors, and not making a general improvement in playing strength.
+We use ``general playing strength'' in an intuitive sense -- we want to avoid focusing on
+particular ``tricks'' that are only effective against the algorithms in the current population.
+Although improvement against the three baseline opponents gives some evidence against this over-specialization, 
+we perform further comparison against a strong outside control. 
+For this, we used the MoHex program from the University of Alberta, compiled and run with its default settings~\cite{mohex}.
+The optimization and domain expertise in MoHex make it a very strong  competitor --  MoHex has won the computer Hex world championships in 2008, 2009 and 2010.
 
@@
 
-The time required to learn greatly depends on the boardsize used. For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and bias toward first-mover advantage.
-% TODO 13
-% comment: explain
-% answer: 
+The time required to learn greatly depends on the boardsize used. For training, a boardsize of 7x7 was chosen as a tradeoff between computation time and the game's deepness and subtlety.
 We now examine how well the policy learned on 7x7 boards generalizes to larger boards. 
@@
 but it remains to be seen how much of this gap can be closed with improved child-selection and update policies.
-% TODO 13
-% comment: yes, I think the results of this paper would be even more interesting if either (i) only continuation policies were compared
-% e.g. by disabling all other techniques in MoHex) or (ii) other advanced techniques (e.g., child-selection policy, the update policy, etc.)
-% were added to Hivemind as well.
-% answer: Comparing just continuation policies is a tricky issue. Hivemind specifically scores candidates on winning a full game using MCTS, it does
-% not attempt to build a strong continuation policy. It is true that a strong continuation policy can lead to stronger overall play, but especially
-% the simulation balancing paper shows that strong playouts are not always the best approach to a stronger overall player.
-% As to (ii), improving Hivemind's other areas is definitely a good idea
 

### 2012-03-24 thesis/paper/paper.tex
Using acm_proc_article-sp style, had to tweak the figures and itemize/enumerate environments to fit into 8 pages.

@@
 \documentclass{acm_proc_article-sp}
+%\documentclass{sig-alternate}
 \usepackage[latin1]{inputenc}
@@
 \usepackage{havannah}
+\usepackage{enumitem}
+
 \newcommand{\hblack}{\texttt{Black}}
@@
 
+\toappear{Permission to make digital or hard copies of all or part of this work for
+   personal or classroom use is granted without fee provided that copies are
+   not made or distributed for profit or commercial advantage and that copies
+   bear this notice and the full citation on the first page. To copy otherwise,
+   to republish, to post on servers or to redistribute to lists, requires prior
+   specific permission and/or a fee. \\
+   FDG '12, May 29-June 1, 2012 Raleigh, NC, USA. \\
+   Copyright (c) 2012 ACM 978-1-4503-1333-9/12/05... \$10.00.}
+
 \maketitle
+
+\setlist{topsep=0.5pt,partopsep=0.5pt,parsep=0.5pt,itemsep=0.5pt}
 
@@
 
+\category{I.2.6}{Computing Methodologies}{Artificial Intelligence}
+
 \keywords{Monte Carlo Tree Search, Computer Hex, Evolution Strategies, Evolutionary Learning, Adversarial Search, Monte Carlo}
@@
 \section{Introduction}
+
 Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
@@
 
-Our experiments answer several questions, including: \\
-$\bullet$  How beneficial are the learned policies ?  \\
-$\bullet$   How stable is the evolutionary learning process? \\
-$\bullet$   Do the learned policies work on different board sizes?
+Our experiments answer several questions, including: 
+\begin{itemize}
+\item How beneficial are the learned policies ? 
+\item How stable is the evolutionary learning process? 
+\item Do the learned policies work on different board sizes?
+\end{itemize}
 
@@
 \begin{figure}[tb]
-\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
+\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
 		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
@@
 \begin{figure}[tb]
-\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
+\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
 		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
@@
 	\begin{center}
-	\includegraphics[width=3.0in]{graphics/tree.pdf}
+	\includegraphics[width=2.5in]{graphics/tree.pdf}
 	\end{center}
@@
 In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
-It is important to ``exploit" those child nodes with high win-rates, both
+It is important to ``exploit'' those child nodes with high win-rates, both
 to discover possible weaknesses and to obtain highly accurate win-rate estimates. 
@@
 We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
-On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move
-could actually be very good.
+On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move could actually be very good.
 
@@
 	\begin{center}
-	\includegraphics[width=3.0in]{graphics/local-pattern.pdf}
+	\includegraphics[width=2.85in]{graphics/local-pattern.pdf}
 	\caption{A 4x4 board with the 3 patterns shown around \hblacks\ last move. Example weights are shown in the center vertex.}
@@
 	\begin{center}
-	\includegraphics[width=3.0in]{graphics/weight-pattern-map.pdf}
+	\includegraphics[width=3.25in]{graphics/weight-pattern-map.pdf}
 	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
@@
 
-%There are three simple continuation policies that we use as baseline comparators for our learned policies.
-In principle any player could be used to evaluate the effectiveness of learning.
-We used three simple continuation policies that were easily implementable in Hivemind as baseline comparators for our learned policies.
-% TODO 9
-% comment: would it be possible to include patterns/weights from competition-level programs such as MoGo (in Go)?
-% answer: yes
-% DPH: the more I think about it the more I think I don't understand the comment.  Is it about representing MoGo patterns with the 
-% pattern encodings, or about using mogo patterns as a comparator?
-
-\begin{description}
-	\item[Default] - The next move is chosen uniformly at random from the set of all legal moves.
-	This policy corresponds to setting all of the weights to 0 in our encoding .
-	\item[Uniform local] - The next move is chosen uniformly at random from the empty locations adjacent to the last move.
-	If no legal move exists on these 6 points, then the move is chosen uniformly from the set of all legal moves 
-	(i.e. using the default policy).  This policy corresponds to setting all the weights to 1.
-	\item[Uniform local with ``tenuki''] - The tenuki (Japanese for  ``play away'') variant is a mixture of the Default and Uniform local policies.  
-	With probability 5/6 the move is chosen with the Uniform Local policy and with probability 1/6 it is chosen with the Default policy. 
-\end{description}
+Note that many kinds of local hand-coded patterns (like those used by MoGo \cite{gelly2006modification} for Go) can be expressed using the Hivemind encoding system.
 
@@
 In addition to the \texttt{Playout} function, the Game Tracker provides the following functionality.
+
 \begin{enumerate}
@@
 \end{enumerate}
+
 A few extra functions also provide for debugging by pretty-printing the tracker's internal state.
@@
 \subsubsection*{Benefits of learned policy}
-To determine the benefits of the learned continuation policy we played a round-robin set of matches between the learned, 
-default, uniform local, and uniform with tenuki policies.
+In principle any player could be used to evaluate the effectiveness of learning.   
+We used a round-robin tournament between the learned policy and the three following continuation policies that were easily implementable in Hivemind.
+
+\begin{description}
+	\item[Default] - The next move is chosen uniformly at random from the set of all legal moves.
+	This policy corresponds to setting all of the weights to 0 in our encoding .
+	\item[Uniform local] - The next move is chosen uniformly at random from the empty locations adjacent to the last move.
+	If no legal move exists on these 6 points, then the move is chosen uniformly from the set of all legal moves 
+	(i.e. using the default policy).  This policy corresponds to setting all the weights to 1.
+	\item[Uniform local with ``tenuki''] - The tenuki (Japanese for  ``play away'') variant is a mixture of the Default and Uniform local policies.  
+	With probability 5/6 the move is chosen with the Uniform Local policy and with probability 1/6 it is chosen with the Default policy. 
+\end{description}
+
 Each match consisted of 200 games on a 7x7 board where the first player alternated was forced to start with a
@@
 
-A natural concern is that the evolution strategies could have large variations in their end results and our first learned policy was
-just a ``lucky'' happenstance. 
+A natural concern is that the evolution strategies could have large variations in their end results and our first learned policy was just a ``lucky'' happenstance. 
 To protect against this, we re-ran the learning process described above another 9 times, creating a total of 10 different learned continuation policies.
@@
 The learned policy provided 22 of the 23 wins against MoHex.
-Whereas the Hivemind programs were faster on the 7x7 boards, MoHex's optimizations allow it to use only 1/5 to 1/6 the CPU time
-on the larger board.
+Whereas the Hivemind programs were faster on the 7x7 boards, MoHex's optimizations allow it to use only 1/5 to 1/6 the CPU time on the larger board.
 Note that the pattern lookup required by the learned weights slightly increases the time per simulation.
@@
 In practice, both resources and time are finite. 
-Bias in the playout policy is extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search
-and evolutionary learning can provide a substantial performance boost over naive patterns. 
+Bias in the playout policy is extremely useful for improving the overall quality and performance of a Monte-Carlo Tree Search and evolutionary learning can provide a substantial performance boost over naive patterns.
 We showed how self-play and evolution can automatically learn good weights, without requiring expert knowledge or manual tuning.
@@
 
-It is worth emphasizing 
-that  the learned continuation policies are coupled with relatively naive update and child-selection
-polices.  
+It is worth emphasizing that  the learned continuation policies are coupled with relatively naive update and child-selection polices.  
 There is still a large gap between the automatically learned Hivemind policy and current champion programs,
@@
 
+\subsection*{Acknowledgements}
+We thank David Doshay and the Hierarchical Systems Research Foundation for the use of the their compute cluster. \\
+
 \bibliography{paper}

### 2012-03-24 thesis/paper/paper.tex
Changed 13x13 figures to 7x7 for better visibility. Larger fonts on the weight-pattern-map figure.

@@
 The object of the game is to form an unbroken chain from one side of the board to the other. 
-Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board, note that the numbers indicate
+Figure \ref{fig:inprogress} shows an example game in progress on a 7x7 board, note that the numbers indicate
 the order in which the markers were played and have no effect on the game.
-Figure \ref{fig:13x13finished} shows the final position, \hblack{} having just played the winning move (71), completing an unbroken chain connecting both his sides.
+Figure \ref{fig:finished} shows the final position, \hblack{} having just played the winning move (25), completing an unbroken chain connecting both his sides.
 
@@
 \begin{figure}[tb]
-\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
-		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
+\resizebox{3.3in}{!}{	\begin{HexBoard}[board size=7]
+		\HGame{d4,d3,e3,f1,c2,d6,b5,e2}
 	\end{HexBoard}
 	}
-	\caption{13x13 Hex Game, \hwhite{} to play.}
-	\label{fig:13x13inprogress}
+	\caption{7x7 Hex Game, \hwhite{} to play.}
+	\label{fig:inprogress}
 \end{figure}
@@
 \begin{figure}[tb]
-\resizebox{3.35in}{!}{	\begin{HexBoard}[board size=13]
-		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
-				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
-				   e8,e9,c10,k8,c9,g6,g7,c11,
-				   d10,h6,h7,i5,j5,j4,k4,k3,
-				   m2,l3,m3,l4,c6,e11,l5,m4,
-				   d11,b7,e6,d12,c12,c7,e7,f5,
-				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
-				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
-				   c13,c5,b6}
+\resizebox{3.3in}{!}{	\begin{HexBoard}[board size=7]
+		\HGame{d4,d3,e3,f1,c2,d6,b5,e2,c4,c3,a4,b6,c6,c5,d5,b7,c7,b4,a5,b3,a3,b1,b2,c1,d1}
 	\end{HexBoard}
 	}
-	\caption{Finished 13x13 Hex Game, \hblack{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
-	\label{fig:13x13finished}
+	\caption{Finished 7x7 Hex Game, \hblack{}'s play 25 (at \loc{d1}) gives him an unbroken path from top to bottom.}
+	\label{fig:finished}
 \end{figure}

### 2012-05-08 presentation/presentation.tex
FDG 2012 presentation.

\documentclass{beamer}
\usetheme{Boadilla}
\usepackage{havannah}

\title[Learning MCTS]{Evolutionary Learning of Policies for MCTS Simulations}
\author[Pettit, Helmbold]{James Pettit, David Helmbold}
\institute[UCSC]{
  University of California, Santa Cruz\\
  \texttt{jpettit@soe.ucsc.edu}
}
\date[July 2012]{July 2012}

\begin{document}

\begin{frame}[plain]
  \titlepage
\end{frame}

\begin{frame}{Overview}
\end{frame}

\begin{frame}{The Game of Hex}
% example board, play a few moves, show final position
% useful for experimentation:
% 1. Easy to program
% 2. Clear-cut winning condition (not so for Go)
% 3. Large problem space
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}
\begin{frame}{The Game of Hex}
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4,d3}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}
\begin{frame}{The Game of Hex}
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4,d3,e3}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}
\begin{frame}{The Game of Hex}
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4,d3,e3,f1}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}
\begin{frame}{The Game of Hex}
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4,d3,e3,f1,c2}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}
\begin{frame}{The Game of Hex}
	\begin{figure}[tb]
	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
		  \HGame{d4,d3,e3,f1,c2,d6,b5,e2,c4,c3,a4,b6,c6,c5,d5,b7,c7,b4,a5,b3,a3,b1,b2,c1,d1}
		\end{HexBoard}
		}
	\end{figure}
\end{frame}

\begin{frame}{MCTS - Overview}
% Minimax game tree cannot be fully constructed
% Use random playouts to estimate minimax value
% Large enough playouts will converge to true minimax value
\end{frame}

\begin{frame}{MCTS - Playout Policy}
% One of the ways to improve the quality of the estimation
\end{frame}

\begin{frame}{Evolutionary Learning}
\end{frame}

\begin{frame}{Encoding}
\end{frame}

\begin{frame}{Results}
\end{frame}

\begin{frame}{Future Work}
\end{frame}

\end{document}

### 2012-05-09 presentation/presentation.tex
More stuff to talk about.

@@
 \begin{frame}{Overview}
+\begin{enumerate}
+	\item The Game of Hex
+	\item Monte-Carlo Tree Search (MCTS)
+	\item Apply Evolutionary Learningto MCTS+Hex
+	\item Results and Future Work
+\end{enumerate}
 \end{frame}
@@
 \begin{frame}{The Game of Hex}
+\begin{itemize}
+	\item 2 player, perfect information
+	\item 6-sided hexagons on a parallelogram board
+	\item Common sizes: 11, 13
+\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Board}
 % example board, play a few moves, show final position
-% useful for experimentation:
-% 1. Easy to program
-% 2. Clear-cut winning condition (not so for Go)
-% 3. Large problem space
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4}
-		\end{HexBoard}
-		}
-	\end{figure}
-\end{frame}
-\begin{frame}{The Game of Hex}
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4,d3}
-		\end{HexBoard}
-		}
-	\end{figure}
-\end{frame}
-\begin{frame}{The Game of Hex}
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4,d3,e3}
-		\end{HexBoard}
-		}
-	\end{figure}
-\end{frame}
-\begin{frame}{The Game of Hex}
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4,d3,e3,f1}
-		\end{HexBoard}
-		}
-	\end{figure}
-\end{frame}
-\begin{frame}{The Game of Hex}
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4,d3,e3,f1,c2}
-		\end{HexBoard}
-		}
-	\end{figure}
-\end{frame}
-\begin{frame}{The Game of Hex}
 	\begin{figure}[tb]
@@
 
-\begin{frame}{MCTS - Overview}
-% Minimax game tree cannot be fully constructed
-% Use random playouts to estimate minimax value
-% Large enough playouts will converge to true minimax value
+\begin{frame}{The Game of Hex - Good for AI}
+\begin{itemize}
+	\item Easy to program
+	\item Clear-cut winning condition
+	\item Large problem space
+	\item Solved for boards up to 7x7
+\end{itemize}
 \end{frame}
 
-\begin{frame}{MCTS - Playout Policy}
-% One of the ways to improve the quality of the estimation
+\begin{frame}{Tree Search}
+\begin{itemize}
+	\item Game tree grows exponentially
+	\item Symmetry can halve space
+	\item Limited opportunities for provable pruning
+	\item No good position ranking heuristic
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Monte Carlo Tree Search}
+\begin{itemize}
+	\item Use random playouts to estimate minimax value
+	\item Large enough playouts will converge to true minimax value
+	\item Seems dumb, actually works
+	\item Computationally very expensive
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Monte Carlo Tree Search - Playout Policy}
+\begin{itemize}
+	\item Naively ``improving'' the strength of the playout policy can hurt overall performance
+	\item Requires careful and expensive testing to verify improvement
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Evolutionary Learning}
+\begin{itemize}
+	\item Idea: Evolve playout policy
+	\item Individual policies compete in a tournament to reproduce
+	\item Self-play yields a self-bootstrapping system
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Encoding}
+\begin{itemize}
+	\item Individual policy is a collection of evolvable weights
+	\item One weight, many interpretations
+	\item Explicitly requires domain knowledge
+	\item Implicitly limits the system
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Results}
+\begin{itemize}
+	\item Strong and robust
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Future Work}
+\begin{itemize}
+	\item Different encodings
+	\item General game play
+	\item Encode expert knowledge in weights
+\end{itemize}
 \end{frame}

### 2012-05-10 presentation/presentation.tex
Added figures and results.

@@
 
-\begin{frame}{Evolutionary Learning}
+\begin{frame}{Evolutionary Learning (Hivemind)}
 \begin{itemize}
@@
 \end{itemize}
+\end{frame}
+
+\begin{frame}{Evolution}
+\begin{enumerate}
+	\item Generate $n$ individuals (population)
+	\item Evaluate fitness\label{eval}
+	\item Rank by fitness
+	\item Select top $c$ children, $c < n$
+	\item Recombine children into $n$ new individuals (next generation)
+	\item Mutate population
+	\item Repeat from \ref{eval}
+\end{enumerate}
 \end{frame}
@@
 
-\begin{frame}{Results}
+\begin{frame}{Encoding - Example}
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=3.25in]{graphics/local-pattern.pdf}
+  \label{fig:encoding}
+  \end{center}
+\end{figure}
 \begin{itemize}
-	\item Strong and robust
+	\item Weight moves local to last-played move
+	\item If local area is filled, use default policy
 \end{itemize}
+\end{frame}
+
+\begin{frame}{Encoding - Example}
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=3.25in]{graphics/weight-pattern-map.pdf}
+  \label{fig:encoding}
+  \end{center}
+\end{figure}
+\begin{itemize}
+	\item Map from 32-bit integer to floating point weight
+	\item Evolution (mutation/recombination) operates on individual's map entries
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Results (Self)}
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c | c c c}
+		& \multispan{3}{\hfil opponent \hfil} \\
+		 player & default & uniform & uniform (tenuki) \\
+		\hline
+		uniform & 70.50\% & & \\
+		uniform (tenuki) & 61.00\% & 50.00\% & \\
+		learned & 90.00\% & 84.00\% & 86.00\% \\
+		\end{tabular}
+	\label{fig:results}
+	\end{center}
+\end{figure}
+All-play-all tournament of the 4 Hivemind variants.
+
+Each element is the percent win-rate of the row variant versus the column variant.
+\end{frame}
+
+\begin{frame}{Results (Self)}
+\centerline{
+\begin{tabular}{c | cc cc}
+ & \multispan{2}{\hfil default \hfil} & \multispan{2}{\hfil uniform local\hfil} \\
+ & 11x11 & 13x13 & 11x11 & 13x13 \\ \hline
+ learned win \%	& 92.5 \% & 94 \%  & 88.5\% & 85\% 
+ \end{tabular}
+}
+\end{frame}
+
+\begin{frame}{Results (MoHex 7x7)}
+\centerline{
+\begin{tabular}{lccc}
+& Learned policy & Uniform local & Default policy \\
+win \% & 42\% & 26\% & 11.75\%
+\end{tabular}
+}
+\end{frame}
+
+\begin{frame}{Results (MoHex 11x11)}
+\centerline{
+		\begin{tabular}{c  | c c c}
+		Hivemind & Win-rate & Relative CPU \\
+		\hline
+		Default & 0.5\% & 510\% \\
+		Local & 0.0\% & 545\% \\
+		Learned & 11\% & 620\% \\
+		\end{tabular}
+}
 \end{frame}

### 2012-05-27 presentation/presentation.tex
Best copy yet.

--- 2012-05-09 presentation/presentation.tex
@@
 \usepackage{havannah}
+\usepackage{color}
+\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
 
@@
 
-\begin{frame}{Overview}
+\begin{frame}{Motivation}
 \begin{enumerate}
-	\item The Game of Hex
-	\item Monte-Carlo Tree Search (MCTS)
-	\item Apply Evolutionary Learningto MCTS+Hex
-	\item Results and Future Work
+	\item Cutting and connecting games are hard
+	\item Computer Go - Next AI Grand Challenge
+	\item Monte Carlo techniques promising
+	\item Tuning policies difficult
+	\item Solution: use evolution and self-play to learn better policies 
 \end{enumerate}
@@
 
-\begin{frame}{The Game of Hex}
+\begin{frame}{The Game of Hex - Good for AI}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{}
+		\end{HexBoard}
+		}
+	\end{figure}
 	\begin{itemize}
 		\item 2 player, perfect information
-	\item 6-sided hexagons on a parallelogram board
-	\item Common sizes: 11, 13
+		\item Easy to program
+		\item Clear-cut winning condition
+		\item Large problem space
+		\item Solved for boards up to 7x7
+		\item Common sizes: 11x11, 13x13
 	\end{itemize}
@@
 
-\begin{frame}{The Game of Hex - Example Board}
-% example board, play a few moves, show final position
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4,d3}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4,d3,e3}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4,d3,e3,f1}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4,d3,e3,f1,c2}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
+	\begin{figure}[tb]
+	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
+		  \HGame{d4,d3,e3,f1,c2,d6}
+		\end{HexBoard}
+		}
+	\end{figure}
+	\begin{itemize}
+		\item Connect opposing sides
+		\item Cut opponents connections
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Game of Hex - Example Game}
 	\begin{figure}[tb]
@@
 	\end{figure}
-\end{frame}
-
-\begin{frame}{The Game of Hex - Good for AI}
 	\begin{itemize}
-	\item Easy to program
-	\item Clear-cut winning condition
-	\item Large problem space
-	\item Solved for boards up to 7x7
+		\item Connect opposing sides
+		\item Cut opponents connections
 	\end{itemize}
@@
 	\item Game tree grows exponentially
-	\item Symmetry can halve space
+	\begin{itemize}
+		\item Hex on a general board is PSPACE-complete
+	\end{itemize}
+	\item No effective position evaluation heuristic
+	\begin{itemize}
+		\item Previous champion until 2006, (\emph{Six}, used circuit simulation
+	\end{itemize}
 	\item Limited opportunities for provable pruning
-	\item No good position ranking heuristic
+	\begin{itemize}
+		\item Current champion, \emph{MoHex}, does lots of this
+	\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
+	\item Grow tree dynamically
 	\item Use random playouts to estimate minimax value
-	\item Large enough playouts will converge to true minimax value
-	\item Seems dumb, actually works
-	\item Computationally very expensive
+	\item In the limit, converges to true minimax value
+	\item Simple idea, remarkably successful
+	\item Computationally expensive
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Monte Carlo Tree Search - Overview}
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=2.25in]{graphics/tree.pdf}
+  \label{fig:encoding}
+  \end{center}
+\end{figure}
+\begin{itemize}
+	\item \emph{Select} - using win/loss statistics, traverse tree to a leaf
+	\item {\bf Playout} - from selected position, \emph{randomly} play out game until final position
+	\item \emph{Update} - using results of playout, update win/loss statistics
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Monte Carlo Tree Search - Playout Policies}
+\begin{itemize}
+	\item Playout policy a critical part of overall performance
+	\item We examine 4 different policies:
+	\begin{itemize}
+		\item Default - uniform random
+		\item Uniform Local - play in local neighborhood first
+		\item Uniform Local with Tekuki (Play-Away) - possibly play in local neighborhood first
+		\item Local Pattern Weighted - stochastic, weighted local neighborhood
+	\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Weighting the Playout Policy}
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=3.25in]{graphics/local-pattern.pdf}
+  \label{fig:encoding}
+  \end{center}
+\end{figure}
+\begin{itemize}
+	\item Use patterns to weight moves
+	\item Weight moves local to last-played move
+	\item If local area is filled, use default policy
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Example Encoding}
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=3.25in]{graphics/weight-pattern-map.pdf}
+  \label{fig:encoding}
+  \end{center}
+\end{figure}
+\begin{itemize}
+	\item Fast lookup table mapping pattern to weight
+	\item Plenty of unused space in upper bits for storing extra info
 \end{itemize}
@@
 \begin{itemize}
+	\item Execution time is critical
+	\item Requires deep insight to develop weights, then careful and expensive testing to verify improvement
 	\item Naively ``improving'' the strength of the playout policy can hurt overall performance
-	\item Requires careful and expensive testing to verify improvement
+	\begin{itemize}
+		\item Subtle interaction between selection, update, and playout
+		\item Example: fully deterministic policy equivalent to an evaluation function
+	\end{itemize}
+	\item We want a technique to learn weights automatically
 \end{itemize}
@@
 	\item Idea: Evolve playout policy
-	\item Individual policies compete in a tournament to reproduce
-	\item Self-play yields a self-bootstrapping system
+	\item Policy = set of all possible local patterns + weights for those patterns
+	\item Individual policies compete via self-play to propagate
+	\item Best individuals are selected for reproduction
+	\item Mutation/recombination operates on individual's pattern weights
+	\item Individuals play \emph{complete} games against each other using \emph{complete} MCTS system
 \end{itemize}
@@
 
-\begin{frame}{Encoding}
+\begin{frame}{Evolution}
 \begin{itemize}
-	\item Individual policy is a collection of evolvable weights
-	\item One weight, many interpretations
-	\item Explicitly requires domain knowledge
-	\item Implicitly limits the system
+	\item Init
+	\begin{enumerate}
+		\item Generate $n$ individuals (population)
+	\end{enumerate}
+	\item Each Generation
+	\begin{enumerate}
+		\item Evaluate fitness
+		\item Rank by fitness
+		\item Select top $c$ individuals, $c < n$
+		\item Recombine individuals into $n$ new children
+		\item Mutate children
+	\end{enumerate}
 \end{itemize}
@@
 
-\begin{frame}{Results}
+\begin{frame}{Results of learning after training on 7x7)}
+\begin{figure}
+	\caption{Trained on 7x7 board}
+	\begin{center}
+		\begin{tabular}{c | c c c}
+		& \multispan{3}{\hfil opponent \hfil} \\
+		 player & default & uniform local & uniform local (tenuki) \\
+		\hline
+		uniform local & 70.50\% & & \\
+		uniform local (tenuki) & 61.00\% & 50.00\% & \\
+		learned & \hilight{90.00\%} & \hilight{84.00\%} & \hilight{86.00\%} \\
+		\end{tabular}
+	\label{fig:results}
+	\end{center}
+\end{figure}
 \begin{itemize}
-	\item Strong and robust
+	\item All-play-all tournament of the 4 policy variants.
+	\item Each element is the percent win-rate of the row variant versus the column variant.
+	\item 200 games per pair.
 \end{itemize}
@@
 
-\begin{frame}{Future Work}
+\begin{frame}{Generalization to Different Board Sizes}
+\centerline{
+\begin{tabular}{c | cc cc}
+ & \multispan{2}{\hfil default \hfil} & \multispan{2}{\hfil uniform local\hfil} \\
+ & 11x11 & 13x13 & 11x11 & 13x13 \\ \hline
+ learned & 92.5 \% & 94 \%  & 88.5\% & 85\%
+ \end{tabular}
+}
+\begin{figure}
+\caption{Trained on 7x7 board}
+	\begin{center}
+		\begin{tabular}{c | c c}
+		& default & uniform local \\
+		\hline
+		learned & 90.00\% & 84.00\% \\
+		\end{tabular}
+	\label{fig:results}
+	\end{center}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Results versus MoHex}
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c | c}
+		& MoHex on 7x7 \\
+		\hline
+		default & 11.75\% \\
+		uniform local & 26\% \\
+		learned & 42\% \\
+		\end{tabular}
+	\end{center}
+\end{figure}
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{c  | c c c}
+		& MoHex on 11x11 & Relative CPU \\
+		\hline
+		default & 0.5\% & 100\% \\
+		uniform local & 0.0\% & 107\% \\
+		learned & 11\% & 122\% \\
+		\end{tabular}
+	\end{center}
+\end{figure}
 \begin{itemize}
-	\item Different encodings
-	\item General game play
-	\item Encode expert knowledge in weights
+	\item MoHex, the current world-champion, uses lots of expert and domain-specific knowledge
+	\item Good results on small board
+	\item Dramatic improvement for modest overhead
 \end{itemize}
@@
 
+\begin{frame}{Conclusion}
+\begin{itemize}
+	\item Computer Hex/Go present a huge problem space
+	\item MCTS is an effective technique to navigate that space
+	\item Shown we can automatically learn to guide MCTS to play computer Hex better
+	\item Future work - apply results to different games
+	\item Source code available at \url{https://github.com/etherealmachine/hivemind}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Conclusion}
+\centerline{Thank you for the time}
+\centerline{Questions?}
+\end{frame}
+
 \end{document}

### 2012-05-31 presentation/presentation.tex
Fixed type on tree search slide relating to Six and MoHex.
Fixed some typos

@@
 	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
-		  \HGame{d4,d3,e3,f1,c2,d6}
-		\end{HexBoard}
-		}
-	\end{figure}
-	\begin{itemize}
-		\item Connect opposing sides
-		\item Cut opponents connections
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{The Game of Hex - Example Game}
-	\begin{figure}[tb]
-	\resizebox{3.3in}{!}{ \begin{HexBoard}[board size=7]
 		  \HGame{d4,d3,e3,f1,c2,d6,b5,e2,c4,c3,a4,b6,c6,c5,d5,b7,c7,b4,a5,b3,a3,b1,b2,c1,d1}
@@
 	\begin{itemize}
-		\item Previous champion until 2006, (\emph{Six}, used circuit simulation
+		\item \emph{Six}, previous champion until 2006, used circuit simulation
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Current champion, \emph{MoHex}, does lots of this
+		\item \emph{MoHex}, current champion, does lots of this
 	\end{itemize}
@@
 		\item Uniform Local - play in local neighborhood first
-		\item Uniform Local with Tekuki (Play-Away) - possibly play in local neighborhood first
+		\item Uniform Local with Tenuki (Play-Away) - possibly play in local neighborhood first
 		\item Local Pattern Weighted - stochastic, weighted local neighborhood
@@
 
-\begin{frame}{Results of learning after training on 7x7)}
+\begin{frame}{Results of learning after training on 7x7}
 \begin{figure}
